### 2022-02-21 paper 2.Rmd
Push for team members

---
title: "Millenial vs gen z births and change in attitudes of young mothers"
output: html_document
---
heck if the data includes the borth years of the mothers, otherwise we cant know
liberalis, causing lower borth?
adoption of more freedom for women?
abortion?
women in workforce already still high
miority women have high birth or just white women?
Analysis by state
Comapre borths from eaely 2000s to now
Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new
Measuring error: the generations are highly arbitrary 
what type of sampling was used? 10 samples from each state. what happened to non response?

subset on age 15-19 and 20-24, as well as from 1980s onward.
caption using all 51 states in america from year to year, put source for inner captions
```{r}
install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
```

```{r}
birth = read_csv("fig_1.csv")
numbirth = read_csv("numbirths_2001_2019.csv")
student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
rent = read_dta("longdiff-rentcost-0419.dta")
religion = read_dta("long-diff-religion.dta")
childcare = read_dta("child_care_costs_under12_0408_1519.dta")
```
STUFF FOR REPRODUCTION

```{r}
birth %>%
  ggplot(data=., aes(x=year, y=brate_all))+
  geom_line()+
  theme_classic()+
  ylim(50,80)

```

```{r}
race = read_csv("figs_2a_2b.csv")
```

```{r}
race %>%
  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024", "brate_2529", "brate_3034", "brate_3539", "brate_4044"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24","Age 25-29","Age 30-34","Age 35-39", "Age 40-44")) +
  theme(legend.position = "bottom")
```

```{r}
race %>%
  filter(year >= 1990)%>%
  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
               names_to = "racegroup",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
  ylim(0,140)+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
  theme(legend.position = "bottom")
```

STUFF FOR ACTUAL PROJECT

does birth control reduce usage for young mothers?
we dont expect income ratio to affect as these are young moms anyway.
religious change, specially in bible belt?

```{r}
birth %>%
  filter(year >= 1999) %>%
  ggplot(data=., aes(x=year, y=brate_all))+
  geom_line()+
  theme_classic()+
  ylim(50,80)
```
```{r}
longdiff = read_dta("longdiff-RHS-1519.dta")
```

```{r}
race %>%
  select(year,brate_1519,brate_2024) %>%
  filter(year >= 2000) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
  theme(legend.position = "bottom")
```
```{r}
race %>%
  filter(year >= 2000) %>%
  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024","brate_2529", "brate_3034", "brate_3539", "brate_4044"),
               names_to = "group",
               values_to = "value"
               ) %>%
  mutate(group = 
           case_when(
             group == "brate_1519" ~ "Ages 15 to 19",
             group == "brate_2024" ~ "Ages 20 to 24",
             TRUE ~ "Other Age groups"
             ) # group scores into separate categories
  ) %>%
  ggplot(data=.)+
  theme_minimal()+
  geom_bar(mapping = aes(x = year, fill = group),position="fill")+ # coloring the different categories
  scale_y_continuous(labels = scales::percent)+ # getting the percentage of scores within each ward
  guides(fill=guide_legend(title = "Score Group"))+
  labs(x = "Ward",
       y = " Percentage of Apartments in ward",
       color = "Evaluation Score") 
```
out of pop 1519 and pop 2024, how many are giving birth?

OR declining population caused by declining birth rates?

This is declining of race as a possible explanation to declining birth rates
```{r}
birth3 = birth2 %>%
  select(year,pop,pop1519,pop2024,popwhitenh1519,popwhitenh2024,popblacknh1519,popblacknh2024,pophispanic1519,pophispanic2024) %>%
  filter(year >= 2000) %>%
  group_by(year)%>%
  summarize(sumpop = sum(pop), sum15 = sum(pop1519), sum20 = sum(pop2024),sumwhite15 = sum(popwhitenh1519), sumwhite20 = sum(popwhitenh2024), sumblack15 = sum(popblacknh1519), sumblack20 = sum(popblacknh2024),sumhsp15 = sum(pophispanic1519), sumhsp20 = sum(pophispanic2024))

```

```{r}
birth2 = birth2 %>%
  filter(year >=2000)

ggplot(birth2, aes(x=pop1519)) +
  geom_histogram(color = "black", fill="white")+ #to remove the bins=30 error
  theme_minimal()+
  xlab("Evaluation Score") +
  ylab("Number of Apartments")
```
```{r}
ggplot(birth2, aes(x=pop2024)) +
  geom_histogram(color = "black", fill="white")+ #to remove the bins=30 error
  theme_minimal()+
  xlab("Evaluation Score") +
  ylab("Number of Apartments")
```
Change in birth rates map for 1519 + 2024 for 2000 to 2019
```{r}
birth1 = read_dta("nchs_births_pop_1990_2019.dta")
map = birth1 %>%
  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
  filter(!stname %in% c("AB","BC","MB","NB","NS","ON","QC","SK","XX")) %>%
  group_by(stname, year) %>%
  summarize(sum = sum(numbirth2034), .groups = "drop")
```

```{r}
birth2 = read_dta("age_race_comp_seer.dta")
map2 = birth2 %>%
  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
  group_by(stname, year) %>%
  summarize(sumpop = sum(pop2034),.groups = "drop")
```

```{r}
map01 = map %>%
  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
map02 = map2 %>%
  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))

```

```{r}
mapjoin = cbind(map01,map02)

mapjoin = mapjoin %>%
  select(-c(4:5))%>%
  mutate(birthrate = (sum/sumpop)*1000)

```

```{r}
mapjoin2019 = mapjoin %>%
  group_by(stname) %>%
  slice_max(order_by = year) %>%
  group_by(stname) %>%
  summarize(average = mean(birthrate))
```

```{r}
mapjoin2000 = mapjoin %>%
  group_by(stname) %>%
  slice_min(order_by = year) %>%
  group_by(stname) %>%
  summarize(average = mean(birthrate))
```

```{r}
mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
  mutate(decline = average.x - average.y) %>%
  mutate(fips = fips(stname)) 
```

```{r, fig.cap = "Fall in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
plot_usmap(data = mapdiff, values = "decline", regions = "states") + 
  scale_fill_continuous(low = "red", high = "green", name = "Change in birth rates (%)", label = scales::comma) +
  theme(panel.background=element_blank()) +
  theme(legend.position = "right")
```

# Table 1

```{r}
table = numbirth %>%
  group_by(year) %>%
  summarize(sumbirth = sum(numbirth1544))

birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

totalpop = birth1 %>%
  group_by(year) %>%
  summarize(sumpop = sum(pop1544), sumbirthwhi15 = sum(numbirthwhite15),sumbirthbla15=sum(numbirthblack15),sumbirthhsp15 = sum(numbirthhisp15),sumbirthwhi20 = sum(numbirthwhite20),sumbirthbla20=sum(numbirthblack20),sumbirthhsp20 = sum(numbirthhisp20))

ageracepop = birth2 %>%
  group_by(year) %>%
  summarize(sumw15 = sum(popwhitenh1519), sumw20 = sum(popwhitenh2024), sumblack15 = sum(popblacknh1519), sumblack20 = sum(popblacknh2024),sumhsp15 = sum(pophispanic1519), sumhsp20 = sum(pophispanic2024))

birth = birth %>%
  filter(year >= 1990 & year < 2020)

fulltable = cbind(table,totalpop,ageracepop,birth)
fulltable = fulltable %>%
  select(-c(3,11,18))
```

```{r}
fulltable %>%

```

implications for teenaged mom is a good thing that the borth rates are decreasing? deop for teenage that teen birth decrease, for 2024 college and grad school. Check how teen pregnancy  and female enollment has increased. Putting of their mom lives for careers?

```{r}
birth2 %>%
  filter(year >= 2000)%>%
  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
               names_to = "racegroup",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
  theme(legend.position = "bottom")
```

Declining hispanic birth rates, white stayed relatively constant, there was a spike

```{r}
studentdebt = student_debt %>%
  group_by(year) %>%
  summarize(meandebt = mean(stud_loan_cap))

renttable = rent %>%
  group_by(year) %>%
  summarize(meanrent = mean(avg_grossrent))

religiontable = religion %>%
  group_by(year) %>%
  mutate(important = very_important+somewhat_important/(very_important+somewhat_important + not_too_important + not_at_all_important )) %>%
  summarize(meanpercentimportant = mean(important)) %>%
  select(year, meanpercentimportant)

childcare = childcare[complete.cases(childcare), ]
childtable = childcare %>%
  group_by(year) %>%
  summarize(meanchildcare = mean(coll_c3))

explanatory_table = cbind(studentdebt,renttable,religiontable,childtable)
explanatory = explanatory_table %>%
  select(-c(1, 3,5,7))

explanatory %>%
  knitr::kable(col.names = c("Mean Student Debt", "Mean rent", "Percentage important","Mean childcosts"), caption = "Changes in possible explanatory variables") 
```
NOTE: for the childcare table, the values for some states are missing. discuss this either in the data or in weaknesses/nextsteps

Weaknesses next steps- scrape the data and form our own initial datasets?
```{r}
sum(is.na(childcare))
```
social explanations: teen birth rate?

### 2022-02-22 paper 2.Rmd
Added intro, some data, and references

@@
 ---
-title: "Millenial vs gen z births and change in attitudes of young mothers"
-output: html_document
+title: "Social and Economic Changes have decreased birth rates for millenial and Gen z Mothers (Work in progress)"
+author: "Olaedo Okpareke, Arsh"
+thanks: 'Code and data are available at: https://github.com/OlaedoOkpareke/Paper2Repository. DOI link:' 
+date: "`r format(Sys.time(), '%d %B %Y')`"
+abstract: " "
+output:
+  bookdown::pdf_document2
+toc: FALSE
+bibliography: references2.bib
+
 ---
-heck if the data includes the borth years of the mothers, otherwise we cant know
+A nation's birthrate for a particular year is the total number of live births for a particular year.
+In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS), using the Vital Statistics Final Birth Report and the 2020 Provisional Birth Report (cite both) aggregates data on birth rates for women by age, race, and educational status.
+The total population of the country, provided to  to plan public services such as health care, education, and transportation is provided by the CDC SEER and also breaks down the population by age and race.
+In this report,we are interested in the trend in birth rates for mothers who would be seen today as millenial and generation Z for two different years 20 years apart, and these breakdown in birth rates by race, as well as the relationship between changes in possible explanatory factors like economic, social and cultural aspects of the United states and cganges in these birth rates.
+
+We obtained the dataset from the American Economic Association (cite website).
+We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
+Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap] and `kableExtra` [@citekable].
+We first created graphs such as lineplots, scatterplots and maps which showed the data: the trend in birth rates over the years, both for different age groups and then different races.
+We then created tables that showed the impact of our chosen groups on total birth rate change, as well as showed the statistics for possible causes or correlations to the birth rate changes.
+
+Young women are not as eager to have children, as this report shows that birth rates have declined for both age groups across the 20 years, and birth rate break downs according to race has also decreased.
+This implies that the decline is a general national issue as opposed to a specific demogrpahic.
+The birth rates for the millenial group and Gen z group have gone through significant declines, as well as birth rates for all races experience significant declines.
+The decline of birth rates for our target groups have also contributed to a significant amount of the total decline in birth rates.
+Birth rates for our demographic in some stated also experiences large decreases, while some other states had little to no decrease.
+The variables for birth ratw decline also increased over the time period, correlating to lower birth rates.
+According to (cite), these rates of birth decline are similar to declines in other high income, developed countries.
+
+While this census is widely used to forecast population growth, biases and problems in data collection could paint an incorrect view of population and birth rates.
+this data may also systematically hide certain groups of people and their experiences from the ranks, as well as lead to the ignition of certain harmful political ideologies.
+The declining birth rates brings to light the problems of increasing cost of living, as well as the opportunity costs of working and childcare for women throughout the country, and helps arguments of economic policies that could further hurt the US populace (cite).
+However the data and statistics also shows a decline in teenage pregnancy, as well as increases in some economic aspects that benefit women.
+There are also concerns about biases in measurement, ethical collection and inclusion of all, as well as the future state of population growth in the US and its consequences.
+
+# Data
+
+## Data Source and Collection
+
+The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" (cite), a paper from the Journal of Economic Perspectives.
+The data used by the paper was compiled by various sources such as the NCHR,CDC SEER, NBER Natality Database, and the Current Population Survey
+Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all).
+The datasets were last accessed on February 16th, 2022.
+These sources collect and estimate data via national census from US citizens by county or state over the years 1969 to 2019.
+The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample in other states (cite).
+The national population data from the census counts the entire population of the United States by county (cite).
+The datasets compiled by the paper authors typically include demographic data by year and for each state.
+The data in mainly numeric, counting the number of births, total population, or women in a particular demographic, age group, or population.
+Meanwhile, categorical data is maimly in the form of state.
+
+## Data cleaning
+
+For data cleaning, our intended sample was young mothers from two generaions: millenial mothers, i.e women who would be in the age groups of 15-19 or 20-24 in the year 2000, and generation z mothers, who would be 15-19 or 20-24 in the year 2020.
+The year chosen was 2000 to 2020, to see the change in birthrates from one generation to the other.
+Millenials are considered to be from 1977 to 1995, and gen z from 1997 to 2012 (cite).
+We also considered birth rates for white, black, and hispanic women, both for all ages and our target ages in order to compare them.
+We thus filtered our datasets based on those criteria, or used the closest criteria when they were not available.
+
+We included a dataset if they were considered important for the question or revealed important information.
+We included the race dataset as it gave the main information of the paper in the form of birth rate information for different age groups and races.
+We also included the numbirth dataset as it had information on birthrates for all the age groups together.
+We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination.
+birth2 had total population data for each age group and race-age combination.
+We used rent, student_debt, religion, and childcare as they provided information on the change in rental cost, student debt, importance of religion and rise in child costs, which might have a relationship with the birth rates.
+We created several of our own datasets by modifying and joining variables.
+These datasets including mapdiff, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, fulltable, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time.
+Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
+The full explanation of all modified variables will be explained according in their relvant table/figure.
+
+Check if the data includes the borth years of the mothers, otherwise we cant know
 liberalis, causing lower borth?
@@
 women in workforce already still high
+how do you decide whos woman or not, black or not (self identififcation, problems with hiding people). other is grouped together
 miority women have high birth or just white women?
@@
 what type of sampling was used? 10 samples from each state. what happened to non response?
-
+who do we decide is gen z or millenial? 
+can't count all, so what happened to non-response?
+The problem of self identification: problems in measuring
+why is hispanic decline so large
+harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants
 subset on age 15-19 and 20-24, as well as from 1980s onward.
 caption using all 51 states in america from year to year, put source for inner captions
-```{r}
+
+```{r global_options, include=FALSE}
+
+knitr::opts_chunk$set(fig.pos = 'H')
+library("kableExtra")
 install.packages("usmap")
@@
 childcare = read_dta("child_care_costs_under12_0408_1519.dta")
+birth2 = read_dta("age_race_comp_seer.dta")
+birth1 = read_dta("nchs_births_pop_1990_2019.dta")
+race = read_csv("figs_2a_2b.csv")
 ```
@@
 
-```
-
-```{r}
-race = read_csv("figs_2a_2b.csv")
 ```
@@
 religious change, specially in bible belt?
-
-```{r}
-birth %>%
-  filter(year >= 1999) %>%
-  ggplot(data=., aes(x=year, y=brate_all))+
-  geom_line()+
-  theme_classic()+
-  ylim(50,80)
-```
-```{r}
-longdiff = read_dta("longdiff-RHS-1519.dta")
-```
 
@@
 ```{r}
-birth1 = read_dta("nchs_births_pop_1990_2019.dta")
 map = birth1 %>%
@@
   group_by(year) %>%
+  filter(year >= 2000)
   summarize(sumbirth = sum(numbirth1544))
@@
   group_by(year) %>%
+  filter(year >= 2000) %>%
   summarize(sumpop = sum(pop1544), sumbirthwhi15 = sum(numbirthwhite15),sumbirthbla15=sum(numbirthblack15),sumbirthhsp15 = sum(numbirthhisp15),sumbirthwhi20 = sum(numbirthwhite20),sumbirthbla20=sum(numbirthblack20),sumbirthhsp20 = sum(numbirthhisp20))
@@
   group_by(year) %>%
+  filter(year >= 2000) %>%
   summarize(sumw15 = sum(popwhitenh1519), sumw20 = sum(popwhitenh2024), sumblack15 = sum(popblacknh1519), sumblack20 = sum(popblacknh2024),sumhsp15 = sum(pophispanic1519), sumhsp20 = sum(pophispanic2024))
@@
 birth = birth %>%
-  filter(year >= 1990 & year < 2020)
+  filter(year >= 2000 & year < 2020)
 
@@
 ```{r}
-birth2 %>%
+race %>%
   filter(year >= 2000)%>%
@@
 Weaknesses next steps- scrape the data and form our own initial datasets?
-```{r}
-sum(is.na(childcare))
-```
+
 social explanations: teen birth rate?

### 2022-02-23 paper 2.Rmd
changes

@@
 A nation's birthrate for a particular year is the total number of live births for a particular year.
-In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS), using the Vital Statistics Final Birth Report and the 2020 Provisional Birth Report (cite both) aggregates data on birth rates for women by age, race, and educational status.
+In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, and educational status.
 The total population of the country, provided to  to plan public services such as health care, education, and transportation is provided by the CDC SEER and also breaks down the population by age and race.
@@
 this data may also systematically hide certain groups of people and their experiences from the ranks, as well as lead to the ignition of certain harmful political ideologies.
+It could also propogate preconceived biases against women living in poor areas and racial minorities as being "irresponsible" with their bodeis.
 The declining birth rates brings to light the problems of increasing cost of living, as well as the opportunity costs of working and childcare for women throughout the country, and helps arguments of economic policies that could further hurt the US populace (cite).
@@
 Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all).
-The datasets were last accessed on February 16th, 2022.
+The paper and datasets were last accessed on February 16th, 2022.
 These sources collect and estimate data via national census from US citizens by county or state over the years 1969 to 2019.
-The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample in other states (cite).
-The national population data from the census counts the entire population of the United States by county (cite).
+The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample which was randomly chosen in other states (cite).
+The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate (cite).
 The datasets compiled by the paper authors typically include demographic data by year and for each state.
@@
 We thus filtered our datasets based on those criteria, or used the closest criteria when they were not available.
+2019 is also used if 2020 is not available.
 
@@
 We also included the numbirth dataset as it had information on birthrates for all the age groups together.
+We used birthhist as it gave the breakdown on birth rates by both age group and race.
 We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination.
@@
 Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
-The full explanation of all modified variables will be explained according in their relvant table/figure.
 
-Check if the data includes the borth years of the mothers, otherwise we cant know
-liberalis, causing lower borth?
-adoption of more freedom for women?
-abortion?
-women in workforce already still high
-how do you decide whos woman or not, black or not (self identififcation, problems with hiding people). other is grouped together
-miority women have high birth or just white women?
-Analysis by state
-Comapre borths from eaely 2000s to now
-Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new
-Measuring error: the generations are highly arbitrary 
-what type of sampling was used? 10 samples from each state. what happened to non response?
-who do we decide is gen z or millenial? 
-can't count all, so what happened to non-response?
-The problem of self identification: problems in measuring
-why is hispanic decline so large
-harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants
-subset on age 15-19 and 20-24, as well as from 1980s onward.
-caption using all 51 states in america from year to year, put source for inner captions
+We modified many variables for the report.
+We first modified the `numbirth` dataset to add variables for the birth rates for each race and age group combination, as they were unavailable in the original dataset.
+We did this by adding up all educational statuses for each race and age group combination, as each particpant of borth rates was also recorded their race and age status.
+The final variables were named `numbirthraceage`, aka.
+`numbirthwhite15` for births for the white 15-19 population.
+The `birthhist` dataset was gotten by removing all other variables from the numbirth dataset that were not the newly created ones, and then compiling the values by state and year.
+
+We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`.
+We also modified the race columns of the dataset (`bratewhitenh, brateblacknh, bratehisp`) to be rows within the column of `racegroup`.
+
+For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of birth and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state.
+The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group.
+From those two new datasets, we changed the year column so that it only stated 2000 or 2019.
+We then joined those two datasets together and divided the number of births by the population to get the birth rate.
+We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and colected the mean of each value.
+To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.
 
@@
 
-does birth control reduce usage for young mothers?
-we dont expect income ratio to affect as these are young moms anyway.
-religious change, specially in bible belt?
+We are interested in the trend of birth rates for woman ages 15-19 and 20-24 over a 20 year period.
+This shows the decline of children and the increase of the average birth age and perhaps increase in abortion.
+This has possible and already happened implications for public social and economic policy.
+show a measure of income inequality in housing, and have possible implications for public policy.
+(Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of birth rates for women in the two age groups, subsectioned by race.
+
+These plot show the range of births in the united states according to year
+We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant year being between 60 to 190 thousand.
+The data seems evenly skewed and unimodal.
+We see that teenage borths tend to be less popular, as the data seems slightly right skewed.
+This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others (cite).
+We also see that black teenaged mothers aare less likely, as they have the lowest births for most years.
+This is most likely because they are a much smaller segemnt of the population (13% of the population).
+However we see that black mothers still do manage to have a moderate amount of births.
+Hispanic teenaged mothers are more middle grtound, having between 60 to 100 thousand births per year.
+White teenaged mothers are the most likely, do to them being the most in population.
+They tend to have most of theuir distribution in the high areas, but none in the low areas.
+
+For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes.
+Due to this it is difficult to determine whether the mean or the median of births will be larger.
+We see that there are far more women giving birth at this age group than in the previous age group.
+We see that young women black mothers are overwhelmingly on the lower end of births, which brings into question why teenage blak mothers had more prevalent births.
+Hispanic mothers also tend to be at the lower end of the scale, but are more spread out.
+White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.
+
+We see that brths for black and hispanic mothers at age 20-24 are only double the births of their teenaged counterparts, while white mothers tend to be as much as 5times the number of their teenaged counterparts, Therefore, teenaged motherhood in blakc and hispanic women may be more likelty.
+
+```{r}
+birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519
+
+birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519
+
+birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519
+
+birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024
+
+birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024
+
+birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024
+
+birthhist = birth1 %>%
+  filter(year >= 2000) %>%
+  group_by(year) %>% 
+  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
+  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20= sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20=sum(numbirthhisp20))
+
+```
+
+```{r}
+birthhist %>%
+  pivot_longer(cols = c("white15", "black15","hisp15"),
+               names_to = "group",
+               values_to = "value"
+               ) %>%
+  ggplot(data =., aes(x=value, color = group, fill=group)) +
+  geom_histogram()+ 
+  theme_minimal()+
+  xlab("Numbe of births") +
+  ylab("Number of Years")
+
+```
+```{r}
+birthhist %>%
+  pivot_longer(cols = c("white20", "black20","hisp20"),
+               names_to = "group",
+               values_to = "value"
+               ) %>%
+  ggplot(data =., aes(x=value, color = group, fill=group)) +
+  geom_histogram()+ 
+  theme_minimal()+
+  xlab("Number of Births") +
+  ylab("Number of Years")
+
+```
+(Figure \@ref(fig:figthree)) shows the trend of births for the two age groups in a 20 year period.
+Millenials would be between ages 15-24 in 2000, and gen Z between ages 15-24 in 2020.
+We see that there has been a significant decrease in the birth rates, especialy in the 15-19 (teen pregnancy), which has dropped to half its 2000 levels.
+The 20-24 age grouped by 20 %.
+This shows the decrease in popualrity for birth at these ages.
+There was a baby boom in 2006 (cite), which is the reason for the slight increase in birth rates around that time.
+However overall birth rates for young women has dropped.
+According to () there is little reason to beleive these birth rates will increase any time soon.
+There are some possilbe explanations, including better sex ed for teenagers which led to a decrease in teen pregnancy, mroe women are being focused on their careers and are postponing births, or the increased acceptance of abortions if they can;t afford a child right now.Thesse declining birth rates could also be the cause of declining population (cite), (talk about immigrants due to delcining in discussion)
 
@@
   geom_line() +
-  geom_point() +
-  theme_classic() +
+  theme_minimal() +
   scale_x_continuous(breaks = seq(1980,2020,by = 5))+
@@
 ```
+(Figure \@ref(fig:figfour)) shows the trend of births for all age groups.
+We see that the birthrate for hispanic groups has dropped the hardest, going from 100 births per 1000 population to just 70 births for 1000 population.
+This can either be because the population has greatly increased, or that the number of births has decreased.
+As the black population has stayed relatively constant for decades, we can deduce that its decrease in population is only due to a decrease in birth rates.
+The white population has seen a decrease in recent years, but the borthrate has only decreased a little, indicating that the birth rates fro whites is either stangant or increasing.
+We can see that while for the millenial population there was a wide spread in the birth rates for different races, for gen z they have become more equal.
+This could be due to economic or social factors.
+As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children.
+
 ```{r}
 race %>%
-  filter(year >= 2000) %>%
-  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
-  pivot_longer(cols = c("brate_1519", "brate_2024","brate_2529", "brate_3034", "brate_3539", "brate_4044"),
-               names_to = "group",
+  filter(year >= 2000)%>%
+  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
+  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
+               names_to = "racegroup",
                values_to = "value"
                ) %>%
-  mutate(group = 
-           case_when(
-             group == "brate_1519" ~ "Ages 15 to 19",
-             group == "brate_2024" ~ "Ages 20 to 24",
-             TRUE ~ "Other Age groups"
-             ) # group scores into separate categories
-  ) %>%
-  ggplot(data=.)+
-  theme_minimal()+
-  geom_bar(mapping = aes(x = year, fill = group),position="fill")+ # coloring the different categories
-  scale_y_continuous(labels = scales::percent)+ # getting the percentage of scores within each ward
-  guides(fill=guide_legend(title = "Score Group"))+
-  labs(x = "Ward",
-       y = " Percentage of Apartments in ward",
-       color = "Evaluation Score") 
-```
-out of pop 1519 and pop 2024, how many are giving birth?
-
-OR declining population caused by declining birth rates?
-
-This is declining of race as a possible explanation to declining birth rates
-```{r}
-birth3 = birth2 %>%
-  select(year,pop,pop1519,pop2024,popwhitenh1519,popwhitenh2024,popblacknh1519,popblacknh2024,pophispanic1519,pophispanic2024) %>%
-  filter(year >= 2000) %>%
-  group_by(year)%>%
-  summarize(sumpop = sum(pop), sum15 = sum(pop1519), sum20 = sum(pop2024),sumwhite15 = sum(popwhitenh1519), sumwhite20 = sum(popwhitenh2024), sumblack15 = sum(popblacknh1519), sumblack20 = sum(popblacknh2024),sumhsp15 = sum(pophispanic1519), sumhsp20 = sum(pophispanic2024))
-
+  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
+  geom_line() +
+  theme_minimal() +
+  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
+  labs(x = " ",
+       y = " ",
+       color = " ") +
+  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
+  theme(legend.position = "bottom")
 ```
@@
 ```{r}
-birth2 = birth2 %>%
-  filter(year >=2000)
+birthhist %>%
+  mutate(whiteall = white15 + white20, blackall= black15 + black20, hispall = hisp15+hisp20) %>%
+  select(year, whiteall, blackall, hispall) %>%
+  pivot_longer(cols = c("whiteall", "blackall", "hispall"),
+               names_to = "racegroup",
+               values_to = "value"
+               ) %>%
+  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
+  geom_line() +
+  theme_minimal() +
+  scale_x_continuous(breaks = seq(1990,2020, by=2))+
+  labs(x = " ",
+       y = " ",
+       color = " ") 
+# add the white only 15 black only 20 populations and divide to get each inidividual birthrate  
+```
 
-ggplot(birth2, aes(x=pop1519)) +
-  geom_histogram(color = "black", fill="white")+ #to remove the bins=30 error
-  theme_minimal()+
-  xlab("Evaluation Score") +
-  ylab("Number of Apartments")
-```
-```{r}
-ggplot(birth2, aes(x=pop2024)) +
-  geom_histogram(color = "black", fill="white")+ #to remove the bins=30 error
-  theme_minimal()+
-  xlab("Evaluation Score") +
-  ylab("Number of Apartments")
-```
 Change in birth rates map for 1519 + 2024 for 2000 to 2019
@@
   group_by(year) %>%
-  filter(year >= 2000)
+  filter(year >= 2000) %>%
   summarize(sumbirth = sum(numbirth1544))
-
-birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519
-
-birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519
-
-birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519
-
-birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024
-
-birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024
-
-birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024
-
-totalpop = birth1 %>%
-  group_by(year) %>%
-  filter(year >= 2000) %>%
-  summarize(sumpop = sum(pop1544), sumbirthwhi15 = sum(numbirthwhite15),sumbirthbla15=sum(numbirthblack15),sumbirthhsp15 = sum(numbirthhisp15),sumbirthwhi20 = sum(numbirthwhite20),sumbirthbla20=sum(numbirthblack20),sumbirthhsp20 = sum(numbirthhisp20))
 
@@
 
-fulltable = cbind(table,totalpop,ageracepop,birth)
+fulltable = cbind(table,birthhist,ageracepop,birth)
 fulltable = fulltable %>%
@@
 ```{r}
-fulltable %>%
+share_of_pop = fulltable %>%
+  filter(year == 2000 | year==2019) %>%
+  mutate(sumw15/sumpop,sumw20/sumpop,sumblack15/sumpop,sumblack20/sumpop,sumhsp15/sumpop,sumhsp20/sumpop) %>%
+  pivot_longer(cols = c("sumw15/sumpop","sumw20/sumpop","sumblack15/sumpop","sumblack20/sumpop","sumhsp15/sumpop","sumhsp20/sumpop"),
+               names_to = "racepercentpop",
+               values_to = "value"
+               ) %>%
+  select(year,racepercentpop,value)
+
+birthrates = fulltable %>%
+  filter(year == 2000 | year==2019) %>%
+  mutate(w15brate = white15/sumw15*1000, w20brate = white20/sumw20*1000, b15brate = black15/sumblack15*1000, b20brate = black20/sumblack20*1000, h15brate = hisp15/sumhsp15*1000, h20brate = hisp20/sumhsp20*1000) %>%
+  pivot_longer(cols = c("w15brate", "w20brate", "b15brate", "b20brate", "h15brate", "h20brate"),
+               names_to = "racepercentbirth",
+               values_to = "birthratevalue"
+               ) %>%
+  select(year,racepercentbirth,birthratevalue)
 
@@
 
-implications for teenaged mom is a good thing that the borth rates are decreasing?
-deop for teenage that teen birth decrease, for 2024 college and grad school.
-Check how teen pregnancy  and female enollment has increased.
-Putting of their mom lives for careers?
-
 ```{r}
-race %>%
-  filter(year >= 2000)%>%
-  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
-  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
-               names_to = "racegroup",
-               values_to = "value"
-               ) %>%
-  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
-  geom_line() +
-  theme_classic() +
-  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
-  labs(x = " ",
-       y = " ",
-       color = " ") +
-  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
-  theme(legend.position = "bottom")
+65.9-58.3
 ```
-
-Declining hispanic birth rates, white stayed relatively constant, there was a spike
 
@@
 social explanations: teen birth rate?
+```{r}
+sample |>
+  group_by(group, favorite_color) |>
+  count() |>
+  ungroup() |>
+  group_by(group) |>
+  mutate(prop = n / sum(n)) |>
+  knitr::kable(
+    caption = "Proportion of the treatment and control group that prefer blue or white",
+    col.names = c("Group", "Preferred color", "Number", "Proportion"),
+    digits = 2,
+    booktabs = TRUE,
+    linesep = ""
+  )
+```

### 2022-02-24 paper 2.Rmd
changes

@@
 We also included the numbirth dataset as it had information on birthrates for all the age groups together.
-We used birthhist as it gave the breakdown on birth rates by both age group and race.
+We used the `birthhist` dataset as it gave the breakdown on birth rates by both age group and race.
 We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination.
@@
 We created several of our own datasets by modifying and joining variables.
-These datasets including mapdiff, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, fulltable, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time.
+These datasets including `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `fulltable`, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time.
 Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
@@
 We modified many variables for the report.
-We first modified the `numbirth` dataset to add variables for the birth rates for each race and age group combination, as they were unavailable in the original dataset.
+We first modified the `numbirth` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset.
 We did this by adding up all educational statuses for each race and age group combination, as each particpant of borth rates was also recorded their race and age status.
@@
 We also modified the race columns of the dataset (`bratewhitenh, brateblacknh, bratehisp`) to be rows within the column of `racegroup`.
+These variables were used to get the trend in birthrates for our age demographocs, and trend in birthrates in race for all demographics.
 
@@
 To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.
+
+For the first table, we got the `table` dataset by grouping `numbirth` dataset by year of more than 2000 and then summing up the entire female population by state, which was stored as `sumbirth`.
+We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations.
+We named these variables for each combination `sumraceage`, e.g `sumblack20`.
+We filtered the `birth` dataset between 2000 and 2020.
+We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`.
+From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable.
+From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by dividing the number of birhs for eact race-age combination by the total number of births.
+These variables were named `raceagebrate`, e.g `b20brate`.
+We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` for 2000.
+
+We got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states.
+We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively.
+We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values and `not_too_important` and `not_at_all_important`, and then getting the mean of all the states.
+Note that for these variables we needed to use 2004 as 2000 was not avaiable.
 
@@
 ```
+
+## Results
+
 (Figure \@ref(fig:figthree)) shows the trend of births for the two age groups in a 20 year period.
@@
   filter(year == 2000 | year==2019) %>%
-  mutate(w15brate = white15/sumw15*1000, w20brate = white20/sumw20*1000, b15brate = black15/sumblack15*1000, b20brate = black20/sumblack20*1000, h15brate = hisp15/sumhsp15*1000, h20brate = hisp20/sumhsp20*1000) %>%
+  mutate(w15brate = white15/sumbirth*1000, w20brate = white20/sumbirth*1000, b15brate = black15/sumbirth*1000, b20brate = black20/sumbirth*1000, h15brate = hisp15/sumbirth*1000, h20brate = hisp20/sumbirth*1000) %>%
   pivot_longer(cols = c("w15brate", "w20brate", "b15brate", "b20brate", "h15brate", "h20brate"),
@@
 ```{r}
-65.9-58.3
+contribution = birthrates %>%
+  pivot_wider(names_from = year, values_from = birthratevalue)
+
+contribution$contribution = ((contribution$"2000" - contribution$"2019")/65.9)*20
+
 ```
@@
 social explanations: teen birth rate?
-```{r}
-sample |>
-  group_by(group, favorite_color) |>
-  count() |>
-  ungroup() |>
-  group_by(group) |>
-  mutate(prop = n / sum(n)) |>
-  knitr::kable(
-    caption = "Proportion of the treatment and control group that prefer blue or white",
-    col.names = c("Group", "Preferred color", "Number", "Proportion"),
-    digits = 2,
-    booktabs = TRUE,
-    linesep = ""
-  )
-```

### 2022-02-25 Output/paper 2.Rmd
changed code, text, added folders

---
title: "Social and Economic Changes have decreased birth rates for millenial and Gen z Mothers (Work in progress)"
author: "Olaedo Okpareke, Arsh"
thanks: 'Code and data are available at: https://github.com/OlaedoOkpareke/Paper2Repository. DOI link:' 
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "PUT SOMETHING HERE"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references2.bib

---
A nation's birthrate for a particular year is the total number of live births for a particular year. In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, and educational status. The total population of the country, provided to  to plan public services such as health care, education, and transportation is provided by the CDC SEER and also breaks down the population by age and race. In this report,we are interested in the trend in birth rates for mothers who would be seen today as millenial and generation Z for two different years 20 years apart, and these breakdown in birth rates by race, as well as the relationship between changes in possible explanatory factors like economic, social and cultural aspects of the United states and cganges in these birth rates.

We obtained the dataset from the American Economic Association (cite website). We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap] and `kableExtra` [@citekable]. We first created graphs such as lineplots, scatterplots and maps which showed the data: the trend in birth rates over the years, both for different age groups and then different races. We then created tables that showed the impact of our chosen groups on total birth rate change, as well as showed the statistics for possible causes or correlations to the birth rate changes.

Young women are not as eager to have children, as this report shows that birth rates have declined for both age groups across the 20 years, and birth rate break downs according to race has also decreased. This implies that the decline is a general national issue as opposed to a specific demogrpahic. The birth rates for the millenial group and Gen z group have gone through significant declines, as well as birth rates for all races experience significant declines. The decline of birth rates for our target groups have also contributed to a significant amount of the total decline in birth rates. Birth rates for our demographic in some stated also experiences large decreases, while some other states had little to no decrease. The variables for birth ratw decline also increased over the time period, correlating to lower birth rates. According to (cite), these rates of birth decline are similar to declines in other high income, developed countries.

While this census is widely used to forecast population growth, biases and problems in data collection could paint an incorrect view of population and birth rates. this data may also systematically hide certain groups of people and their experiences from the ranks, as well as lead to the ignition of certain harmful political ideologies. It could also propogate preconceived biases against women living in poor areas and racial minorities as being "irresponsible" with their bodeis. The declining birth rates brings to light the problems of increasing cost of living, as well as the opportunity costs of working and childcare for women throughout the country, and helps arguments of economic policies that could further hurt the US populace (cite). However the data and statistics also shows a decline in teenage pregnancy, as well as increases in some economic aspects that benefit women. There are also concerns about biases in measurement, ethical collection and inclusion of all, as well as the future state of population growth in the US and its consequences.

# Data

## Data Source and Collection

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" (cite), a paper from the Journal of Economic Perspectives. The data used by the paper was compiled by various sources such as the NCHR,CDC SEER, NBER Natality Database, and the Current Population Survey
Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all). The paper and datasets were last accessed on February 16th, 2022. These sources collect and estimate data via national census from US citizens by county or state over the years 1969 to 2019. The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample which was randomly chosen in other states (cite). The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate (cite). The datasets compiled by the paper authors typically include demographic data by year and for each state. The data in mainly numeric, counting the number of births, total population, or women in a particular demographic, age group, or population. Meanwhile, categorical data is maimly in the form of state.

## Data cleaning

For data cleaning, our intended sample was young mothers from two generaions: millenial mothers, i.e women who would be in the age groups of 15-19 or 20-24 in the year 2000, and generation z mothers, who would be 15-19 or 20-24 in the year 2020. The year chosen was 2000 to 2020, to see the change in birthrates from one generation to the other. Millenials are considered to be from 1977 to 1995, and gen z from 1997 to 2012 (cite). We also considered birth rates for white, black, and hispanic women, both for all ages and our target ages in order to compare them. We thus filtered our datasets based on those criteria, or used the closest criteria when they were not available. 2019 is also used if 2020 is not available.

We included a dataset if they were considered important for the question or revealed important information. We included the race dataset as it gave the main information of the paper in the form of birth rate information for different age groups and races. We also included the numbirth dataset as it had information on birthrates for all the age groups together. We used the `birthhist` dataset as it gave the breakdown on birth rates by both age group and race. We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination. birth2 had total population data for each age group and race-age combination. We used rent, student_debt, religion, and childcare as they provided information on the change in rental cost, student debt, importance of religion and rise in child costs, which might have a relationship with the birth rates. 
We created several of our own datasets by modifying and joining variables.  These datasets including `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `fulltable`, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time. Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We first modified the `numbirth` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset. We did this by adding up all educational statuses for each race and age group combination, as each particpant of borth rates was also recorded their race and age status. The final variables were named `numbirthraceage`, aka. `numbirthwhite15` for births for the white 15-19 population. The `birthhist` dataset was gotten by removing all other variables from the numbirth dataset that were not the newly created ones, and then compiling the values by state and year.

We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`. We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates. These variables were used to get the trend in birthrates for our age demographocs, and trend in birthrates in race for our age demographics.

For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of birth and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state. The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group. From those two new datasets, we changed the year column so that it only stated 2000 or 2019. We then joined those two datasets together and divided the number of births by the population to get the birth rate. We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and colected the mean of each value. To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000. In this dataset we usd to 20-24 age group as the target age groups specifically were not available.

For the first table, we got the `table` dataset by grouping `numbirth` dataset by year of more than 2000 and then summing up the entire female population by state, which was stored as `sumbirth`. We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations. We named these variables for each combination `sumraceage`, e.g `sumblack20`. We filtered the `birth` dataset between 2000 and 2020. We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`. From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable. From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by multiplying the number of births for each race-age combination by 1000 and them dividing them by the total female population. These variables were named `raceagebrate`, e.g `b20brate`. We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` for 2000 and multipliying by 200. We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.

For the second table, we got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states. We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively. For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source. We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values and `not_too_important` and `not_at_all_important`, and then getting the mean of all the states. Note that for these variables we needed to use 2004 as 2000 was not available.

```{r global_options, include=FALSE}

knitr::opts_chunk$set(fig.pos = 'H')
library("kableExtra")
install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
library("gridExtra")
```

```{r}
birth = read_csv("fig_1.csv")
numbirth = read_csv("numbirths_2001_2019.csv")
student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
rent = read_dta("longdiff-rentcost-0419.dta")
religion = read_dta("long-diff-religion.dta")
childcare = read_dta("child_care_costs_under12_0408_1519.dta")
birth2 = read_dta("age_race_comp_seer.dta")
birth1 = read_dta("nchs_births_pop_1990_2019.dta")
race = read_csv("figs_2a_2b.csv")
policy = read_dta("policyvars01_19.dta")
```

STUFF FOR REPRODUCTION

```{r}
birth %>%
  ggplot(data=., aes(x=year, y=brate_all))+
  geom_line(color="blue")+
  theme_classic()+
  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) + 
  ylim(50,80) +
  ylab("Births per 1,000 women age 15–44")+
  xlab("")

```

```{r}
race %>%
  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024", "brate_2529", "brate_3034", "brate_3539", "brate_4044"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = " ",
       y = " ",
       color = " ") +
  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24","Age 25-29","Age 30-34","Age 35-39", "Age 40-44")) +
  theme(legend.position = "bottom")
```

```{r}
race %>%
  filter(year >= 1990)%>%
  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
               names_to = "racegroup",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
  ylim(0,140)+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
  theme(legend.position = "bottom")
```

STUFF FOR ACTUAL PROJECT

We are interested in the trend of birth rates for woman ages 15-19 and 20-24 over a 20 year period. This shows the decline of children and the increase of the average birth age and perhaps increase in abortion. This has possible and already happened implications for public social and economic policy. show a measure of income inequality in housing, and have possible implications for public policy. (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of birth rates for women in the two age groups, subsectioned by race.

These plot show the range of births in the united states according to year
We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant year being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage borths tend to be less popular, as the data seems slightly right skewed. This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others (cite). We also see that black teenaged mothers aare less likely, as they have the lowest births for most years. This is most likely because they are a much smaller segemnt of the population (13% of the population). However we see that black mothers still do manage to have a moderate amount of births. Hispanic teenaged mothers are more middle grtound, having between 60 to 100 thousand births per year. White teenaged mothers are the most likely, do to them being the most in population. They tend to have most of theuir distribution in the high areas, but none in the low areas.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that young women black mothers are overwhelmingly on the lower end of births, which brings into question why teenage blak mothers had more prevalent births. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.

We see that brths for black and hispanic mothers at age 20-24 are only double the births of their teenaged counterparts, while white mothers tend to be as much as 5times the number of their teenaged counterparts, Therefore, teenaged motherhood in blakc and hispanic women may be more likelty.

In contrast, the distribution for the total population of women for the past 20 years seems uniformal, with alsmost all years having their own value. This makes sense as the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are two years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration, but is most likely due to the increased lifespan of the elderly to offset the birthrates. The two dips corespind to the previous histogram of two years with the same value.

```{r}
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519

birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519

birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024

birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024

birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024

birthhist = birth1 %>%
  filter(year >= 2000) %>%
  group_by(year) %>% 
  select(year,pop1544, numbirthwhite15,numbirthwhite20,numbirthhisp15,numbirthhisp20,numbirthblack15,numbirthblack20) %>%
  summarize(sumpop = sum(pop1544),white15 = sum(numbirthwhite15), white20 = sum(numbirthwhite20), black15 = sum(numbirthblack15), black20= sum(numbirthblack20), hisp15 = sum(numbirthhisp15), hisp20=sum(numbirthhisp20))

```

```{r}
birthhist %>%
  pivot_longer(cols = c("white15", "black15","hisp15"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = group, fill=group)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Numbe of births") +
  ylab("Number of Years")

```
```{r}
birthhist %>%
  pivot_longer(cols = c("white20", "black20","hisp20"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = group, fill=group)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of Births") +
  ylab("Number of Years")

```

```{r}
par(mfrow=c(2,2))
pop_plot = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(data =., aes(x=population)) +
  geom_histogram(color="black", fill="white",bins=20)+ 
  theme_minimal()+
  xlab("Total Female Population") +
  ylab("Number of Years")

pop_plot1 = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(aes(x = year, y =population)) +
  geom_point()+ 
  theme_minimal()+
  xlab("Year") +
  ylab("Total Female Population")

grid.arrange(pop_plot,pop_plot1, ncol =2)
```

## Results

(Figure \@ref(fig:figthree)) shows the trend of births for the two age groups in a 20 year period. Millenials would be between ages 15-24 in 2000, and gen Z between ages 15-24 in 2020. We see that there has been a significant decrease in the birth rates, especialy in the 15-19 (teen pregnancy), which has dropped to half its 2000 levels. The 20-24 age grouped by 20 %. This shows the decrease in popualrity for birth at these ages. There was a baby boom in 2006 (cite), which is the reason for the slight increase in birth rates around that time. However overall birth rates for young women has dropped.  According to () there is little reason to beleive these birth rates will increase any time soon. There are some possilbe explanations, including better sex ed for teenagers which led to a decrease in teen pregnancy, mroe women are being focused on their careers and are postponing births, or the increased acceptance of abortions if they can;t afford a child right now.Thesse declining birth rates could also be the cause of declining population (cite), (talk about immigrants due to delcining in discussion). Polcy to increase families can be made to increase the borthrates, and fixing the economic system.

```{r}
race %>%
  select(year,brate_1519,brate_2024) %>%
  filter(year >= 2000) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
  theme(legend.position = "bottom")
```
(Figure \@ref(fig:figfour)) shows the trend of births for of different races for women of ages 15 to 19. We see that the birthrate for hispanic groups has dropped the hardest, going from about 62 births per 1000 population to just about 48 births for 1000 population. This can either be because the population has greatly increased, or that the number of births has decreased. The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000. The white birth rates however has taken the largest decrease however, going from about 125 births to about 65 births through the 20 years. We can see that while for the millenial population there was a wide spread in the birth rates for different races, for gen z they have much more close. This could be due to economic or social factors. As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children. White women's decrease could also be due to economic factors. Racism in healthcare and the decrease of 'adultification' of gen z black women could be the cause of the decline.

```{r}
birthhist %>%
  mutate(whiteall = white15*10000/sumpop + white20*10000/sumpop, blackall= black15*10000/sumpop + black20*10000/sumpop, hispall = hisp15*10000/sumpop +hisp20*10000/sumpop) %>%
  select(year, whiteall, blackall, hispall) %>%
  pivot_longer(cols = c("whiteall", "blackall", "hispall"),
               names_to = "racegroup",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1990,2020, by=5))+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", "White, non-Hispanic")) +
  theme(legend.position = "bottom")

```
(Figure \@ref(fig:figfive)) shows the average number of births for those aged 15-24 for every state from 2000 to 2019. The color of the points indicates the level of the employment rate.

We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier. However between the years of approximately 2003 and 2006, there was an increase in birth rates. According to (https://www.catholicnewsagency.com/news/11476/us-birth-rate-spikes-in-2006-creating-boomlet), in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom.  At this point, unemployment rates were relatively low as signified by the darker blue points between 2000 and 2007. However, this trend doesn’t continue past 2006 as there seems to be a large decrease in births which continues until 2010 and becomes less severe between 2010 and 2013. During these times, we observe lighter blue points signifying that unemployment was higher than previous years, approximately around the 6-8% range. The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births. Between 2013 and 2014 the births increase, however following 2014 they begin to decrease further each year with low unemployment rates. Although the trends seen between 2000-2013 make sense with the changing unemployment rates, the trends from 2013 onwards tell us that there are other variables likely affecting birth rates as well.

```{r}
pro = group_by(policy, year) %>% select(year, unemp)
tew <- summarise(pro, Unemployment = mean(unemp)) %>% select(Unemployment)

correct <- race

correct <- correct %>% filter(year >= 2001 & year < 2020)
graph <- cbind(correct, tew)

g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + labs(x = "Year", y = "Average Birth Rate") + guides(fill = guide_legend(title = "Unemployment Rate"))

g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + labs(x= "Year", y = "Average Birth Rate") + guides(fill = guide_legend(title = "Unemployment Rate"))

grid.arrange(g1519,g2024, nrow =2)
```

(Figure \@ref(fig:figsix)) shows the percentage decline in birth rates for the 20-34 (young women) age group by state from 2000 to 2019. We see that throughout the US there has been a decline in birth rates fro most states, with some states having more severe decline than others. We see that states on the west coast tend to have a larger decline than states on the east coast, with california having very high decline in birth rates, while that of North Dakota is nearly 0 change. This could be attriuted to the wwest coast being more liberal and refusing of gender roles for woman such that they dontnhave children as early or at all (cite). Compare that to east coast that tends to be more conservatove, religious or old days type, who encourage the standards of early marriage, especially in the southern and midwestern states of the US that more accept gender norms and refusal of new changes in culture (cite) . The exception seems to be Utah, which is consevrative religious due to the mormonism culture but has the largest decline in birth rates. This could be due to the downward turn of the Utahn economy (cite). however, the general trend still holds. Pressuring of women in religious areas and less sex ed as accuatomed ton those areas could also be the cause of the fewer decrease in south and midwest.

```{r}
map = birth1 %>%
  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
  filter(!stname %in% c("AB","BC","MB","NB","NS","ON","QC","SK","XX")) %>%
  group_by(stname, year) %>%
  summarize(sum = sum(numbirth2034), .groups = "drop")
```

```{r}
birth2 = read_dta("age_race_comp_seer.dta")
map2 = birth2 %>%
  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
  group_by(stname, year) %>%
  summarize(sumpop = sum(pop2034),.groups = "drop")
```

```{r}
map01 = map %>%
  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
map02 = map2 %>%
  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))

```

```{r}
mapjoin = cbind(map01,map02)

mapjoin = mapjoin %>%
  select(-c(4:5))%>%
  mutate(birthrate = (sum/sumpop)*1000)

```

```{r}
mapjoin2019 = mapjoin %>%
  group_by(stname) %>%
  slice_max(order_by = year) %>%
  group_by(stname) %>%
  summarize(average = mean(birthrate))
```

```{r}
mapjoin2000 = mapjoin %>%
  group_by(stname) %>%
  slice_min(order_by = year) %>%
  group_by(stname) %>%
  summarize(average = mean(birthrate))
```

```{r}
mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
  mutate(decline = average.x - average.y) %>%
  mutate(fips = fips(stname)) 
```

```{r, fig.cap = "Fall in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
plot_usmap(data = mapdiff, values = "decline", regions = "states") + 
  scale_fill_continuous(low = "red", high = "green", name = "Change in birth rates (%)", label = scales::comma) +
  theme(panel.background=element_blank()) +
  theme(legend.position = "right")
```

# Table 1

```{r}
table = numbirth %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(sumbirth = sum(numbirth1544))

ageracepop = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(sumw15 = sum(popwhitenh1519), sumw20 = sum(popwhitenh2024), sumblack15 = sum(popblacknh1519), sumblack20 = sum(popblacknh2024),sumhsp15 = sum(pophispanic1519), sumhsp20 = sum(pophispanic2024))

birth = birth %>%
  filter(year >= 2000 & year < 2020)

fulltable = cbind(table,birthhist,ageracepop,birth)
fulltable = fulltable %>%
  select(-c(3,11,18))
```

```{r}
share_of_pop = fulltable %>%
  filter(year == 2000 | year==2019) %>%
  mutate(sumw15/sumpop,sumw20/sumpop,sumblack15/sumpop,sumblack20/sumpop,sumhsp15/sumpop,sumhsp20/sumpop) %>%
  pivot_longer(cols = c("sumw15/sumpop","sumw20/sumpop","sumblack15/sumpop","sumblack20/sumpop","sumhsp15/sumpop","sumhsp20/sumpop"),
               names_to = "racepercentpop",
               values_to = "value"
               ) %>%
  select(year,racepercentpop,value)

birthrates = fulltable %>%
  filter(year == 2000 | year==2019) %>%
  mutate(w15brate = (white15*1000)/sumpop, w20brate = (white20*1000)/sumpop, b15brate = (black15*1000)/sumpop, b20brate = (black20*1000)/sumpop, h15brate = (hisp15*1000)/sumpop, h20brate = (hisp20*1000)/sumpop) %>%
  pivot_longer(cols = c("w15brate", "w20brate", "b15brate", "b20brate", "h15brate", "h20brate"),
               names_to = "racepercentbirth",
               values_to = "birthratevalue"
               ) %>%
  select(year,racepercentbirth,birthratevalue)

```

```{r}
contribution = birthrates %>%
  pivot_wider(names_from = year, values_from = birthratevalue)

contribution$contribution = ((contribution$"2000" - contribution$"2019")/65.9)*200

percentpop = share_of_pop %>%
  pivot_wider(names_from = year, values_from = value)

percentpop$"2000" = percentpop$"2000"*100
percentpop$"2019"= percentpop$"2019"*100

table1 = cbind(contribution,percentpop)
table1 = table1 %>%
  select(-c(2,3,5))

table1 %>%
  knitr::kable(col.names = c("Group", "Relative contribution to birth rate decline (%)", "Share of female population 2000 (%)","Share of female population 2019 (%)"), caption = "Contribution of target groups to total decline in birth rates from 2000 to 2019") 
```

(Table \@ref(fig:tabtwo)) compares average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019. The mean student debt is total debt while the mean rent is monthly rent. Due to the data provided, the data corresponding to 2004 is representative of the early 2000’s. All these variables are related tpo social and economic factoes of the united states which may explain the change in birth rates. We see that the average student debt has almost doubled in 2019, the mean monthly rent has increased signifcantly, the percentage of people who believe religion is important has decreased, and the monthly childcare costs have also significantly increased. These costs have all risen past inflation, indicating that it is now more expensive to live in the us in 2019 than it is in 2000. As gen z mothers want to be more financially comfortable before having kids, these could cause them to postpone having kids, or not having as many kids as its too expendice.

```{r}
studentdebt = student_debt %>%
  group_by(year) %>%
  summarize(meandebt = mean(stud_loan_cap)*10000)

renttable = rent %>%
  group_by(year) %>%
  summarize(meanrent = mean(avg_grossrent))

religiontable = religion %>%
  group_by(year) %>%
  mutate(important = very_important+somewhat_important/(very_important+somewhat_important + not_too_important + not_at_all_important )) %>%
  summarize(meanpercentimportant = mean(important)) %>%
  select(year, meanpercentimportant)

childcare = childcare[complete.cases(childcare), ]
childtable = childcare %>%
  group_by(year) %>%
  summarize(meanchildcare = mean(coll_c3))

explanatory_table = cbind(studentdebt,renttable,religiontable,childtable)
explanatory = explanatory_table %>%
  select(-c(1, 3,5,7))

explanatory %>%
  knitr::kable(col.names = c("Mean Student Debt", "Mean rent", "Percentage important","Mean childcosts"), caption = "Changes in possible explanatory variables") 
```
# Discussion
The notion of God saying “Be fruitful and increase in number; multiply on the Earth” has changed throughout time. Many Millennials have parents who are Baby Boomers and they have expressed to their children that it’s important to think for themselves. This results in a “do-it- yourself” attitude toward religion. Millennials due to this have more focus on themselves in terms of having a good education and better job. Due to this, women are more career oriented. Before 2000s, women didn’t have many choices regarding these situations and used to marry at an early age and have kids. They used to take care of the family and do household chores while men used to earn by working outside. Since then, there has been a major change which has seen
women making progress in all sorts of field such as sports, business or becoming leaders of their respected countries. Having a kid result in more responsibility being taken up which restricts things like working hours. Women having to juggle both work and home creates difficulties for themselves which can have major effect on their health. Maternity leave in the US on an average result in 10 weeks while the paid leave is 8 weeks. This can cause an issue in terms of finance as living expenses has increased from 2000s and especially since the recession in 2008. Two-parent households are on the decline in the United States as divorce, remarriage and cohabitation are on the rise. And families are smaller now, both due to the growth of single- parent households and the drop in fertility. Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed. The reason of decrease in births because women are becoming mothers later in life and the reasons are tied to increases in educational attainment, growing labour force participation and delays in marriage. The average age of mothers at first birth is 27, up from 23 in 2010.
https://www.pewresearch.org/fact-tank/2016/01/08/qa-why-millennials-are-less-religious-than-
older-americans/

https://www.pewresearch.org/social-trends/2011/10/12/in-a-down-economy-fewer-births/

## Politics

White birth rates have decreased significantly, and the 15-24 groups contribute quite a lot at more than 11% of declining birth rates. Decrease in white women borth rates often toutued by far right, which has used to harass people of color, into a "white genocide" stuff. This has also caused the increase of attacks on planned parenthood clinics as they are believed to have caused the decrease in these birth rates. A declining population is one of the main reasons people are against abortion(cite), and the increase and proloferation of far right ideologoies that white birth rates are decreaing leading to a les white country are fuled in part by the way that these statistcis can be showsn, and withiut the proper context to show why they are decreasing. As birth rates for poc has not decrwed as much it could also perpetuate harmful stereotpyes on poc women having too many hcildren, too young, or are having lazy kids on welfare.

This decrease in birth rates has led to the push for immigration policies, not jutst here but in other industrialized societies. This has led to political cotneoversy on having too many imigrants which takes jobs from the native populations. This can then cause demographic decline as it alreadty has in the dust belt(), which  further cotnributes to declining birth rates as things become too expensice (cite). These then promote feelings of hatreds towards immigrants,both for "Stealing" jobs, as well as usually having more children than the native populations. These decrease i nbirth rates can thus be used by both capitalists wanting to use immigration, or by right winged groups as a crisis that immigrants are coming and to have more kids.

## Unemployment
While unemployment rates were a factor which affected birth rates between 2000-2013, there are other variables at play as well.
The increase in birth rates from 2000 onwards is likely due to the economic and job growth that the United States had leading up to 2006. Between 2001 and 2007, the United States was going through an expansion period which contributed to its economic growth. This expansion was one that was weaker than the post WWII expansion and the one that had occured in 1990 with annual growth rates of 4.3% and 3.3% respectively however, there was still a 2.8% annual growth when looking at GDP. This growth is not evident when looking at birthrates for the 15-19 age group as the declining birth rate continues until 2005. In contrast, the results of this expansion period are more visible in the 20-24 age range as it seems that the birth rate from 2001 to 2005 decreased at a lower rate each consecutive year until 2005, where it began to increase. This is likely because of the fact that the 15-19 age group isn’t affected by an economic expansion as much as the 20-24 age group as at such a young age, most of them would still be in school. On the contrary, an economic expansion provides a lot more jobs and income for young adults who are done or almost done with school.  (https://www.cbpp.org/research/how-robust-was-the-2001-2007-economic-expansion)

In a time like this, families are more likely to be in a position where they can afford to have a child because interest rates and debts that they owe are lower than usual. With more people employed, there is generally more income for families to support the needs of an infant. (cite)

 However, that growth began to decline in 2007 where consumers began to pay more for debt that they had to repay. It is stated that economic growth had decreased to 2% in the third quarter compared to the 5.6% of first quarter growth the country had. This time also marked the end of the housing boom which is generally caused by economic prosperity and lower credits. Due to this, the decline of the economy also meant the end of the housing boom and also, the beginning of the declining birth rates.

((https://www.americanprogress.org/article/the-u-s-economy-in-review-2006/))

Due to a higher number of unemployed workers and the declining economy, affording a child became difficult which is why we notice the sudden decrease in average births across the United States. This steep decline of births seems to slow down in 2010 which is also the time that the great recession ended.

https://pubmed.ncbi.nlm.nih.gov/22066128/#:~:text=In%20most%20countries%2C%20the%20recession,decade%20of%20rising%20fertility%20rates.

## STudent debt, childcare costs, religion (by state?), rent

NOTE: for the childcare table, the values for some states are missing. discuss this either in the data or in weaknesses/nextsteps

what caused the great decline in hispanic borth rate decrease?

Elaborate on how decreased religion, increased childcare, mean rent, student debt

Weaknesses next steps- scrape the data and form our own initial datasets?

social explanations: teen birth rate?

Limitations: we couldnt get the exact age groups fror some variables, so we had to take the ages of all women or the n20-34 age groups to approximate, so we may not get the accuarte results as the general trend may not apply to the sepcific trend of teenaged and young mothers.

# Bias and Ethical Concerns

It is important to note that the fact that this data is based on people, there are likely to be some ethical implications that could limit the true accuracy of our data. Another consideration we should make a note of are potential biases that we can observe from the data.

A problem with these datasets and the original sources as well is that in terms of gender, it seems that they’ve only looked at births that happen through a female. The issue with this is that it does not account for the trans and non-binary population who are still able to get pregnant and give birth to children. This is another group of people that are not included in this study which hinders the accuracy of results. Non binary people are systematically removed from society in many offical processes and unofficial. For the fact that the official us census even in 2020 did not include a non binary or other gender is also bad. Even if they did include an other, that is grouping a lot of different gender identities into one huge group, that may be completely different fromeach other. This would then make analysis confusing as different gender identities may have different experiences. The datasets also also did not include transgender men, who can also get pregnant. This invalidates transgender men's existences by calling them women.

Adding to this, non hispanic black and whites got their own group but hispanics did not, they were all grouped into one. This is an error as hispanics can also be any race and althought they are within the hispanic ethnicity, the experiences and data for black hispanics will likely be a lot different than that of white hispanics due to racism within the hispanic community (cite). This problem may also be because most of society sees all hispanics as mestizo, thus erasing castizo and afro-hispanic experiences.

The sources for birth rates did not include illegal immigrants in their info on birth and population. This is because illegal mothers are much more likely to give birth at home or less formal places, due to fear of eviction from the country. Thus they often do not have birth certifictaes, or have inaccurate birth certificates. As these records are used for *put what they were used for in the introduction*, not including them perpetuates rhe discultison of these people from society and societal improvements, many of whom pay taxes, as they are also misused for labour, in prisons, and by ICE (cite). ALthough census information does include everybody irregardless of status, many illegals also fear coming otu due to fear of being extrajudicially removed, which has happened before (cite). this census data is also used for funding in particular areas and for infrastruture improvement, so these inaccurate results mean that not only are their experiences being drowned out, but areas where there are many of them are not getting the funding required.

# Data collection error
problems in measurement of what is black etc. did they decide on self identification, which could lead to wholely inacurate results, or did they decide for themselves, wbich could strip people of their gender identity?

Adding onto this, they did not record the data for some childcare costs, for the childcare dataset and so we do not have the full picture.

# race
south poorest area in the us, have higher births due to les education, as previously said. As black people are cocentrated there (cite), these lack of resources disproprtionately affect them and so they have higher birth rates so thats why they dont have as much contribution to declining birth rates. Whites who have more resources and better schools that can teach sex ed have less birth rates due to that better education. Hispanics also have less resoruces and less good schools. They also have les outside opportunties like school and career that could postpone birth at those ages. 
Black poor so cant find good doctors for natal care, which could affect birth rates.

*look at two links sent by Arsh at 7:58 pm and 8:28pm*

# State by state changes

The change in birthrates for some states were very different than others. Some states were able to bounce back from the economic depression (cite), while some states have always had cheaper costs of living. From *FIGURE MAP* These states, primarily the midwest, are still heavily affected by the ecomic recession that started due to the fall of the steel industry (cite) and outside sourcing of labour. However cost of living and childcare are also relatively cheap in these states, so it didnt affect birth rates as much (cite). The west coast was able to bounce back better than the midwest, but as the cost of lviving is still expensive, especially in washington, oregon and california, this could deter women from having children quite early until they are fiancially ready. In the east coast and the south, the cost of living is also quite high but as these states tend to be more religious and conservative and value the family more (cite) than the liberal west coast, birth rates did not decline as far as the west coast. It is also known that the east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher. 
However as gen z is now more liberal than their millenial counterparts they are less likely to have more kids nationwide.

# Teen birth rates, race and economic status

Teenaged birth rates have decreased in the past 20 years for gen z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted ven with the cultural change. This is due to lack of opportunities and that is all they know, as well as higher rates of poverty and low education. While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back. It is of note that although white teenagers are mothers at a larger number, and thus make up most of teen mothers, most of the concern is based on the poc women and these data can perpetuate these biases. More education is needed in these communities. Poor women irrepsective of race are also more likely to get pregnant earlier, this is mainly in the south as they still practice abstinence only sex ed. This report, which shows an increase in irreligiousnes could perpetuate the false notions that teaching respinsible sex leads to more recklenss, when it is the other way around (cite).  Increasing more liberal, science based forms of sex-ed, while controversial, would decrease teen birth rates. However, there is push back because... (cite). WHile still being religious, these people have high levels of teen pregnancy, so policy is needed in schools to fight back agaisnt abstinence based sex-ed. It is knowsn to also shame women for having sexual thought,s cotnributing to a misogynistic society(cite)

# Economic changes

Costs are increasing, higher student debt, and lack of safety nets due to the country's indivualistic nature means that gen z women are not as eager tp have kids as millenial women. 
Table 1 examined other economic factors that played a role in the lives of young women. We noticed student debt, rent and childcare costs increased from the early 2000s to late 2010s.  Most students between the ages of 15-24 do not have a steady source of income and due to the rising expenses, it becomes difficult for young women to properly raise a child. The cost of housing in the United States has doubled which plays a huge role in the decision to have a child. (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4685765/). The increased rent seen from Table 1 is the main reason that families have second thoughts about having a child. In some cities, rent is increasing by up to 40% and increases like that would definitely put families wanting to have a child in a less favorable position. (https://www.theguardian.com/us-news/2022/feb/16/renters-rent-increases-us-lease) In a similar way, the rise of student debt in the United States is also leaving young adults with less money to care for their families. According to studies, women in the United States hold approximately 67% of student and many of them state that their debt influences their decision to have a child(https://blog.futurefamily.com/press-release-student-dept-holds-back-american-fertility-rates/). When considering these two costs, and considering the fact that there are child care expenses to tend to as well, one can see how expensive it is becoming to have a child. According to a study from David M. Blau and Phillip K. Robbins, a dollar increase in weekly childcare costs implies a 2% decrease in birth rate. (https://www-jstor-org.myaccess.library.utoronto.ca/stable/2061526?seq=8#metadata_info_tab_contents) Along with this, a dollar increase in childcare cost also means that the rate of entering employment decreases by 3%. From this, one can see how there is not only a correlation between increasing childcare costs and unemployment but also a correlation with increasing childcare costs and a decreasing birth rate.

# Culture changes

Insudtrialized societies have led to a further breakdown of conservative beliefs and women becoming more idnependent. In these days more women have a choice than 20 years ago on if they want to have children at all. WOmen are also  choosing to enter careers and focus on child care in their later years and not their early 20s. This can be seen as women in states with more stay at home moms have higher 20-24 birth rates. Women are also not marrying just after college anymore (cite). the rise of lgbtq families has also led to more adoptions or not wanting to have chidren in general.
Women are also getting married later (cite us economic and social trends since 2000) and the gen z having sex is decreasing (cite), which also leads to further birthrate decline.

# Weakness / Next Steps

There were datasets in which we had to make generalizations about a specific population based on data of a general population. For example, in Figure *US MAP FIGURE*, the map data was composed of responses taken from women aged from 20-34. This serves as a limitation because it ignores an entire age group from the study that is being conducted since this paper is focused on births for those aged between 15-24. Due to this, conclusions that are made from this map are not as accurate as they could be because the true statistics of the 15-19 age group aren’t provided. In a similar manner, Table 1 also carries a degree of error as due to the data that was provided, we are restricted to making conclusions about ages 15-24 based on data which does not specify age.  Our research question wanted to focus on the trend of birth rates over a 20 year period, specifically from 2000 to 2020. However, for several datasets only data up to 2019 was available. Due to this, instead of being able to compare 2000 and 2020 explicitly, we had to generalize trends that we see in 2019 to be at least similar to trends that we’d see in 2020. We had the same problem with Table 2, that we had to approximate 2004 to 2000 even though situations would be different.

https://ifstudies.org/blog/higher-rent-fewer-babies-housing-costs-and-fertility-decline

As more women do not want to give birth and opt for adoption, that could lead to further decreases in birth rates in the future. We wuld like to make a graph that shows the relaitionships betwen adiptions and birth rates Adoption could be one reason for impact on the map.

Could falling testosterone rates(cite) also attribute to these? we would like to make a table that shows the change in the 20 years.

### 2022-02-26 Output/paper 2.Rmd
Created new files, edited code and writing

@@
 author: "Olaedo Okpareke, Arsh"
-thanks: 'Code and data are available at: https://github.com/OlaedoOkpareke/Paper2Repository. DOI link:' 
+thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
 date: "`r format(Sys.time(), '%d %B %Y')`"

### 2022-02-26 Main Code repository with data and paper/3 Graph Replication.R

@@ -1,0 +1,64 @@
+# Loading in datasets
+
+birth = read_csv("fig_1.csv")
+race = read_csv("figs_2a_2b.csv")
+write_csv(birth, "Replication-Birth Rates/birth.csv")
+write_csv(race, "Replication-Birth Rates/race.csv")
+
+# Replication of Figure 1
+birth %>%
+  ggplot(data=., aes(x=year, y=brate_all))+
+  geom_line(color="blue")+
+  theme_classic()+
+  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) + 
+  ylim(50,80) +
+  ylab("Births per 1,000 women age 15–44")+
+  xlab("") +
+  annotate(geom="text", x=2009, y=72, label="2007",
+           color="black")
+
+# Replication of Figure 2A
+race %>%
+  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
+  pivot_longer(cols = c("brate_1519", "brate_2024", "brate_2529", "brate_3034", "brate_3539", "brate_4044"),
+               names_to = "group",
+               values_to = "value"
+  ) %>%
+  ggplot(mapping = aes(x = year, y = value, color = group)) +
+  geom_line() +
+  theme_classic() +
+  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
+  scale_y_continuous(breaks = seq(0,140,by = 20))+
+  labs(x = " ",
+       y = "Births per 1,000 women in relevant population subgroup",
+       color = " ") +
+  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) +
+  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24","Age 25-29","Age 30-34","Age 35-39", "Age 40-44")) +
+  theme(legend.position = "bottom") +
+  annotate(geom="text", x=2009, y=120, label="2007",
+           color="black")
+
+
+# Replication of Figure 2B
+
+race %>%
+  filter(year >= 1990)%>%
+  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
+  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
+               names_to = "racegroup",
+               values_to = "value"
+  ) %>%
+  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
+  geom_line() +
+  theme_classic() +
[…]

### 2022-02-26 Scripts/Rscript.R

@@ -1,1 +1,39 @@
-Rscript
+#### Preamble ####
+# Purpose: Load the data from the data section of "The Puzzle of Falling US Birth Rates since the Great Recession", from  the American Economic Journal 
+# Author: Olaedo Okpareke, Arsh Lakhanpal,Swarnadeep Chattopadhyay
+# Data: 25 February 2022
+# Contact: olaedo.okpareke@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+# - Need to have installed the tidyverse, haven and tidyr packages. 
+
+
+
+#### Workspace setup ####
+
+library(haven)
+library(tidyverse)
+
+# Reading in the data in the raw data.
+birth = read_csv("fig_1.csv")
+numbirth = read_csv("numbirths_2001_2019.csv")
+student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
+rent = read_dta("longdiff-rentcost-0419.dta")
+religion = read_dta("long-diff-religion.dta")
+childcare = read_dta("child_care_costs_under12_0408_1519.dta")
+birth2 = read_dta("age_race_comp_seer.dta")
+birth1 = read_dta("nchs_births_pop_1990_2019.dta")
+race = read_csv("figs_2a_2b.csv")
+policy = read_dta("policyvars01_19.dta")
+
+# create datasets from dta files
+
+write_csv(numbirth, "Inputs/Data/numbirth.csv")
+write_csv(race, "Inputs/Data/race.csv")
+write_dta(policy, "Inputs/Data/policy.csv")
+write_csv(student_debt, "Inputs/Data/student_debt.csv")
+write_csv(rent, "Inputs/Data/rent.csv")
+write_csv(religion, "Inputs/Data/religion.csv")
+write_csv(childcare, "Inputs/Data/childcare.csv")
+write_csv(birth2, "Inputs/Data/birth2.csv")
+write_csv(birth1, "Inputs/Data/birth1.csv")

### 2022-02-27 Output/paper 2.Rmd
Added many changes to text and code

--- 2022-02-25 Output/paper 2.Rmd
@@
 ---
-title: "Social and Economic Changes have decreased birth rates for millenial and Gen z Mothers (Work in progress)"
-author: "Olaedo Okpareke, Arsh"
-thanks: 'Code and data are available at: https://github.com/OlaedoOkpareke/Paper2Repository. DOI link:' 
+title: "Decline in U.S Birthrates is Indicative of Socio-economic Changes"
+author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
+thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
 date: "`r format(Sys.time(), '%d %B %Y')`"
-abstract: "PUT SOMETHING HERE"
+abstract: "birth rates are merely a symptom of cultursl changes betweeen two generations and the economic situation.
+It is important to see that these are caused by everything more expensive now.
+make sure to cite R in the abstract."
 output:
@@
 ---
-A nation's birthrate for a particular year is the total number of live births for a particular year.
-In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, and educational status.
-The total population of the country, provided to  to plan public services such as health care, education, and transportation is provided by the CDC SEER and also breaks down the population by age and race.
-In this report,we are interested in the trend in birth rates for mothers who would be seen today as millenial and generation Z for two different years 20 years apart, and these breakdown in birth rates by race, as well as the relationship between changes in possible explanatory factors like economic, social and cultural aspects of the United states and cganges in these birth rates.
+better economic-industrial society to expensive for children, industrial has low birth rates
+adoption of more freedom for women?
+abortion? restricting of abortion could lead to higher birth rats, but to the detriment of choice
 
-We obtained the dataset from the American Economic Association (cite website).
+Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new
+
+why is hispanic decline so large
+harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants
+
+implications for teenaged mom is a good thing that the borth rates are decreasing?
+
+Check how teen pregnancy  and female enollment has increased.
+
+Its good black teen birth rate decreased (table 1), how can we increase this decrease?
+
+does birth control reduce usage for young mothers?
+we dont expect income ratio to affect as these are young moms anyway.
+religious change, specially in bible belt?
+
+possible bias: To what extent is random allocation ethical and fair?
+Some argue that shortages mean it is reasonable to randomly allocate, but that may depend on how linear the benefits are.
+It may also be difficult to establish definitions, and the power imbalance between those making these decisions and those being treated should be considered
+
+# Introduction
+
+A nation's birthrate for a particular year is the total number of live births per 1000 people for that year.
+In order to measure the birth rates and therefore fertility levels, as well as forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors.
+The total population of the country is the total number of citizens and other residents in the country, regardless of legal status.
+It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics.
+The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER).
+This information also breaks down the population by age, race, and socio-economic factors.
+In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographic of the mother.
+We are also interested in the relationship between  changes in economic, social and cultural aspects of the United states and the changes in these birth rates.
+
+We obtained the dataset from the American Economic Association [@citewebsite].
 We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
-Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap] and `kableExtra` [@citekable].
-We first created graphs such as lineplots, scatterplots and maps which showed the data: the trend in birth rates over the years, both for different age groups and then different races.
-We then created tables that showed the impact of our chosen groups on total birth rate change, as well as showed the statistics for possible causes or correlations to the birth rate changes.
+Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable].
+We first created histograms and scatterplots which showed the most important variables.
+We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states.
+We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.
 
-Young women are not as eager to have children, as this report shows that birth rates have declined for both age groups across the 20 years, and birth rate break downs according to race has also decreased.
-This implies that the decline is a general national issue as opposed to a specific demogrpahic.
-The birth rates for the millenial group and Gen z group have gone through significant declines, as well as birth rates for all races experience significant declines.
-The decline of birth rates for our target groups have also contributed to a significant amount of the total decline in birth rates.
-Birth rates for our demographic in some stated also experiences large decreases, while some other states had little to no decrease.
-The variables for birth ratw decline also increased over the time period, correlating to lower birth rates.
-According to (cite), these rates of birth decline are similar to declines in other high income, developed countries.
+Young women are in no rush to have children, as this report shows that birth rates have declined for the 15-24 age groups over the past 20 years.
+The report also shows that birth rates have declined for all races analyzed over the same period of time.
+This implies that this decline is a general national issue attributed to national factors, as opposed to factors that mainly affect a specific demographic.
+The birth rates for our target demographics have declined to the point that they make up a rather large contribution to the total decline in US birth rates.
+Birth rates for our age demographics have experienced large decreases in some states, while those in other states have had a relatively smaller decrease or no decrease at all.
+This may be due to the fact that some economic or social changes of the country affect some geographic areas more than others.
+These economic and social factors suspected to have an influence on birth rates also changed over the past 20 years in ways that would correlate with declining birth rates.
+According to [@citeadsera], these rates of birth decline are similar to declines in other high income, developed countries.
 
-While this census is widely used to forecast population growth, biases and problems in data collection could paint an incorrect view of population and birth rates.
-this data may also systematically hide certain groups of people and their experiences from the ranks, as well as lead to the ignition of certain harmful political ideologies.
-It could also propogate preconceived biases against women living in poor areas and racial minorities as being "irresponsible" with their bodeis.
-The declining birth rates brings to light the problems of increasing cost of living, as well as the opportunity costs of working and childcare for women throughout the country, and helps arguments of economic policies that could further hurt the US populace (cite).
-However the data and statistics also shows a decline in teenage pregnancy, as well as increases in some economic aspects that benefit women.
-There are also concerns about biases in measurement, ethical collection and inclusion of all, as well as the future state of population growth in the US and its consequences.
+While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population.
+The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry].
+This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their systematic discrimination in society.
+The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being "irresponsible" with their bodies.
+This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].
+There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.
 
@@
 
-The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" (cite), a paper from the Journal of Economic Perspectives.
-The data used by the paper was compiled by various sources such as the NCHR,CDC SEER, NBER Natality Database, and the Current Population Survey
+The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite].
+The data used by the paper was compiled by various sources such as the NCHS [@citenchs] CDC SEER [@citecdcseer], NBER Natality Database, and the Current Population Survey
 Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all).
 The paper and datasets were last accessed on February 16th, 2022.
-These sources collect and estimate data via national census from US citizens by county or state over the years 1969 to 2019.
-The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample which was randomly chosen in other states (cite).
-The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate (cite).
-The datasets compiled by the paper authors typically include demographic data by year and for each state.
-The data in mainly numeric, counting the number of births, total population, or women in a particular demographic, age group, or population.
-Meanwhile, categorical data is maimly in the form of state.
+The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019.
+The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which was randomly sampled [@citenchs].
+The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer].
+The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics.
+These datasets were gotten from surveys or website data, so non-response was often recorded as 0 or as `NA`.
+The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity.
+Meanwhile, categorical data mainly appeared in the form of state.
 
@@
 
-For data cleaning, our intended sample was young mothers from two generaions: millenial mothers, i.e women who would be in the age groups of 15-19 or 20-24 in the year 2000, and generation z mothers, who would be 15-19 or 20-24 in the year 2020.
-The year chosen was 2000 to 2020, to see the change in birthrates from one generation to the other.
-Millenials are considered to be from 1977 to 1995, and gen z from 1997 to 2012 (cite).
-We also considered birth rates for white, black, and hispanic women, both for all ages and our target ages in order to compare them.
-We thus filtered our datasets based on those criteria, or used the closest criteria when they were not available.
-2019 is also used if 2020 is not available.
+For data cleaning, our target sample and main focus was young mothers from two generations: millennial mothers, who were between 15 to 24 years in 2000, and generation Z mothers, who were between 15 to 24 years old in 2020.
+We observed birthrates from 2000 to 2020, in order to see the effect of relatively short term effects on birthrates between the generations.
+Millennial are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock].
+We also considered birth rates for white, black, and hispanic women that were in our target age ranges as we wanted to compare the birth rates between them.
+We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e.
+2019 data is used if 2020 data is not available.
+We dropped any `NA` values from the dataset if they appeared.
 
-We included a dataset if they were considered important for the question or revealed important information.
-We included the race dataset as it gave the main information of the paper in the form of birth rate information for different age groups and races.
-We also included the numbirth dataset as it had information on birthrates for all the age groups together.
-We used the `birthhist` dataset as it gave the breakdown on birth rates by both age group and race.
-We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination.
-birth2 had total population data for each age group and race-age combination.
-We used rent, student_debt, religion, and childcare as they provided information on the change in rental cost, student debt, importance of religion and rise in child costs, which might have a relationship with the birth rates.
+## Data Modification
+
+We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information.
+We included the `race` dataset as it gave the main data of analysis  in the form of birth rates for different age groups and races.
+We included the `birth` dataset as it included the total birthrate for all women.
+We also included the `numbirth` dataset as it had information on birthrates for all the age groups combined.
+We used the `birth1` dataset as well as it had data on birth rates for younger age groups, as well as total population data and demographic data for each race and age combination.
+`birth2` had total population data for each age group and race-age combination.
+We used`policy`, `rent`, `student_debt`, `religion`, and `childcare` as they provided information on the changes in unemployment, rental cost, student debt, importance of religion and rise in child costs throughout the US, all of which are expected to influence birth rates to varying degrees.
+
 We created several of our own datasets by modifying and joining variables.
-These datasets including `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `fulltable`, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time.
-Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
+These datasets include `birthhist`, which was gotten from the `birth1` dataset by adding all the educational status for each age-race combination to get the total birth race for that comnination, and then compiling the values by state and year.
+We also got the `sumpop` variable by adding the total population for each state in a year.
+We then obtained `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `table1`, which was used to show how our target demographics contributed to birth rate decline, and `explanatory`, which was used to display the changes in possible explanatory variables over a period of time.
+Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
 
 We modified many variables for the report.
-We first modified the `numbirth` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset.
-We did this by adding up all educational statuses for each race and age group combination, as each particpant of borth rates was also recorded their race and age status.
-The final variables were named `numbirthraceage`, aka.
-`numbirthwhite15` for births for the white 15-19 population.
-The `birthhist` dataset was gotten by removing all other variables from the numbirth dataset that were not the newly created ones, and then compiling the values by state and year.
-
+We first modified the `birth1` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset.
+We did this by adding up all educational statuses for each race and age group combination, as each person whose birth rates was recorded also had their educational status recorded.
+The final variables were named `numbirthraceage`, i.e `numbirthwhite15` for births for the white 15-19 population.
 We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`.
 We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates.
-These variables were used to get the trend in birthrates for our age demographocs, and trend in birthrates in race for our age demographics.
+These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
 
-For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of birth and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state.
+For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of births and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state.
 The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group.
-From those two new datasets, we changed the year column so that it only stated 2000 or 2019.
+From those two new datasets, we changed the `year` column so that it only stated 2000 or 2019.
 We then joined those two datasets together and divided the number of births by the population to get the birth rate.
-We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and colected the mean of each value.
+We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and collected the mean of each value.
 To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.
-In this dataset we usd to 20-24 age group as the target age groups specifically were not available.
 
-For the first table, we got the `table` dataset by grouping `numbirth` dataset by year of more than 2000 and then summing up the entire female population by state, which was stored as `sumbirth`.
+For the first table, we got the `table` dataset by grouping the `numbirth` dataset by year, beyond 2000, and then summing up the entire female population by state, which was stored as `sumbirth`.
 We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations.
@@
 These variables were named `raceagebrate`, e.g `b20brate`.
-We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` for 2000 and multipliying by 200.
+We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` value for 2000 and multiplying by 200.
 We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.
@@
 For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source.
-We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values and `not_too_important` and `not_at_all_important`, and then getting the mean of all the states.
+We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values plus `not_too_important` and `not_at_all_important`, and then getting the mean of all the states.
 Note that for these variables we needed to use 2004 as 2000 was not available.
@@
 library("kableExtra")
-install.packages("usmap")
+# install.packages("usmap")
 library("tidyverse")
@@
 
-```{r}
-birth = read_csv("fig_1.csv")
-numbirth = read_csv("numbirths_2001_2019.csv")
-student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
-rent = read_dta("longdiff-rentcost-0419.dta")
-religion = read_dta("long-diff-religion.dta")
-childcare = read_dta("child_care_costs_under12_0408_1519.dta")
-birth2 = read_dta("age_race_comp_seer.dta")
-birth1 = read_dta("nchs_births_pop_1990_2019.dta")
-race = read_csv("figs_2a_2b.csv")
-policy = read_dta("policyvars01_19.dta")
+```{r,include = FALSE}
+birth = read_csv(here::here("Inputs/Data/birth.csv"))
+numbirth = read_csv(here::here("Inputs/Data/numbirth.csv"))
+student_debt = read_csv(here::here("Inputs/Data/student_debt.csv"))
+rent = read_csv(here::here("Inputs/Data/rent.csv"))
+religion = read_csv(here::here("Inputs/Data/religion.csv"))
+childcare = read_csv(here::here("Inputs/Data/childcare.csv"))
+birth2 = read_csv(here::here("Inputs/Data/birth2.csv"))
+birth1 = read_csv(here::here("Inputs/Data/birth1.csv"))
+race = read_csv(here::here("Inputs/Data/race.csv"))
+policy = read_csv(here::here("Inputs/Data/policy.csv"))
 ```
+## View of the Data
 
-STUFF FOR REPRODUCTION
+We are interested in the trend of birth rates for women ages 15-19 and 20-24 over a 20 year period.
+This shows the decline in popularity of young motherhood and thus the increase of the average birth age.
+This has ongoing implications for public social and economic policy.
+(Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of number of births for women in the two age groups for 20 years, sectioned by race.
 
-```{r}
-birth %>%
-  ggplot(data=., aes(x=year, y=brate_all))+
-  geom_line(color="blue")+
-  theme_classic()+
-  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) + 
-  ylim(50,80) +
-  ylab("Births per 1,000 women age 15–44")+
-  xlab("")
-
-```
-
-```{r}
-race %>%
-  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
-  pivot_longer(cols = c("brate_1519", "brate_2024", "brate_2529", "brate_3034", "brate_3539", "brate_4044"),
-               names_to = "group",
-               values_to = "value"
-               ) %>%
-  ggplot(mapping = aes(x = year, y = value, color = group)) +
-  geom_line() +
-  theme_classic() +
-  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
-  scale_y_continuous(breaks = seq(0,140,by = 20))+
-  labs(x = " ",
-       y = " ",
-       color = " ") +
-  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) +
-  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24","Age 25-29","Age 30-34","Age 35-39", "Age 40-44")) +
-  theme(legend.position = "bottom")
-```
-
-```{r}
-race %>%
-  filter(year >= 1990)%>%
-  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
-  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
-               names_to = "racegroup",
-               values_to = "value"
-               ) %>%
-  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
-  geom_line() +
-  theme_classic() +
-  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
-  ylim(0,140)+
-  labs(x = " ",
-       y = " ",
-       color = " ") +
-  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
-  theme(legend.position = "bottom")
-```
-
-STUFF FOR ACTUAL PROJECT
-
-We are interested in the trend of birth rates for woman ages 15-19 and 20-24 over a 20 year period.
-This shows the decline of children and the increase of the average birth age and perhaps increase in abortion.
-This has possible and already happened implications for public social and economic policy.
-show a measure of income inequality in housing, and have possible implications for public policy.
-(Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of birth rates for women in the two age groups, subsectioned by race.
-
-These plot show the range of births in the united states according to year
-We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant year being between 60 to 190 thousand.
+We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand.
 The data seems evenly skewed and unimodal.
-We see that teenage borths tend to be less popular, as the data seems slightly right skewed.
-This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others (cite).
-We also see that black teenaged mothers aare less likely, as they have the lowest births for most years.
-This is most likely because they are a much smaller segemnt of the population (13% of the population).
-However we see that black mothers still do manage to have a moderate amount of births.
-Hispanic teenaged mothers are more middle grtound, having between 60 to 100 thousand births per year.
-White teenaged mothers are the most likely, do to them being the most in population.
-They tend to have most of theuir distribution in the high areas, but none in the low areas.
+We see that teenage births tend to be less popular, as the data seems slightly right skewed.
+This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012].
+We also see that black teenage mothers are less common, as they have the lowest births for most years.
+This is most likely because they are a much smaller segment of the population than the others.
+However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportiantely appearing as teenage mothers.
+Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year.
+White teenage mothers are the most common, due to them being the largest in population.
+They tend to have most of their distribution in the high areas towards the higher numbers.
 
@@
 We see that there are far more women giving birth at this age group than in the previous age group.
-We see that young women black mothers are overwhelmingly on the lower end of births, which brings into question why teenage blak mothers had more prevalent births.
+We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts.
 Hispanic mothers also tend to be at the lower end of the scale, but are more spread out.
@@
 
-We see that brths for black and hispanic mothers at age 20-24 are only double the births of their teenaged counterparts, while white mothers tend to be as much as 5times the number of their teenaged counterparts, Therefore, teenaged motherhood in blakc and hispanic women may be more likelty.
+We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.
 
-In contrast, the distribution for the total population of women for the past 20 years seems uniformal, with alsmost all years having their own value.
-This makes sense as the population of a nation either increases or decreases, rarely staying the same.
+In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value.
+This is because the population of a nation either increases or decreases, rarely staying the same.
 Due to this we can say that the mean and median population are approximately the same.
-However there are two years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year.
+However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year.
 From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase.
-This increase in the population could be due to increased amounts of immigration, but is most likely due to the increased lifespan of the elderly to offset the birthrates.
-The two dips corespind to the previous histogram of two years with the same value.
+This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates.
+The two dips in the plot correspond to the previous histogram of years with similar value.
 
-```{r}
+```{r, include=FALSE}
 birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519
@@
 
-```{r}
+```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020"}
 birthhist %>%
   pivot_longer(cols = c("white15", "black15","hisp15"),
-               names_to = "group",
+               names_to = "Race",
                values_to = "value"
                ) %>%
-  ggplot(data =., aes(x=value, color = group, fill=group)) +
+  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
   geom_histogram()+ 
   theme_minimal()+
-  xlab("Numbe of births") +
+  xlab("Number of births") +
   ylab("Number of Years")
@@
 ```
-```{r}
+```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020" }
 birthhist %>%
   pivot_longer(cols = c("white20", "black20","hisp20"),
-               names_to = "group",
+               names_to = "Race",
                values_to = "value"
                ) %>%
-  ggplot(data =., aes(x=value, color = group, fill=group)) +
+  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
   geom_histogram()+ 
@@
 
-```{r}
+```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
 par(mfrow=c(2,2))
@@
 
-grid.arrange(pop_plot,pop_plot1, ncol =2)
+grid.arrange(pop_plot,pop_plot1, nrow =2)
 ```
@@
 
-(Figure \@ref(fig:figthree)) shows the trend of births for the two age groups in a 20 year period.
-Millenials would be between ages 15-24 in 2000, and gen Z between ages 15-24 in 2020.
-We see that there has been a significant decrease in the birth rates, especialy in the 15-19 (teen pregnancy), which has dropped to half its 2000 levels.
-The 20-24 age grouped by 20 %.
-This shows the decrease in popualrity for birth at these ages.
-There was a baby boom in 2006 (cite), which is the reason for the slight increase in birth rates around that time.
+(Figure \@ref(fig:figfour)) shows the trend of births for the two age groups in a 20 year period.
+We see that there has been a significant decrease in the birth rates, especially in the 15-19 age group, dropping to half of its 2000 levels.
+This indicates a decrease in teenage pregnancy nationwide.
+The birth rates for the 20-24 age dropped by around 40% over the same time period.
+This may be because generation Z women are deciding to have children later in life.
+There was a baby boom in 2006 [@citecatholic], which is the reason for the slight increase in birth rates around that time.
 However overall birth rates for young women has dropped.
-According to () there is little reason to beleive these birth rates will increase any time soon.
-There are some possilbe explanations, including better sex ed for teenagers which led to a decrease in teen pregnancy, mroe women are being focused on their careers and are postponing births, or the increased acceptance of abortions if they can;t afford a child right now.Thesse declining birth rates could also be the cause of declining population (cite), (talk about immigrants due to delcining in discussion).
-Polcy to increase families can be made to increase the borthrates, and fixing the economic system.
+According to [@Caldwell2006] there is little reason to believe these birth rates will increase any time soon.
+There are some possible explanations behind this phenomenon, including better sex education for teenagers leading to a decrease in teen pregnancy, more women focusing on their careers and thus postponing births [@citeexecutive], or the increased acceptance of abortions.
+These declining birth rates could also be the cause of declining population.
+There are also economic factors correlated to this decline, including the rising cost of living.
 
-```{r}
+```{r figfour, echo = FALSE, fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020"}
+
 race %>%
@@
   theme_minimal() +
-  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
+  scale_x_continuous(breaks = seq(2000,2020,by = 2))+
   scale_y_continuous(breaks = seq(0,140,by = 20))+
-  labs(x = " ",
-       y = " ",
-       color = " ") +
+  labs(x = "Year",
+       y = "Birth Rate",
+       color = "Age group") +
   scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
@@
 ```
-(Figure \@ref(fig:figfour)) shows the trend of births for of different races for women of ages 15 to 19.
-We see that the birthrate for hispanic groups has dropped the hardest, going from about 62 births per 1000 population to just about 48 births for 1000 population.
-This can either be because the population has greatly increased, or that the number of births has decreased.
+
+(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years.
+We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population.
+This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased.
 The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000.
-The white birth rates however has taken the largest decrease however, going from about 125 births to about 65 births through the 20 years.
-We can see that while for the millenial population there was a wide spread in the birth rates for different races, for gen z they have much more close.
-This could be due to economic or social factors.
-As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children.
-White women's decrease could also be due to economic factors.
-Racism in healthcare and the decrease of 'adultification' of gen z black women could be the cause of the decline.
+The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years.
+We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation z they have converged more closely due to the decrease in birth rates for all races.
+This could be due to economic factors directly related to race, such as inuequality in the healthcare system [@Davis2018] .
+As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
 
-```{r}
+```{r figfive, echo = FALSE, fig.cap = "Trend in Birth rates by Race from 2000 to 2020"}
 birthhist %>%
@@
   theme_minimal() +
-  scale_x_continuous(breaks = seq(1990,2020, by=5))+
-  labs(x = " ",
-       y = " ",
-       color = " ") +
+  scale_x_continuous(breaks = seq(2000,2020, by=2))+
+  labs(x = "Year",
+       y = "Birth Rate",
+       color = "Race") +
   scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", "White, non-Hispanic")) +
@@
 ```
-(Figure \@ref(fig:figfive)) shows the average number of births for those aged 15-24 for every state from 2000 to 2019. The color of the points indicates the level of the employment rate.
+(Table \@ref(fig:tabone)) shows the contribution of our main groups to the total decline in birth rates from 2000 to 2019.
+It also shows each group's share of the total female population for those years.
+We see that white women in both age groups have contributed the most to the decline, with a total of 18.2% of the total decrease.
+White 15 to 19 year olds have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011].
+White 20 to 24 year olds have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2017].
+Black women have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population.
+Black women are disproportionately poorer, and low economic status leads to higher teenage pregnancy rates [@Gold2001].
+The contribution of Hispanic women of ages 20 to 24 is less than their teenage counterparts, showing that there is not a large decrease in their birth rates as opposed to white women.
+They are also more likely to suffer from disproportional poverty.
+We see that white womens' share of the population from 2000 to 2019 has decreased due to the large decrease in birth rates.
+However black and hispanic womens' share of the population over this time period has either increased or not changed, likely due to increasing immigration from these two races over the past 20 years.
 
-We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier.
-However between the years of approximately 2003 and 2006, there was an increase in birth rates.
-According to (https://www.catholicnewsagency.com/news/11476/us-birth-rate-spikes-in-2006-creating-boomlet), in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom.
-At this point, unemployment rates were relatively low as signified by the darker blue points between 2000 and 2007.
-However, this trend doesn’t continue past 2006 as there seems to be a large decrease in births which continues until 2010 and becomes less severe between 2010 and 2013.
-During these times, we observe lighter blue points signifying that unemployment was higher than previous years, approximately around the 6-8% range.
-The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births.
-Between 2013 and 2014 the births increase, however following 2014 they begin to decrease further each year with low unemployment rates.
-Although the trends seen between 2000-2013 make sense with the changing unemployment rates, the trends from 2013 onwards tell us that there are other variables likely affecting birth rates as well.
-
-```{r}
-pro = group_by(policy, year) %>% select(year, unemp)
-tew <- summarise(pro, Unemployment = mean(unemp)) %>% select(Unemployment)
-
-correct <- race
-
-correct <- correct %>% filter(year >= 2001 & year < 2020)
-graph <- cbind(correct, tew)
-
-g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + labs(x = "Year", y = "Average Birth Rate") + guides(fill = guide_legend(title = "Unemployment Rate"))
-
-g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + labs(x= "Year", y = "Average Birth Rate") + guides(fill = guide_legend(title = "Unemployment Rate"))
-
-grid.arrange(g1519,g2024, nrow =2)
-```
-
-(Figure \@ref(fig:figsix)) shows the percentage decline in birth rates for the 20-34 (young women) age group by state from 2000 to 2019.
-We see that throughout the US there has been a decline in birth rates fro most states, with some states having more severe decline than others.
-We see that states on the west coast tend to have a larger decline than states on the east coast, with california having very high decline in birth rates, while that of North Dakota is nearly 0 change.
-This could be attriuted to the wwest coast being more liberal and refusing of gender roles for woman such that they dontnhave children as early or at all (cite).
-Compare that to east coast that tends to be more conservatove, religious or old days type, who encourage the standards of early marriage, especially in the southern and midwestern states of the US that more accept gender norms and refusal of new changes in culture (cite) .
-The exception seems to be Utah, which is consevrative religious due to the mormonism culture but has the largest decline in birth rates.
-This could be due to the downward turn of the Utahn economy (cite).
-however, the general trend still holds.
-Pressuring of women in religious areas and less sex ed as accuatomed ton those areas could also be the cause of the fewer decrease in south and midwest.
-
-```{r}
-map = birth1 %>%
-  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
-  filter(!stname %in% c("AB","BC","MB","NB","NS","ON","QC","SK","XX")) %>%
-  group_by(stname, year) %>%
-  summarize(sum = sum(numbirth2034), .groups = "drop")
-```
-
-```{r}
-birth2 = read_dta("age_race_comp_seer.dta")
-map2 = birth2 %>%
-  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
-  group_by(stname, year) %>%
-  summarize(sumpop = sum(pop2034),.groups = "drop")
-```
-
-```{r}
-map01 = map %>%
-  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
-map02 = map2 %>%
-  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
-
-```
-
-```{r}
-mapjoin = cbind(map01,map02)
-
-mapjoin = mapjoin %>%
-  select(-c(4:5))%>%
-  mutate(birthrate = (sum/sumpop)*1000)
-
-```
-
-```{r}
-mapjoin2019 = mapjoin %>%
-  group_by(stname) %>%
-  slice_max(order_by = year) %>%
-  group_by(stname) %>%
-  summarize(average = mean(birthrate))
-```
-
-```{r}
-mapjoin2000 = mapjoin %>%
-  group_by(stname) %>%
-  slice_min(order_by = year) %>%
-  group_by(stname) %>%
-  summarize(average = mean(birthrate))
-```
-
-```{r}
-mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
-  mutate(decline = average.x - average.y) %>%
-  mutate(fips = fips(stname)) 
-```
-
-```{r, fig.cap = "Fall in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
-plot_usmap(data = mapdiff, values = "decline", regions = "states") + 
-  scale_fill_continuous(low = "red", high = "green", name = "Change in birth rates (%)", label = scales::comma) +
-  theme(panel.background=element_blank()) +
-  theme(legend.position = "right")
-```
-
-# Table 1
-
-```{r}
+```{r, include = FALSE}
 table = numbirth %>%
@@
   select(-c(3,11,18))
-```
 
-```{r}
 share_of_pop = fulltable %>%
@@
 
-```{r}
+```{r tabone, echo = FALSE}
 contribution = birthrates %>%
@@
 table1 %>%
-  knitr::kable(col.names = c("Group", "Relative contribution to birth rate decline (%)", "Share of female population 2000 (%)","Share of female population 2019 (%)"), caption = "Contribution of target groups to total decline in birth rates from 2000 to 2019")
+  knitr::kable(col.names = c("Group", "Contribution to decline (%)", "2000 Population share (%)","2019 Population Share (%)"), caption = "Contribution of demographics of interest to total decline in birth rates from 2000 to 2019")%>%
+  kable_styling(latex_options = "HOLD_position")
+```
+
+(Figure \@ref(fig:figsix)) shows the number of births for those aged 15-24 for every from 2000 to 2019, according to the year's unemployment rate.
+We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier.
+Between 2004 and 2005, the birth rate for the 15-19 group continued to decrease whereas the birth rate for the 20-24 age group began to increase.
+We see that the unemployment rate is low in this period.
+According to [@citecatholic], in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom.
+However after this, the birth rate began to quickly decline and unemployment rates were high, reaching 8% by 2008 and continuing until 2012.
+This is likely due to the great recession that occurred during this period, such that it is possible women could not afford bringing new children.
+The birth rate, while still decreasing for both age groups, becomes less severe from 2012 onward, and we see that unemployment reduces.
+The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births.
+However beyond 2014, despite the low unemployment rate, we see the birth rate continue to decrease.
+We note that the unemployment rate is more likely to affect women in the 20-24 age group compared to their younger counterparts, as they are more involved in the workforce.
+
+```{r figsix, echo = FALSE, fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020" }
+pro = group_by(policy, year) %>% select(year, unemp)
+tew <- summarise(pro, Unemployment = mean(unemp)) %>% select(Unemployment)
+
+correct <- race
+
+correct <- correct %>% filter(year >= 2001 & year < 2020)
+graph <- cbind(correct, tew)
+
+g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x = "Year", y = "Birth Rate for Ages 15-19") +
+  labs(color= "Unemployment Rate (%)")
+
+g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x= "Year", y = "Birth Rate for Ages 20-24") +
+  labs(color= "Unemployment Rate (%)")
+
+grid.arrange(g1519,g2024, nrow =2)
+```
+(Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically the the 20-34 age group, by state from 2000 to 2019.
+We see that throughout the U.S there has been a decline in birth rates in most states, with some states having a more severe decline than others.
+We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, while those of North Dakota have seen little change.
+This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering  gender roles, such that women have children later or not at all.
+In comparison the east coast, midwestern and southern states have less severe changes in their birth rates, with some having no change or a slight increase in birth rates.
+Coincidentally, these states tend to be more conservative or religious, so they accept gender norms and refusal of new changes in culture.
+The exception seems to be Utah, which is conservative and religious due to Mormonism but has the largest decline in birth rates.
+However, the general trend still holds, and as some states are more costly to live in than others, this could also explain the differences.
+
+```{r, include = FALSE}
+map = birth1 %>%
+  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
+  filter(!stname %in% c("AB","BC","MB","NB","NS","ON","QC","SK","XX")) %>%
+  group_by(stname, year) %>%
+  summarize(sum = sum(numbirth2034), .groups = "drop")
+
+map2 = birth2 %>%
+  filter((year >= 1999 & year <= 2000) | (year>=2018 & year<=2019)) %>%
+  group_by(stname, year) %>%
+  summarize(sumpop = sum(pop2034),.groups = "drop")
+
+map01 = map %>%
+  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
+map02 = map2 %>%
+  mutate(year = ifelse(year >= 1999 & year <= 2000, 2000, 2019))
+
+mapjoin = cbind(map01,map02)
+
+mapjoin = mapjoin %>%
+  select(-c(4:5))%>%
+  mutate(birthrate = (sum/sumpop)*1000)
+
+mapjoin2019 = mapjoin %>%
+  group_by(stname) %>%
+  slice_max(order_by = year) %>%
+  group_by(stname) %>%
+  summarize(average = mean(birthrate))
+
+mapjoin2000 = mapjoin %>%
+  group_by(stname) %>%
+  slice_min(order_by = year) %>%
+  group_by(stname) %>%
+  summarize(average = mean(birthrate))
+
+mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
+  mutate(decline = average.x - average.y) %>%
+  mutate(fips = fips(stname)) 
+```
+
+```{r figseven, fig.cap = "Decline in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
+plot_usmap(data = mapdiff, values = "decline", regions = "states") + 
+  scale_fill_continuous(low = "red", high = "green", name = "Change in birth rates (%)", label = scales::comma) +
+  theme(panel.background=element_blank()) +
+  theme(legend.position = "right")
 ```
@@
 (Table \@ref(fig:tabtwo)) compares average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019.
-The mean student debt is total debt while the mean rent is monthly rent.
-Due to the data provided, the data corresponding to 2004 is representative of the early 2000’s.
-All these variables are related tpo social and economic factoes of the united states which may explain the change in birth rates.
-We see that the average student debt has almost doubled in 2019, the mean monthly rent has increased signifcantly, the percentage of people who believe religion is important has decreased, and the monthly childcare costs have also significantly increased.
-These costs have all risen past inflation, indicating that it is now more expensive to live in the us in 2019 than it is in 2000.
-As gen z mothers want to be more financially comfortable before having kids, these could cause them to postpone having kids, or not having as many kids as its too expendice.
+The mean student debt is lifetime debt while the mean rent is monthly rent.
+All these variables are social and economic attributes of the U.S, which may explain the change in birth rates.
+We see that the average student debt has almost doubled from 2004 to 2019, the mean monthly rent has increased significantly, the percentage of people who believe religion is important has decreased, and the monthly childcare costs have also significantly increased.
+Millennial women were more religious than their younger counterparts, which may have influenced their decisions to have children relatively young.
+We notice that these costs have all risen past inflation, indicating that it is now more expensive to live in the U.S in 2019 than it was in the early 2000s.
+As generation Z mothers want to be more financially comfortable before having children, this causes them to postpone having kids, or not having as many children due to expense.
+Expensive childcare and lack of affordable living has caused lower fertility rates [@citenargund].
 
-```{r}
+```{r tabtwo,echo = FALSE }
 studentdebt = student_debt %>%
@@
 explanatory = explanatory_table %>%
-  select(-c(1, 3,5,7))
+  select(-c(3,5,7))
 
 explanatory %>%
-  knitr::kable(col.names = c("Mean Student Debt", "Mean rent", "Percentage important","Mean childcosts"), caption = "Changes in possible explanatory variables") 
+  knitr::kable(col.names = c("Year","Student Debt", "Rent", "Religious Importance","Child Care Cost"), caption = "Changes in Variables that may Explain Decline in Birth Rates")%>% 
+  kable_styling(latex_options = "HOLD_position") 
 ```
+
 # Discussion
-The notion of God saying “Be fruitful and increase in number; multiply on the Earth” has changed throughout time.
-Many Millennials have parents who are Baby Boomers and they have expressed to their children that it’s important to think for themselves.
-This results in a “do-it- yourself” attitude toward religion.
-Millennials due to this have more focus on themselves in terms of having a good education and better job.
-Due to this, women are more career oriented.
-Before 2000s, women didn’t have many choices regarding these situations and used to marry at an early age and have kids.
-They used to take care of the family and do household chores while men used to earn by working outside.
-Since then, there has been a major change which has seen
-women making progress in all sorts of field such as sports, business or becoming leaders of their respected countries.
-Having a kid result in more responsibility being taken up which restricts things like working hours.
-Women having to juggle both work and home creates difficulties for themselves which can have major effect on their health.
+
+## Bias and Ethical Concerns
+
+As this data is based on people, there are likely to be some ethical implications that could limit the true accuracy of our data.
+
+A problem with these datasets and the original sources as well is that in terms of gender, the authors have only looked at births that happen through a female.
+The issue with this is that it does not account for the trans and non-binary population who are still able to get pregnant and give birth to children.
+As this group of people and their experiences are not included in the original study, it hinders both the accuracy and the main points that the author intended to make.
+Non binary people are systematically removed from society in many official and unofficial processes [@Tabaac2017], in such a way that the official U.S census even in 2020 did not include factor variables for non binary or other genders.
+Even if the data did include an 'other' option, that is grouping a lot of different gender identities into one large group; genders that are  completely different from each other.
+This would then make analysis confusing as different gender identities may have different experiences [@Tabaac2017].
+The datasets also did not include transgender men, who can also get pregnant, bringing the worrisome question of if pregnant transgender men were listed as 'male' in the U.S census, why the datasets were filtered by female only.
+
+Adding to this, non hispanic black and white people got their own group but hispanic black and white people did not, as they were all grouped into one race.
+This is inaccurate as hispanics can be any race and although they are all within the hispanic ethnicity, the experiences and thus data for black hispanics will likely be a lot different than that of white hispanics due to racism within the hispanic community [@Haywood2017].
+This problem may also arise because most of society sees all hispanics as mixed indigenous and european, thus erasing black and indigenous hispanic experiences.
+
+The NCHS dataset did not include undocumented immigrants in the information on birth and population.
+This is because undocumented mothers are much more likely to give birth at home or less formal places, due to fear of eviction from the country.
+Thus they often do not have birth certificates, or have inaccurate birth certificates.
+As these records are used for planning infrastructure and amenities to benefit the populace, not including them perpetuates their exclusion (many of whom pay taxes [@citetax]) from society and societal improvements.
+This perpetuates discrimination against them as they are also misused for labor, and targeted by ICE.
+ALthough census information does include everybody irregardless of status, many undocumented immigrants also fear coming out due to fear of being extrajudicially removed.
+This census data is also used for funding in particular areas and for infrastructure improvement, so these inaccurate results mean that not only are their experiences being left out, but areas where there are many undocumented are not getting the funding required.
+
+## Data collection concerns
+
+While the datasets included information on race, there are possible problems in data collection due to the subjectivity of race.
+There was no option for mixed race, so it is assumed that every person in the data were classified according to three arbitrary groups.
+For example, the sources did not provide at what point someone was no longer considered to be black, or if white hispanics were classified under white or under hispanic.
+What may be considered white or considered black may also change depending on the state or the community; i.e a mixed race person in a black community would be considered biracial, but in a white community may be considered black.
+If during data collection, the sources decided on self identification of survey takers, this could lead to inaccurate results.
+However if the data collectors decided to choose the race for themselves, this could strip people of their ethnic identity.
+As previously stated, undocumented immigrants were not included in the birth rate data.
+As they make up a significant portion of the country, the results presented in the data are likely very different from the true situation of women living in the country.
+The Current Population Survey did not record the data for child care costs for some midwestern states, and the reason for doing so was not stated.
+As the states which were excluded make up a decent portion of the U.S population and tend to have low costs of living than the national average, this could have led to a higher nationwide average child care cost in the report than is factual.
+
+## Socio-cultural Causes and Impacts of Declining Birth Rates
+
+### Stances on Women in society
+
+In [@Hayford2008], women took a survey on their adherence to religion and based on their answer, it shows those who answered 'very' have higher fertility than those that said 'somewhat' or 'not'.
+The article uses data from the 2002 National Survey of Family Growth (NSFG) to show the results.
+Factors such as unwanted fertility, age at childbearing, or degree of fertility postponement seem not to contribute to religiosity differentials in fertility.The article states how studies consistently find higher fertility among more religiously active women across denominations, although the magnitude of the effect varies with the measures used for fertility and for religiosity.
+The results from the survey show how women who value religion higher in their daily life have more children compared to those who value their religion somewhat to none.
+This shows how being religious can have a direct correlation to a higher fertility which causes more birth.
+Society in general has become less religious which is shown in the declining birth rates.
+[@citeusatrends] shows how marriage for women in the 20 to 24 age group has decreased, showing that generation Z are choosing the marriage and children route at later stages in their lives.
+In 1970, about 36% of the population in the age group did not get married compared to 2008 where 80% of the population in the age group were not married.
+Since women are marrying late, they are giving birth at a later stage in life compared to before.
+However it should be noted that the amount of women who have children without being married has increased in the past 20 years.
+
+The notion of female independence has become increasingly more popular in recent years.
+While millennial women are already career oriented, generation Z women are even more so.
+Women staying at home to take care of children, while still nowhere near as popular a notion as in the 1950s, was still an accepted norm in the 1990s and 2000s [@Maher2007].
+With the rise of feminist movements in the past two decades there has been a major change which has seen women making progress in all sorts of field such as sports, business or becoming leaders of their respected countries.
+The choice to not have children is not as shameful in society compared to 20 years ago.
+Having children results in more responsibility being taken up which can restrict career advancements [@citeexecutive], as western culture still has expectations that women do  childcare when men go to work, so women often feel they can't do both [@Maher2007].
+The cost of delaying their careers to have children also increases as they get deeper into their profession.
+Women who do 'have the best of both worlds' need to juggle both work and home, creating difficulties for themselves which can have major effect on their health.
 Maternity leave in the US on an average result in 10 weeks while the paid leave is 8 weeks.
-This can cause an issue in terms of finance as living expenses has increased from 2000s and especially since the recession in 2008.
+This can cause an issue in terms of finance as living expenses has increased from the 2000s, especially since the recession in 2008.
 Two-parent households are on the decline in the United States as divorce, remarriage and cohabitation are on the rise.
-And families are smaller now, both due to the growth of single- parent households and the drop in fertility.
-Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed.
-The reason of decrease in births because women are becoming mothers later in life and the reasons are tied to increases in educational attainment, growing labour force participation and delays in marriage.
-The average age of mothers at first birth is 27, up from 23 in 2010.
-https://www.pewresearch.org/fact-tank/2016/01/08/qa-why-millennials-are-less-religious-than-
-older-americans/
+Families are smaller now, both due to the growth of single- parent households and the drop in fertility.
+The rise of LGBT families has also led to an increase in adoptions and decrease in birth rates.
+Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed.The increase in women achieving higher education [@citeusatrends] has also led to a pause to the child rearing stage for many, which has also caused a decrease in the birth rates for young women as this causes women to have children later in life after they are done with education.The causes of these decrease in births are tied to increases in educational attainment, growing labor force participation.
 
-https://www.pewresearch.org/social-trends/2011/10/12/in-a-down-economy-fewer-births/
+### Politics
 
-## Politics
+White birth rates have decreased significantly, and the 15-24 groups contribute at more than 18% of declining birth rates.
+This decrease in white birth rates is a well known talking point of the far right, which is used to harass people of color, especially mothers.
+This concern over the white birth rate decline (often coded as 'True Americans') and the notion of 'being replaced' has led to harmful anti-immigration policies, as well as the rise of far right hate groups throughout the country [@Chermak2013].The increase and proliferation of far right ideologies that white birth rates are decreasing leading to a less white country are fueled in part by the way that these statistics can be shown; without the proper context to show why they are decreasing.
+Many pro-life advocates have also noted declining birth rates as a major reason for their antipathy towards abortion, which has caused an increase in attacks on planned parenthood clinics as they are believed to be a symptom of societal decay.
+[@Primrose].
 
-White birth rates have decreased significantly, and the 15-24 groups contribute quite a lot at more than 11% of declining birth rates.
-Decrease in white women borth rates often toutued by far right, which has used to harass people of color, into a "white genocide" stuff.
-This has also caused the increase of attacks on planned parenthood clinics as they are believed to have caused the decrease in these birth rates.
-A declining population is one of the main reasons people are against abortion(cite), and the increase and proloferation of far right ideologoies that white birth rates are decreaing leading to a les white country are fuled in part by the way that these statistcis can be showsn, and withiut the proper context to show why they are decreasing.
-As birth rates for poc has not decrwed as much it could also perpetuate harmful stereotpyes on poc women having too many hcildren, too young, or are having lazy kids on welfare.
+A decline in birth rates has led to a decline in population growth and thus the need for immigrants.
+This decrease in birth rates has led to the push for immigration policies, not just in the U.S, but in other industrialized societies.
+This has led to political controversy based on immigrants taking jobs from citizens.
+As the economic recession devastated many sectors of the economy and areas of the country, an increase in immigrants taking jobs in the U.S promotes feelings of animosity towards them.
+These feelings of animosity are also fuled by immigrants tendency to have more children than the citizen populations.
+However these arguments often do not take into account the fact that second and third generation immigrants tend to reduce their birth rates to match the rest of the population [@immigbirth].
+This decrease in birth rates is thus used both  as a justification for immigration and job outsourcing, as well as an 'invasion crisis'.
 
-This decrease in birth rates has led to the push for immigration policies, not jutst here but in other industrialized societies.
-This has led to political cotneoversy on having too many imigrants which takes jobs from the native populations.
-This can then cause demographic decline as it alreadty has in the dust belt(), which  further cotnributes to declining birth rates as things become too expensice (cite).
-These then promote feelings of hatreds towards immigrants,both for "Stealing" jobs, as well as usually having more children than the native populations.
-These decrease i nbirth rates can thus be used by both capitalists wanting to use immigration, or by right winged groups as a crisis that immigrants are coming and to have more kids.
+### Race
 
-## Unemployment
+Women of color such as black and hispanic women are less likely to receive adequate natal care and resources compared to their white counterparts.
+This has caused the decreased fertility rates and higher negative birth outcomes for black women [@Davis2018], [@Yang2003] which could decrease their number of births and thus their birth rates.
+Shifts in fertility have disproportionately happened to white women and the higher educated, who also tend to be white [@Yang2003].
+Hispanic women also have less than stellar access to fertility health care, as well as have lower IVF success rates.
+Black people are concentrated in the south, which has very high levels of teenage pregnancy [@citeteen] and low level of resources on sexual education, so they are disproportionately affected.
+This may be why they did not contribute much to declining birth rates in the report.
+Black teenagers are also less likely to have access to quality education,careers and outside opportunities[@citeusatrends], which is a major factor to the decline in birth rates.
+Hispanics also are affected by many of the same problems; lack of funding and proper education in their communities.
+As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.
+This shows yet another way in which women of color are negatively treated in society.
+
+## Economic Causes and Impacts of Declining Birth Rates
+
+Table 1 examined economic factors that may play a role in the living standards and motherhood decision of young women.
+Student debt, rent and childcare costs increased from the early 2000s to late 2010s.
+Most students between the ages of 15-24 do not have a steady source of income and due to the rising expenses, it becomes difficult for young women to properly raise a child.
+The cost of housing in the United States has doubled which plays a huge role in the decision to have a child [@citewomenhousing] .
+Increasing rent and lack of social safety nets due to the U.S's individualistic tendencies contributes to why generation Z couples have second thoughts about having a child[@citerentbabies].
+In some cities in 2022, rent is increasing by up to 40% [@cite40] and increases like that  push women against having children in less favorable positions [@livingston2020].
+In a similar way, the rise of student debt in the United States is also leaving young adults, more so young women, with less money to care for their families.
+According to [@citewomenloans], women in the United States hold approximately 67% of student loans and many of them state that their debt influences their decision to have a child.
+When considering these two costs, and considering the fact that there are child care expenses to tend to as well, it becomes apparent how expensive it is becoming to have a child.
+According to [@citechildcarecost], a dollar increase in weekly childcare costs implies a 2% decrease in birth rate.
+Along with this, a dollar increase in childcare cost also means that the rate of entering employment decreases by 3%.
+From this, one can see how there is not only a correlation between increasing childcare costs and unemployment but also a correlation with increasing childcare costs and a decreasing birth rate.In addition to these, the rise of urbanization has caused a decrease in birth rates [@citenargund], and urbanization is directly correlated to increase in rent.
+
+### Unemployment
 While unemployment rates were a factor which affected birth rates between 2000-2013, there are other variables at play as well.
@@
 
-## STudent debt, childcare costs, religion (by state?), rent
-
-NOTE: for the childcare table, the values for some states are missing. discuss this either in the data or in weaknesses/nextsteps
-
 what caused the great decline in hispanic borth rate decrease?
@@
 Limitations: we couldnt get the exact age groups fror some variables, so we had to take the ages of all women or the n20-34 age groups to approximate, so we may not get the accuarte results as the general trend may not apply to the sepcific trend of teenaged and young mothers.
-
-# Bias and Ethical Concerns
-
-It is important to note that the fact that this data is based on people, there are likely to be some ethical implications that could limit the true accuracy of our data.
-Another consideration we should make a note of are potential biases that we can observe from the data.
-
-A problem with these datasets and the original sources as well is that in terms of gender, it seems that they’ve only looked at births that happen through a female.
-The issue with this is that it does not account for the trans and non-binary population who are still able to get pregnant and give birth to children.
-This is another group of people that are not included in this study which hinders the accuracy of results.
-Non binary people are systematically removed from society in many offical processes and unofficial.
-For the fact that the official us census even in 2020 did not include a non binary or other gender is also bad.
-Even if they did include an other, that is grouping a lot of different gender identities into one huge group, that may be completely different fromeach other.
-This would then make analysis confusing as different gender identities may have different experiences.
-The datasets also also did not include transgender men, who can also get pregnant.
-This invalidates transgender men's existences by calling them women.
-
-Adding to this, non hispanic black and whites got their own group but hispanics did not, they were all grouped into one.
-This is an error as hispanics can also be any race and althought they are within the hispanic ethnicity, the experiences and data for black hispanics will likely be a lot different than that of white hispanics due to racism within the hispanic community (cite).
-This problem may also be because most of society sees all hispanics as mestizo, thus erasing castizo and afro-hispanic experiences.
-
-The sources for birth rates did not include illegal immigrants in their info on birth and population.
-This is because illegal mothers are much more likely to give birth at home or less formal places, due to fear of eviction from the country.
-Thus they often do not have birth certifictaes, or have inaccurate birth certificates.
-As these records are used for *put what they were used for in the introduction*, not including them perpetuates rhe discultison of these people from society and societal improvements, many of whom pay taxes, as they are also misused for labour, in prisons, and by ICE (cite).
-ALthough census information does include everybody irregardless of status, many illegals also fear coming otu due to fear of being extrajudicially removed, which has happened before (cite).
-this census data is also used for funding in particular areas and for infrastruture improvement, so these inaccurate results mean that not only are their experiences being drowned out, but areas where there are many of them are not getting the funding required.
-
-# Data collection error
-problems in measurement of what is black etc.
-did they decide on self identification, which could lead to wholely inacurate results, or did they decide for themselves, wbich could strip people of their gender identity?
-
-Adding onto this, they did not record the data for some childcare costs, for the childcare dataset and so we do not have the full picture.
-
-# race
-south poorest area in the us, have higher births due to les education, as previously said.
-As black people are cocentrated there (cite), these lack of resources disproprtionately affect them and so they have higher birth rates so thats why they dont have as much contribution to declining birth rates.
-Whites who have more resources and better schools that can teach sex ed have less birth rates due to that better education.
-Hispanics also have less resoruces and less good schools.
-They also have les outside opportunties like school and career that could postpone birth at those ages.
-Black poor so cant find good doctors for natal care, which could affect birth rates.
-
-*look at two links sent by Arsh at 7:58 pm and 8:28pm*
 
@@
 However as gen z is now more liberal than their millenial counterparts they are less likely to have more kids nationwide. 
+and problems with birth control and abortion in more conservative states.
+ Pressuring of women in religious areas and less sex ed as accuatomed ton those areas could also be the cause of the fewer decrease in south and midwest.
+
+Why do some states have a larger decrease in birth rates compared to others?
+
+The rent is a huge factor when it comes to why some states were hit harder than others.
+Rent on a two-bedroom apartment in Henderson, Nevada has risen 23 percent to nearly $1,600 a month.
+This makes it almost impossible for a single mother to raise her kids and hence it causes them to move to a different state where the rent is rather lower.
+North Dakoda is a prime example of why it is beneficial to live there and raise your family.
+The average rent in the state is around $850 which is roughly half of what it is in Nevada.
+This is why they choose to move out of that state causing that state to have a higher decrease.
+
+https://worldpopulationreview.com/state-rankings/average-rent-by-state
+
+https://www.washingtonpost.com/business/2022/01/30/rent-inflation-housing/
 
@@
 Teenaged birth rates have decreased in the past 20 years for gen z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted ven with the cultural change.
-This is due to lack of opportunities and that is all they know, as well as higher rates of poverty and low education.
+This is due to lack of opportunities and that is all they know, as well as higher rates of poverty and low education [@Gold2001].
 While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back.
@@
 It is knowsn to also shame women for having sexual thought,s cotnributing to a misogynistic society(cite)
+they have liitle change of advancement (doi: 10.1257/jep.26.2.141)
 
-# Economic changes
+# Conclusion:
 
-Costs are increasing, higher student debt, and lack of safety nets due to the country's indivualistic nature means that gen z women are not as eager tp have kids as millenial women. 
-Table 1 examined other economic factors that played a role in the lives of young women.
-We noticed student debt, rent and childcare costs increased from the early 2000s to late 2010s.
-Most students between the ages of 15-24 do not have a steady source of income and due to the rising expenses, it becomes difficult for young women to properly raise a child.
-The cost of housing in the United States has doubled which plays a huge role in the decision to have a child.
-(https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4685765/).
-The increased rent seen from Table 1 is the main reason that families have second thoughts about having a child.
-In some cities, rent is increasing by up to 40% and increases like that would definitely put families wanting to have a child in a less favorable position.
-(https://www.theguardian.com/us-news/2022/feb/16/renters-rent-increases-us-lease) In a similar way, the rise of student debt in the United States is also leaving young adults with less money to care for their families.
-According to studies, women in the United States hold approximately 67% of student and many of them state that their debt influences their decision to have a child(https://blog.futurefamily.com/press-release-student-dept-holds-back-american-fertility-rates/).
-When considering these two costs, and considering the fact that there are child care expenses to tend to as well, one can see how expensive it is becoming to have a child.
-According to a study from David M.
-Blau and Phillip K.
-Robbins, a dollar increase in weekly childcare costs implies a 2% decrease in birth rate.
-(https://www-jstor-org.myaccess.library.utoronto.ca/stable/2061526?seq=8#metadata_info_tab_contents) Along with this, a dollar increase in childcare cost also means that the rate of entering employment decreases by 3%.
-From this, one can see how there is not only a correlation between increasing childcare costs and unemployment but also a correlation with increasing childcare costs and a decreasing birth rate.
-
-# Culture changes
-
-Insudtrialized societies have led to a further breakdown of conservative beliefs and women becoming more idnependent.
-In these days more women have a choice than 20 years ago on if they want to have children at all.
-WOmen are also  choosing to enter careers and focus on child care in their later years and not their early 20s.
-This can be seen as women in states with more stay at home moms have higher 20-24 birth rates.
-Women are also not marrying just after college anymore (cite).
-the rise of lgbtq families has also led to more adoptions or not wanting to have chidren in general.
-Women are also getting married later (cite us economic and social trends since 2000) and the gen z having sex is decreasing (cite), which also leads to further birthrate decline.
-
+It is necessary... (from)
 # Weakness / Next Steps
@@
 
+The years for millenial and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
+Thus the years chosen for this report are completely arbitrary.
+
 As more women do not want to give birth and opt for adoption, that could lead to further decreases in birth rates in the future.
-We wuld like to make a graph that shows the relaitionships betwen adiptions and birth rates Adoption could be one reason for impact on the map.
+In further research we would like to make a graph that shows the relationship between adoptions and birth rates.
+There has also been much controversy in the past two decades over the falling sperm counts in men [@JohnAitken2013].
+We are interested in if this has had any effect on birth rates, and in future research we would like to explore the relationship between them.
 
-Could falling testosterone rates(cite) also attribute to these? we would like to make a table that shows the change in the 20 years.
+for next steps: Could falling testosterone rates(cite) also attribute to these? we would like to make a table that shows the change in the 20 years.
+
+# References

### 2022-02-27 Main Code repository with data and paper/3 Graph Replication.R

@@ -1,7 +1,7 @@
 # Loading in datasets
 
-birth = read_csv("fig_1.csv")
-race = read_csv("figs_2a_2b.csv")
+birth = read_csv("Inputs/Data/fig_1.csv")
+race = read_csv("Inputs/Data/figs_2a_2b.csv")
 write_csv(birth, "Replication-Birth Rates/birth.csv")
 write_csv(race, "Replication-Birth Rates/race.csv")
 

### 2022-02-27 Scripts/Rscript.R

@@ -15,22 +15,22 @@ library(haven)
 library(tidyverse)
 
 # Reading in the data in the raw data.
-birth = read_csv("fig_1.csv")
-numbirth = read_csv("numbirths_2001_2019.csv")
-student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
-rent = read_dta("longdiff-rentcost-0419.dta")
-religion = read_dta("long-diff-religion.dta")
-childcare = read_dta("child_care_costs_under12_0408_1519.dta")
-birth2 = read_dta("age_race_comp_seer.dta")
-birth1 = read_dta("nchs_births_pop_1990_2019.dta")
-race = read_csv("figs_2a_2b.csv")
-policy = read_dta("policyvars01_19.dta")
+birth = read_csv("Inputs/Data/fig_1.csv")
+numbirth = read_csv("Inputs/Data/numbirths_2001_2019.csv")
+student_debt = read_dta("Inputs/Data/nyfed-student-debt-per-capita-0419.dta")
+rent = read_dta("Inputs/Data/longdiff-rentcost-0419.dta")
+religion = read_dta("Inputs/Data/long-diff-religion.dta")
+childcare = read_dta("Inputs/Data/child_care_costs_under12_0408_1519.dta")
+birth2 = read_dta("Inputs/Data/age_race_comp_seer.dta")
+birth1 = read_dta("Inputs/Data/nchs_births_pop_1990_2019.dta")
+race = read_csv("Inputs/Data/figs_2a_2b.csv")
+policy = read_dta("Inputs/Data/policyvars01_19.dta")
 
 # create datasets from dta files
-
+write_csv(birth, "Inputs/Data/birth.csv")
 write_csv(numbirth, "Inputs/Data/numbirth.csv")
 write_csv(race, "Inputs/Data/race.csv")
-write_dta(policy, "Inputs/Data/policy.csv")
+write_csv(policy, "Inputs/Data/policy.csv")
 write_csv(student_debt, "Inputs/Data/student_debt.csv")
 write_csv(rent, "Inputs/Data/rent.csv")
 write_csv(religion, "Inputs/Data/religion.csv")

### 2022-02-27 Output/paper 2.Rmd
adding changes

@@
 ---
-better economic-industrial society to expensive for children, industrial has low birth rates
-adoption of more freedom for women?
-abortion? restricting of abortion could lead to higher birth rats, but to the detriment of choice
-
-Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new
-
-why is hispanic decline so large
-harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants
-
-implications for teenaged mom is a good thing that the borth rates are decreasing?
-
-Check how teen pregnancy  and female enollment has increased.
-
-Its good black teen birth rate decreased (table 1), how can we increase this decrease?
-
-does birth control reduce usage for young mothers?
-we dont expect income ratio to affect as these are young moms anyway.
-religious change, specially in bible belt?
-
-possible bias: To what extent is random allocation ethical and fair?
-Some argue that shortages mean it is reasonable to randomly allocate, but that may depend on how linear the benefits are.
-It may also be difficult to establish definitions, and the power imbalance between those making these decisions and those being treated should be considered
 
@@
 A nation's birthrate for a particular year is the total number of live births per 1000 people for that year.
-In order to measure the birth rates and therefore fertility levels, as well as forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors.
+In order to measure the birth rates and fertility levels to forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors.
 The total population of the country is the total number of citizens and other residents in the country, regardless of legal status.
@@
 This information also breaks down the population by age, race, and socio-economic factors.
-In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographic of the mother.
-We are also interested in the relationship between  changes in economic, social and cultural aspects of the United states and the changes in these birth rates.
+In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographics of the mother.
+We are also interested in the relationship between the change in economic, social and cultural aspects of the United states and the change in these birth rates.
 
@@
 Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable].
-We first created histograms and scatterplots which showed the most important variables.
+We first created histograms and scatter plots which showed the most important variables.
 We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states.
@@
 The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry].
-This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their systematic discrimination in society.
-The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being "irresponsible" with their bodies.
+This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society.
+The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies.
 This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].
@@
 The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite].
-The data used by the paper was compiled by various sources such as the NCHS [@citenchs] CDC SEER [@citecdcseer], NBER Natality Database, and the Current Population Survey
-Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all).
+The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer],the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) , the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums].
 The paper and datasets were last accessed on February 16th, 2022.
 The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019.
-The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which was randomly sampled [@citenchs].
-The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer].
+The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, and this was randomly sampled [@citenchs].
+The national population data comes from the census, a long running survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer].
 The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics.
-These datasets were gotten from surveys or website data, so non-response was often recorded as 0 or as `NA`.
-The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity.
+These datasets were gotten from surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or estimated.The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity.
 Meanwhile, categorical data mainly appeared in the form of state.
@@
 We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information.
-We included the `race` dataset as it gave the main data of analysis  in the form of birth rates for different age groups and races.
-We included the `birth` dataset as it included the total birthrate for all women.
-We also included the `numbirth` dataset as it had information on birthrates for all the age groups combined.
-We used the `birth1` dataset as well as it had data on birth rates for younger age groups, as well as total population data and demographic data for each race and age combination.
-`birth2` had total population data for each age group and race-age combination.
-We used`policy`, `rent`, `student_debt`, `religion`, and `childcare` as they provided information on the changes in unemployment, rental cost, student debt, importance of religion and rise in child costs throughout the US, all of which are expected to influence birth rates to varying degrees.
+We included the datasets that had information on race, as this was needed to compare birth rates for different races.
+We also had datasets on total number of births and/or birth rates, both of which were used to analyze the decline in births.
+We had total population datasets, used to get the birth rates and our demographic's share of the population.
+We had datasets on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.
+We also included datasets that included information on government policy, rent costs, student debt, religious beliefs, and childcare costs.
+These datasets were used to signify a possible relationship between these variables and birth rates.
 
 We created several of our own datasets by modifying and joining variables.
-These datasets include `birthhist`, which was gotten from the `birth1` dataset by adding all the educational status for each age-race combination to get the total birth race for that comnination, and then compiling the values by state and year.
-We also got the `sumpop` variable by adding the total population for each state in a year.
-We then obtained `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `table1`, which was used to show how our target demographics contributed to birth rate decline, and `explanatory`, which was used to display the changes in possible explanatory variables over a period of time.
+One dataset was gotten by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year.
+We also got a variable for total population by adding the total population for each state in a year.
+We obtained the final dataset used to create the map of birth rate decline by combining various other datasets by year, and subtracting the birth rates of the later years by that of the earlier years.
+The final dataset that showed how our demographics of interest contributed to total birth rate decline was created by joining several datasets all which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions.
+The final dataset used to display the change in explanatory variables over time was gotten by combining the datasets, each of which explained the changes in a single variable.
 Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
@@
 We modified many variables for the report.
-We first modified the `birth1` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset.
-We did this by adding up all educational statuses for each race and age group combination, as each person whose birth rates was recorded also had their educational status recorded.
-The final variables were named `numbirthraceage`, i.e `numbirthwhite15` for births for the white 15-19 population.
-We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`.
-We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates.
-These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
+We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rates was recorded also had their educational status recorded.
+We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within the column of `group`.
+We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
+We obtained a variable which got the decline in birth rates from 2000 to 2019 by state, by creating dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.
 
-For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of births and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state.
-The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group.
-From those two new datasets, we changed the `year` column so that it only stated 2000 or 2019.
-We then joined those two datasets together and divided the number of births by the population to get the birth rate.
-We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and collected the mean of each value.
-To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.
-
-For the first table, we got the `table` dataset by grouping the `numbirth` dataset by year, beyond 2000, and then summing up the entire female population by state, which was stored as `sumbirth`.
-We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations.
-We named these variables for each combination `sumraceage`, e.g `sumblack20`.
-We filtered the `birth` dataset between 2000 and 2020.
-We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`.
-From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable.
-From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by multiplying the number of births for each race-age combination by 1000 and them dividing them by the total female population.
-These variables were named `raceagebrate`, e.g `b20brate`.
-We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` value for 2000 and multiplying by 200.
-We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.
-
-For the second table, we got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states.
-We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively.
-For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source.
-We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values plus `not_too_important` and `not_at_all_important`, and then getting the mean of all the states.
-Note that for these variables we needed to use 2004 as 2000 was not available.
+We got a variable for the total number of births by adding up all the births by state and then grouping them by year.
+We got a variable that stated the total population of each race-age combination by similar methods.
+we obtained a variable got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100.
+We had a variable that got contribution to total birth rate decline by subtracting 2019 values from the 2000 values and dividing that by the total birthrate value for 2000 and multiplying by 200.
+We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states.
+For the average debt, we multiplied by 10,000 as that is how it was shown in the original source.
+We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.Note that for these variables we needed to use 2004 as 2000 was not available.
 
 ```{r global_options, include=FALSE}
-
+# laoding in relevant libraries
 knitr::opts_chunk$set(fig.pos = 'H')
@@
 ```{r,include = FALSE}
+#loading in relevant datasets
 birth = read_csv(here::here("Inputs/Data/birth.csv"))
@@
 ```
+
 ## View of the Data
@@
 We are interested in the trend of birth rates for women ages 15-19 and 20-24 over a 20 year period.
-This shows the decline in popularity of young motherhood and thus the increase of the average birth age.
+The data shows the decline in popularity of young motherhood and thus the increase of the average birth age.
 This has ongoing implications for public social and economic policy.
@@
 We see that teenage births tend to be less popular, as the data seems slightly right skewed.
-This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012].
+This also means that the average number of births will be below the median number of births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012].
 We also see that black teenage mothers are less common, as they have the lowest births for most years.
 This is most likely because they are a much smaller segment of the population than the others.
-However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportiantely appearing as teenage mothers.
+However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportionately appearing as teenage mothers.
 Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year.
 White teenage mothers are the most common, due to them being the largest in population.
-They tend to have most of their distribution in the high areas towards the higher numbers.
+They tend to have most of their distribution towards the higher numbers.
 
@@
 Hispanic mothers also tend to be at the lower end of the scale, but are more spread out.
-White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.
-
-We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.
+White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.
 
@@
 This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates.
-The two dips in the plot correspond to the previous histogram of years with similar value.
+However, the two dips in the scatterplot correspond to the previous histogram of which had years with similar values.
 
 ```{r, include=FALSE}
+# getting total number of births for white people from 15 to 19 years old
 birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519
 
+# getting total number of births for black people from 15 to 19 years old  
 birth1$numbirthblack15 = birth1$numbirth_lesshs_blacknh_1519 + birth1$numbirth_educ_miss_blacknh_1519 + birth1$numbirth_gtc_blacknh_1519 + birth1$numbirth_somecoll_blacknh_1519 + birth1$numbirth_hsgrad_blacknh_1519 + birth1$numbirth_ba_blacknh_1519
 
+# getting total number of births for hispanics from 15 to 19 years old  
 birth1$numbirthhisp15 = birth1$numbirth_lesshs_hisp_1519 + birth1$numbirth_educ_miss_hisp_1519 + birth1$numbirth_gtc_hisp_1519 + birth1$numbirth_somecoll_hisp_1519 + birth1$numbirth_hsgrad_hisp_1519 + birth1$numbirth_ba_hisp_1519
 
+# getting total number of births for white people from 20 to 24 years old
 birth1$numbirthwhite20 = birth1$numbirth_lesshs_whitenh_2024 + birth1$numbirth_educ_miss_whitenh_2024 + birth1$numbirth_gtc_whitenh_2024 + birth1$numbirth_somecoll_whitenh_2024 + birth1$numbirth_hsgrad_whitenh_2024 + birth1$numbirth_ba_whitenh_2024
 
+# getting total number of births for black people from 20 to 24 years old
 birth1$numbirthblack20 = birth1$numbirth_lesshs_blacknh_2024 + birth1$numbirth_educ_miss_blacknh_2024 + birth1$numbirth_gtc_blacknh_2024 + birth1$numbirth_somecoll_blacknh_2024 + birth1$numbirth_hsgrad_blacknh_2024 + birth1$numbirth_ba_blacknh_2024
 
+# getting total number of births for hispanics from 20 to 24 years old  
 birth1$numbirthhisp20 = birth1$numbirth_lesshs_hisp_2024 + birth1$numbirth_educ_miss_hisp_2024 + birth1$numbirth_gtc_hisp_2024 + birth1$numbirth_somecoll_hisp_2024 + birth1$numbirth_hsgrad_hisp_2024 + birth1$numbirth_ba_hisp_2024
 
+# gets the total number of births by year, compiled nationally
 birthhist = birth1 %>%
@@
 
-```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020"}
+```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020", message = FALSE}
+
+# converts the columns to rows and creates the histogram
 birthhist %>%
@@
 ```
-```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020" }
+```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020", message = FALSE}
+
+# converts the columns to rows and creates the histogram
+
 birthhist %>%
@@
 ```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
-par(mfrow=c(2,2))
+
+# groups the total female population by year and compiles nationally, then creates the histogram
 pop_plot = birth2 %>%
@@
 
+# groups the total female population by year and compiles nationally, then creates the scatterplot  
 pop_plot1 = birth2 %>%
@@
 
+# putting two figures into one figure space
 grid.arrange(pop_plot,pop_plot1, nrow =2)
@@
 
+# converting the age column to rows, then creates a line graphing the the trend of the two age groups
 race %>%
@@
   theme_minimal() +
-  scale_x_continuous(breaks = seq(2000,2020,by = 2))+
-  scale_y_continuous(breaks = seq(0,140,by = 20))+
+  scale_x_continuous(breaks = seq(2000,2020,by = 2))+ # labeling by 2
+  scale_y_continuous(breaks = seq(0,140,by = 20))+ # labeling by 20
   labs(x = "Year",
@@
 The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years.
-We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation z they have converged more closely due to the decrease in birth rates for all races.
-This could be due to economic factors directly related to race, such as inuequality in the healthcare system [@Davis2018] .
+We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation Z they have converged more closely due to the decrease in birth rates for all races.
+This could be due to economic factors directly related to race, such as inequality in the healthcare system [@Davis2018] .
 As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
@@
 ```{r figfive, echo = FALSE, fig.cap = "Trend in Birth rates by Race from 2000 to 2020"}
+
+# gets the birth rate for each race by adding the number of births from the 15 to 19 and 20 to 24 age group for each race, then multiplying by 10,000 and dividing by the female population
+
 birthhist %>%
@@
 We see that white women in both age groups have contributed the most to the decline, with a total of 18.2% of the total decrease.
-White 15 to 19 year olds have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011].
-White 20 to 24 year olds have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2017].
-Black women have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population.
+White 15 to 19 year old women have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011].
+White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2017].
+Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population.
 Black women are disproportionately poorer, and low economic status leads to higher teenage pregnancy rates [@Gold2001].
@@
 ```{r, include = FALSE}
+
+# total number of births in the U.S 
 table = numbirth %>%
@@
 
+# total population for each race-age combination
 ageracepop = birth2 %>%
@@
 
+# filters birth rates beyond 2000
 birth = birth %>%
@@
 
+# joining the previous datasets into one
 fulltable = cbind(table,birthhist,ageracepop,birth)
@@
 
+# getting each race - age combination's share of the total population, then converting the columns to rows
 share_of_pop = fulltable %>%
@@
 
+# getting the birth rates for each race - age combination
 birthrates = fulltable %>%
@@
 ```{r tabone, echo = FALSE}
+
+# getting each race - age combination's total contribution to declining birth rates
 contribution = birthrates %>%
   pivot_wider(names_from = year, values_from = birthratevalue) 
-
 contribution$contribution = ((contribution$"2000" - contribution$"2019")/65.9)*200
 
+# getting the percentage of the population
 percentpop = share_of_pop %>%
   pivot_wider(names_from = year, values_from = value)
-
 percentpop$"2000" = percentpop$"2000"*100
@@
 
+# joining the two previous datasets to one and removing unwanted columns
 table1 = cbind(contribution,percentpop)
@@
 The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births.
-However beyond 2014, despite the low unemployment rate, we see the birth rate continue to decrease.
+However beyond 2014, despite the low unemployment rate, we see the birth rate continues to decrease.
 We note that the unemployment rate is more likely to affect women in the 20-24 age group compared to their younger counterparts, as they are more involved in the workforce.
@@
 ```{r figsix, echo = FALSE, fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020" }
+
+# grouping by year and policy, and then selecting relevant columns
 pro = group_by(policy, year) %>% select(year, unemp)
+
+# summarizing these values by year and policy, and getting the average unemployment
 tew <- summarise(pro, Unemployment = mean(unemp)) %>% select(Unemployment)
 
+# filtering by the relevant years and combining the datasets
 correct <- race
-
 correct <- correct %>% filter(year >= 2001 & year < 2020)
@@
 
+# creating scatterplot for the 15 to 19 age group
 g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x = "Year", y = "Birth Rate for Ages 15-19") +
@@
 
+# creating scatterplot for the 20 to 24 age group
 g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x= "Year", y = "Birth Rate for Ages 20-24") +
@@
 
+# putting two figures into one figure space
 grid.arrange(g1519,g2024, nrow =2)
@@
 We see that throughout the U.S there has been a decline in birth rates in most states, with some states having a more severe decline than others.
-We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, while those of North Dakota have seen little change.
+We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, and North Dakota having relatively little change.
 This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering  gender roles, such that women have children later or not at all.
@@
 ```{r, include = FALSE}
+
+# getting births for these two year ranges only, and grouping by year and state
 map = birth1 %>%
@@
 
+# getting the population for these two year ranges, and grouping by year and state
 map2 = birth2 %>%
@@
 
+# making the year variables have only two options
 map01 = map %>%
@@
 
+# joining the two variables
 mapjoin = cbind(map01,map02)
 
+# getting the birth rates
 mapjoin = mapjoin %>%
@@
 
+# getting the average 2019 birth rates for each state  
 mapjoin2019 = mapjoin %>%
@@
 
+# getting the average 2000 birth rates for each state
 mapjoin2000 = mapjoin %>%
@@
 
+# geting the difference between birthrates between 2000 and 2019
 mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
@@
 ```{r figseven, fig.cap = "Decline in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
+# plotting the map with change in birthrates for each state
 plot_usmap(data = mapdiff, values = "decline", regions = "states") + 
@@
 ```{r tabtwo,echo = FALSE }
+
+# mean student debt by year
 studentdebt = student_debt %>%
@@
 
+# mean rent by year, actually recorded monthly
 renttable = rent %>%
@@
 
+# getting the percentage of people who believe religion is important
 religiontable = religion %>%
@@
 
+# mean child care cost by year
 childcare = childcare[complete.cases(childcare), ]
@@
 
+# joining the previous datasets into one and removing unnecessary variables
 explanatory_table = cbind(studentdebt,renttable,religiontable,childtable)
@@
 As the states which were excluded make up a decent portion of the U.S population and tend to have low costs of living than the national average, this could have led to a higher nationwide average child care cost in the report than is factual.
+For some datasets not everybody responded, so they either put 0 or estimated values.
+This may lead to inaccurate results, as their estimates are likely not the true value.
 
@@
 
-### Stances on Women in society
+### Position of Women in society
 
@@
 Families are smaller now, both due to the growth of single- parent households and the drop in fertility.
-The rise of LGBT families has also led to an increase in adoptions and decrease in birth rates.
+The rise of LGBT families for generation Z has also led to an increase in adoptions and decrease in birth rates.
 Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed.The increase in women achieving higher education [@citeusatrends] has also led to a pause to the child rearing stage for many, which has also caused a decrease in the birth rates for young women as this causes women to have children later in life after they are done with education.The causes of these decrease in births are tied to increases in educational attainment, growing labor force participation.
@@
 As the economic recession devastated many sectors of the economy and areas of the country, an increase in immigrants taking jobs in the U.S promotes feelings of animosity towards them.
-These feelings of animosity are also fuled by immigrants tendency to have more children than the citizen populations.
+These feelings of animosity are also fueled by immigrants tendency to have more children than the citizen populations.
 However these arguments often do not take into account the fact that second and third generation immigrants tend to reduce their birth rates to match the rest of the population [@immigbirth].
 This decrease in birth rates is thus used both  as a justification for immigration and job outsourcing, as well as an 'invasion crisis'.
+Further impacts on population growth imply that the U.S may be moving to Japan levels of decline, in which the burden of the economic system is placed on the decreasing numbers of young people, as the increasing numbers of old people rely on them.
 
@@
 Shifts in fertility have disproportionately happened to white women and the higher educated, who also tend to be white [@Yang2003].
-Hispanic women also have less than stellar access to fertility health care, as well as have lower IVF success rates.
+Hispanic women also have less than stellar access to fertility health care, as well as have lower IVF success rates [@citeracefertility].
 Black people are concentrated in the south, which has very high levels of teenage pregnancy [@citeteen] and low level of resources on sexual education, so they are disproportionately affected.
@@
 Black teenagers are also less likely to have access to quality education,careers and outside opportunities[@citeusatrends], which is a major factor to the decline in birth rates.
-Hispanics also are affected by many of the same problems; lack of funding and proper education in their communities.
-As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.
-This shows yet another way in which women of color are negatively treated in society.
+Hispanics are also affected by many of the same problems; lack of funding and proper education in their communities.
+As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.Black women are also at a higher risk of unwanted pregnancies compared to other populations, as shown by [@Bryant2010], which surveyed women who had recently gotten pregnant and asked questions such as if they intended on getting pregnant.
+Among the proportion of women who answered no to this question, black women represented 33% of the women not intending for pregnancy.
+Teenage birth rates have decreased for generation z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted even with the cultural change.
+This is due to lack of opportunities, as well as higher rates of poverty and low education [@Gold2001].
+While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back.This shows yet another way in which women of color are negatively treated in society.
 
@@
 
-### Unemployment
-While unemployment rates were a factor which affected birth rates between 2000-2013, there are other variables at play as well.
-The increase in birth rates from 2000 onwards is likely due to the economic and job growth that the United States had leading up to 2006.
-Between 2001 and 2007, the United States was going through an expansion period which contributed to its economic growth.
-This expansion was one that was weaker than the post WWII expansion and the one that had occured in 1990 with annual growth rates of 4.3% and 3.3% respectively however, there was still a 2.8% annual growth when looking at GDP.
+### Economic Expansion
+
+As seen in the results section, we noticed an increase in birth rates from 2002 onward.
+This is likely because of the economic and job growth that the United States had leading up to 2006.
+Between 2001 and 2007, the United States was going through an expansion period which contributed to its economic growth, with an annual GDP growth of 2.8% [@citeexpansion].
+Studies show that younger adults struggle to be financially stable, so the opportunity that an expansion brings for them would positively affect their financial status putting them in a better position to have a child [@citeyoungadult].
 This growth is not evident when looking at birthrates for the 15-19 age group as the declining birth rate continues until 2005.
 In contrast, the results of this expansion period are more visible in the 20-24 age range as it seems that the birth rate from 2001 to 2005 decreased at a lower rate each consecutive year until 2005, where it began to increase.
-This is likely because of the fact that the 15-19 age group isn’t affected by an economic expansion as much as the 20-24 age group as at such a young age, most of them would still be in school.
+This is likely because of the fact that the 15-19 age group is not affected by an economic expansion as much as the 20-24 age group.
 On the contrary, an economic expansion provides a lot more jobs and income for young adults who are done or almost done with school.
-(https://www.cbpp.org/research/how-robust-was-the-2001-2007-economic-expansion)
-
-In a time like this, families are more likely to be in a position where they can afford to have a child because interest rates and debts that they owe are lower than usual.
-With more people employed, there is generally more income for families to support the needs of an infant.
-(cite)
-
+In addition, in times of expansion families are more likely to be in a position where they can afford to have a child because interest rates and the real value of debts are lower than usual.
+With more people employed, there is generally more income for families to support the needs of children.
 However, that growth began to decline in 2007 where consumers began to pay more for debt that they had to repay.
@@
 This time also marked the end of the housing boom which is generally caused by economic prosperity and lower credits.
-Due to this, the decline of the economy also meant the end of the housing boom and also, the beginning of the declining birth rates.
-
-((https://www.americanprogress.org/article/the-u-s-economy-in-review-2006/))
-
-Due to a higher number of unemployed workers and the declining economy, affording a child became difficult which is why we notice the sudden decrease in average births across the United States.
+Due to this, the decline of the economy also meant the end of the housing boom and also, the beginning of the declining birth rates [@citeusaecon2006].
+In (Figure \@ref(fig:figsix)), we observe a large decrease in births for both age groups as seen by the steep negative slope,  likely caused by the great recession which began in late 2006 and early 2007.
+The points on the graph also show higher unemployment rates due to the recession.
+Due to a higher number of unemployed workers and the declining economy, affording a child became difficult which is why we notice the sudden decrease in average births across the United States[@citeeconrecession].
+Along with this, couples in the United States are concerned with the fluctuating economy of the United States since they would prefer that they start their family in a time where they felt safe(@citewomenhousing).
 This steep decline of births seems to slow down in 2010 which is also the time that the great recession ended.
-
-https://pubmed.ncbi.nlm.nih.gov/22066128/#:~:text=In%20most%20countries%2C%20the%20recession,decade%20of%20rising%20fertility%20rates.
-
-what caused the great decline in hispanic borth rate decrease?
-
-Elaborate on how decreased religion, increased childcare, mean rent, student debt
-
-Weaknesses next steps- scrape the data and form our own initial datasets?
-
-social explanations: teen birth rate?
-
-Limitations: we couldnt get the exact age groups fror some variables, so we had to take the ages of all women or the n20-34 age groups to approximate, so we may not get the accuarte results as the general trend may not apply to the sepcific trend of teenaged and young mothers.
+However it is important to note that women living in poverty are more likely to give birth young, compared to middle income women [@Kearney2012].
 
@@
 The change in birthrates for some states were very different than others.
-Some states were able to bounce back from the economic depression (cite), while some states have always had cheaper costs of living.
-From *FIGURE MAP* These states, primarily the midwest, are still heavily affected by the ecomic recession that started due to the fall of the steel industry (cite) and outside sourcing of labour.
-However cost of living and childcare are also relatively cheap in these states, so it didnt affect birth rates as much (cite).
-The west coast was able to bounce back better than the midwest, but as the cost of lviving is still expensive, especially in washington, oregon and california, this could deter women from having children quite early until they are fiancially ready.
-In the east coast and the south, the cost of living is also quite high but as these states tend to be more religious and conservative and value the family more (cite) than the liberal west coast, birth rates did not decline as far as the west coast.
-It is also known that the east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher.
-However as gen z is now more liberal than their millenial counterparts they are less likely to have more kids nationwide. 
-and problems with birth control and abortion in more conservative states.
- Pressuring of women in religious areas and less sex ed as accuatomed ton those areas could also be the cause of the fewer decrease in south and midwest.
+Some states were able to bounce back from the economic depression, while some states have always had cheaper costs of living.
+From *FIGURE MAP* These states, primarily the midwest, are still heavily affected by the economic recession that started due to the fall of the steel industry [@Meyer2003] and outside sourcing of labor.
+However cost of living and childcare are also relatively cheap in these states [@citerent], so it birth rates were not heavily affected.
+The west and east coast was able to bounce back better than the midwest, but as the cost of living is still expensive [@McMahon1991], this could deter women from having children quite early until they are financially ready.
+In the south, the cost of living is also quite high but as these states tend to be more religious and conservative than the liberal west coast, birth rates did not decline as far as the west coast.The east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher despite the expensive cost of living.
+Southern states also tend to have more problems with abortion and aim to restrict them, which could also be a reason behind the comparatively higher birth rates.
+However as generation Z tends to be more liberal than their millennial counterparts they are less likely to have more kids nationwide.
 
-Why do some states have a larger decrease in birth rates compared to others?
-
-The rent is a huge factor when it comes to why some states were hit harder than others.
-Rent on a two-bedroom apartment in Henderson, Nevada has risen 23 percent to nearly $1,600 a month.
-This makes it almost impossible for a single mother to raise her kids and hence it causes them to move to a different state where the rent is rather lower.
-North Dakoda is a prime example of why it is beneficial to live there and raise your family.
-The average rent in the state is around $850 which is roughly half of what it is in Nevada.
-This is why they choose to move out of that state causing that state to have a higher decrease.
-
-https://worldpopulationreview.com/state-rankings/average-rent-by-state
-
-https://www.washingtonpost.com/business/2022/01/30/rent-inflation-housing/
-
-# Teen birth rates, race and economic status
-
-Teenaged birth rates have decreased in the past 20 years for gen z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted ven with the cultural change.
-This is due to lack of opportunities and that is all they know, as well as higher rates of poverty and low education [@Gold2001].
-While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back.
-It is of note that although white teenagers are mothers at a larger number, and thus make up most of teen mothers, most of the concern is based on the poc women and these data can perpetuate these biases.
-More education is needed in these communities.
-Poor women irrepsective of race are also more likely to get pregnant earlier, this is mainly in the south as they still practice abstinence only sex ed.
-This report, which shows an increase in irreligiousnes could perpetuate the false notions that teaching respinsible sex leads to more recklenss, when it is the other way around (cite).
-Increasing more liberal, science based forms of sex-ed, while controversial, would decrease teen birth rates.
-However, there is push back because...
-(cite).
-WHile still being religious, these people have high levels of teen pregnancy, so policy is needed in schools to fight back agaisnt abstinence based sex-ed.
-It is knowsn to also shame women for having sexual thought,s cotnributing to a misogynistic society(cite)
-they have liitle change of advancement (doi: 10.1257/jep.26.2.141)
-
-# Conclusion:
-
-It is necessary... (from)
-# Weakness / Next Steps
+# Weakness and Next Steps
 
 There were datasets in which we had to make generalizations about a specific population based on data of a general population.
-For example, in Figure *US MAP FIGURE*, the map data was composed of responses taken from women aged from 20-34.
-This serves as a limitation because it ignores an entire age group from the study that is being conducted since this paper is focused on births for those aged between 15-24.
-Due to this, conclusions that are made from this map are not as accurate as they could be because the true statistics of the 15-19 age group aren’t provided.
-In a similar manner, Table 1 also carries a degree of error as due to the data that was provided, we are restricted to making conclusions about ages 15-24 based on data which does not specify age.
+For example, in Figure *US MAP FIGURE*, the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group were not available.
+Due to this, conclusions that are made from this map are likely not accurate as they are not the true statistics of the 15 to 24 age group.
 Our research question wanted to focus on the trend of birth rates over a 20 year period, specifically from 2000 to 2020.
@@
 Due to this, instead of being able to compare 2000 and 2020 explicitly, we had to generalize trends that we see in 2019 to be at least similar to trends that we’d see in 2020.
-We had the same problem with Table 2, that we had to approximate 2004 to 2000 even though situations would be different.
-
-https://ifstudies.org/blog/higher-rent-fewer-babies-housing-costs-and-fertility-decline
-
-The years for millenial and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
+We had the same problem with *Table 2*, in which we had to approximate 2004 to 2000 even though situations would be different.
+In addition to this, the years for millennial and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
 Thus the years chosen for this report are completely arbitrary.
@@
 We are interested in if this has had any effect on birth rates, and in future research we would like to explore the relationship between them.
-
-for next steps: Could falling testosterone rates(cite) also attribute to these? we would like to make a table that shows the change in the 20 years.
+We would also like to download the data directly from the sources to clean and form our own initial datasets.
 

### 2022-02-27 Scripts/Rscript.R
deleted files

@@ -14,7 +14,7 @@
 library(haven)
 library(tidyverse)
 
-# Reading in the data in the raw data.
+# Reading in the data.
 birth = read_csv("Inputs/Data/fig_1.csv")
 numbirth = read_csv("Inputs/Data/numbirths_2001_2019.csv")
 student_debt = read_dta("Inputs/Data/nyfed-student-debt-per-capita-0419.dta")

### 2022-02-28 Output/paper 2.Rmd
changes to code to make sure it knits well

@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
-abstract: "birth rates are merely a symptom of cultursl changes betweeen two generations and the economic situation.
-It is important to see that these are caused by everything more expensive now.
-make sure to cite R in the abstract."
+abstract: "Birth rates provide useful information on community health, and an above-replacement birth rate is indicative of stability.
+However, previous research has found that birth rates in industrialized societies such as the U.S have been decreasing in recent years.
+We use data from the American Economic Association to analyze this by using graphs and tables to observe the trend in birth rates of different demographics, using R [@citeR] and other packages.
+We also observe the trends behind changing economic and cultural factors that could affect the birth rates.
+We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have been on an increase.
+This report shows that the decline in birth rates are a symptom of cultural and economic changes on women between two generations.
+While this indicates that the U.S has gotten more culturally liberated and industrialized, the declining birth rates shows the systemic effects of the worsening economy and increasing costs on the average american woman."
 output:
@@
 We obtained the dataset from the American Economic Association [@citewebsite].
-We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
-Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable].
+We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr],`haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
+Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegridExtra] and `kableExtra` [@citekable].
 We first created histograms and scatter plots which showed the most important variables.
@@
 
-We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information.
-We included the datasets that had information on race, as this was needed to compare birth rates for different races.
-We also had datasets on total number of births and/or birth rates, both of which were used to analyze the decline in births.
-We had total population datasets, used to get the birth rates and our demographic's share of the population.
-We had datasets on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.
-We also included datasets that included information on government policy, rent costs, student debt, religious beliefs, and childcare costs.
-These datasets were used to signify a possible relationship between these variables and birth rates.
+We included datasets and variable if they were considered important to answering the question behind birthrates, or revealed important information.
+We included the variables that had information on race, as this was needed to compare birth rates for different races.
+We also had variables on total number of births and/or birth rates, both of which were used to analyze the decline in births.
+We had a total population variable, used to get the birth rates and our demographic's share of the population.
+We had variables on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.
+We also included variables on government policy, rent costs, student debt, religious beliefs, and childcare costs.
+These variables were used to signify a possible relationship between them and birth rates.
 
@@
 We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rates was recorded also had their educational status recorded.
-We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within the column of `group`.
+We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within one column.
 We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
@@
 we obtained a variable got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100.
-We had a variable that got contribution to total birth rate decline by subtracting 2019 values from the 2000 values and dividing that by the total birthrate value for 2000 and multiplying by 200.
+We produced a variable on contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birthrates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200.
 We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states.
@@
 ```{r global_options, include=FALSE}
-# laoding in relevant libraries
+# loading in relevant libraries
 knitr::opts_chunk$set(fig.pos = 'H')
 library("kableExtra")
-# install.packages("usmap")
+#install.packages("usmap")
 library("tidyverse")
 library("haven")
-library("janitor")
 library("dplyr")
@@
 #loading in relevant datasets
+
 birth = read_csv(here::here("Inputs/Data/birth.csv"))
@@
 ```{r, include=FALSE}
+
 # getting total number of births for white people from 15 to 19 years old
@@
 
-```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020", message = FALSE}
+```{r figone, out.width="90%", echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020", message = FALSE}
 
@@
                ) %>%
-  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
+  ggplot(data =., aes(x=value, fill = Race)) +
   geom_histogram()+ 
@@
   xlab("Number of births") +
-  ylab("Number of Years")
+  ylab("Number of Years") +
+  scale_fill_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", "White, non-Hispanic"))
+```
 
-```
-```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020", message = FALSE}
+```{r figtwo, out.width="90%", echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020", message = FALSE}
 
@@
                ) %>%
-  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
+  ggplot(data =., aes(x=value,fill=Race)) +
   geom_histogram()+ 
@@
   xlab("Number of Births") +
-  ylab("Number of Years")
+  ylab("Number of Years")+
+  scale_fill_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", "White, non-Hispanic"))
 
@@
 
-```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
+```{r figthree, out.width="90%", echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
 
@@
 
-```{r figfour, echo = FALSE, fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020"}
+```{r figfour, out.width="90%", fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020", echo = FALSE}
 
@@
 We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation Z they have converged more closely due to the decrease in birth rates for all races.
-This could be due to economic factors directly related to race, such as inequality in the healthcare system [@Davis2018] .
+This could be due to economic factors directly related to race, such as inequality in the health care system [@Davis2018] .
 As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
 
-```{r figfive, echo = FALSE, fig.cap = "Trend in Birth rates by Race from 2000 to 2020"}
+```{r figfive, out.width="90%", fig.cap = "Trend in Birth rates by Race from 2000 to 2020", echo = FALSE}
 
@@
 ```
-(Table \@ref(fig:tabone)) shows the contribution of our main groups to the total decline in birth rates from 2000 to 2019.
+
+(Table \@ref(tab:tabone)) shows the contribution of our main groups to the total decline in birth rates from 2000 to 2019.
 It also shows each group's share of the total female population for those years.
@@
 White 15 to 19 year old women have caused a 7% drop in the total birth rate, indicating a rapid decline in the teenage pregnancy for this race [@Wingo2011].
-White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2017].
+White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2007].
 Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population.
-Black women are disproportionately poorer, and low economic status leads to higher teenage pregnancy rates [@Gold2001].
+Black women are disproportionately poorer [@citepovertyrace], and low economic status leads to higher teenage pregnancy rates [@Gold2001].
 The contribution of Hispanic women of ages 20 to 24 is less than their teenage counterparts, showing that there is not a large decrease in their birth rates as opposed to white women.
@@
 
-```{r figsix, echo = FALSE, fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020" }
+```{r figsix, out.width="90%", fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020", echo = FALSE}
 
@@
 ```
+
 (Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically the the 20-34 age group, by state from 2000 to 2019.
@@
 
-# geting the difference between birthrates between 2000 and 2019
+# getting the difference between birthrates between 2000 and 2019
 mapdiff = left_join(mapjoin2019, mapjoin2000, by="stname") %>%
@@
 
-```{r figseven, fig.cap = "Decline in Birth rates for young women from 1999-2000 to 2018-2019 by state"}
+```{r figseven, out.width="90%", fig.cap = "Decline in Birth rates for young women from 1999-2000 to 2018-2019 by state", echo = FALSE}
 # plotting the map with change in birthrates for each state
@@
 
-(Table \@ref(fig:tabtwo)) compares average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019.
+(Table \@ref(tab:tabtwo)) compares average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019.
 The mean student debt is lifetime debt while the mean rent is monthly rent.
@@
 
-```{r tabtwo,echo = FALSE }
+```{r tabtwo, echo = FALSE }
 
 # mean student debt by year
-studentdebt = student_debt %>%
+studentdebttable = student_debt %>%
   group_by(year) %>%
@@
 # joining the previous datasets into one and removing unnecessary variables
-explanatory_table = cbind(studentdebt,renttable,religiontable,childtable)
+explanatory_table = cbind(studentdebttable,renttable,religiontable,childtable)
 explanatory = explanatory_table %>%
@@
 explanatory %>%
-  knitr::kable(col.names = c("Year","Student Debt", "Rent", "Religious Importance","Child Care Cost"), caption = "Changes in Variables that may Explain Decline in Birth Rates")%>% 
+  knitr::kable(col.names = c("Year","Average Student Debt", "Average Rent", "Percentage of Religious People","Average Child Care Cost"), caption = "Changes in Variables that may Explain Decline in Birth Rates")%>%
   kable_styling(latex_options = "HOLD_position") 
@@
 Hispanics are also affected by many of the same problems; lack of funding and proper education in their communities.
-As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.Black women are also at a higher risk of unwanted pregnancies compared to other populations, as shown by [@Bryant2010], which surveyed women who had recently gotten pregnant and asked questions such as if they intended on getting pregnant.
+As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.
+Black women are also at a higher risk of unwanted pregnancies compared to other populations, as shown by [@Bryant2010], which surveyed women who had recently gotten pregnant and asked questions such as if they intended on getting pregnant.
 Among the proportion of women who answered no to this question, black women represented 33% of the women not intending for pregnancy.

### 2022-02-28 Output/paper 2.Rmd
final changes

@@
 date: "`r format(Sys.time(), '%d %B %Y')`"
-abstract: "Birth rates provide useful information on community health, and an above-replacement birth rate is indicative of stability.
+abstract: "Birth rates provide useful information on population growth, and an above-replacement birth rate is indicative of stability.
 However, previous research has found that birth rates in industrialized societies such as the U.S have been decreasing in recent years.
@@
 We also observe the trends behind changing economic and cultural factors that could affect the birth rates.
-We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have been on an increase.
-This report shows that the decline in birth rates are a symptom of cultural and economic changes on women between two generations.
-While this indicates that the U.S has gotten more culturally liberated and industrialized, the declining birth rates shows the systemic effects of the worsening economy and increasing costs on the average american woman."
+We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have increased.
+This report shows that the decline in birth rates are a symptom of cultural and economic changes between two generations of women.
+While this indicates that the U.S has gotten more culturally liberated, the declining birth rates shows the systemic effects of the worsening economy and increasing costs on the average american woman."
 output:
@@
 The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry].
-This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society.
+This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford children without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society.
 The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies.
@@
 
-For data cleaning, our target sample and main focus was young mothers from two generations: millennial mothers, who were between 15 to 24 years in 2000, and generation Z mothers, who were between 15 to 24 years old in 2020.
-We observed birthrates from 2000 to 2020, in order to see the effect of relatively short term effects on birthrates between the generations.
-Millennial are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock].
-We also considered birth rates for white, black, and hispanic women that were in our target age ranges as we wanted to compare the birth rates between them.
+For data cleaning, our target sample and focus was young mothers from two generations: millennials, who were between 15 to 24 years in 2000, and generation Z, who were between 15 to 24 years old in 2020.
+We observed birthrates from 2000 to 2020, in order to see the effect of relatively short-term effects on birthrates between the generations.
+Millennials are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock].
+We also considered birth rates for White, Black, and Hispanic women that were in our target age ranges as we wanted to compare the birth rates between them.
 We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e.
@@
 
-We included datasets and variable if they were considered important to answering the question behind birthrates, or revealed important information.
-We included the variables that had information on race, as this was needed to compare birth rates for different races.
-We also had variables on total number of births and/or birth rates, both of which were used to analyze the decline in births.
-We had a total population variable, used to get the birth rates and our demographic's share of the population.
+We included datasets and variables if they were considered important for answering the question behind birthrates or revealed important information.
+We included variables on race, as this was needed to compare birth rates for different races.
+We also had variables on total number of births and birth rates, both of which were used to analyze the decline in births.
+We had a total population variable, used to get the birth rates and our demographics' share of the population.
 We had variables on birth rates for different age groups, states, and years, all of which were used to analyze the trends in birth rates for different groups.
@@
 These variables were used to signify a possible relationship between them and birth rates.
+All the variables listed were numeric, with the exception of race and age group.
 
@@
 We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within one column.
-We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
+We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.
+These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
 We obtained a variable which got the decline in birth rates from 2000 to 2019 by state, by creating dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.
 
-We got a variable for the total number of births by adding up all the births by state and then grouping them by year.
-We got a variable that stated the total population of each race-age combination by similar methods.
-we obtained a variable got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100.
+We created a variable for the total number of births by adding up all the births by state and then grouping them by year.
+We also created a variable that stated the total population of each race-age combination by similar methods.
+We obtained a variable that got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100.
 We produced a variable on contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birthrates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200.
@@
 For the average debt, we multiplied by 10,000 as that is how it was shown in the original source.
-We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.Note that for these variables we needed to use 2004 as 2000 was not available.
+We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.
+Note that for these variables we needed to use 2004 as 2000 was not available.
 
@@
 
-## View of the Data
+## Data Visualization
 
@@
                ) %>%
-  select(year,racepercentbirth,birthratevalue)
+  mutate(demographic = c("White, 15 to 19", "White, 20 to 24", "Black, 15 to 19", "Black, 20 to 24", "Hispanic, 15 to 19", "Hispanic, 20 to 24","White, 15 to 19","White, 20 to 24","Black, 15 to 19","Black, 20 to 24","Hispanic, 15 to 19", "Hispanic, 20 to 24")) %>%
+  select(year,demographic,birthratevalue)
 
@@
 We notice that these costs have all risen past inflation, indicating that it is now more expensive to live in the U.S in 2019 than it was in the early 2000s.
-As generation Z mothers want to be more financially comfortable before having children, this causes them to postpone having kids, or not having as many children due to expense.
+As generation Z mothers want to be more financially comfortable before having children, this causes them to postpone having children, or not having as many children due to expense.
 Expensive childcare and lack of affordable living has caused lower fertility rates [@citenargund].
@@
 This would then make analysis confusing as different gender identities may have different experiences [@Tabaac2017].
-The datasets also did not include transgender men, who can also get pregnant, bringing the worrisome question of if pregnant transgender men were listed as 'male' in the U.S census, why the datasets were filtered by female only.
+The datasets also did not include transgender men, who can also get pregnant, bringing the question of if pregnant transgender men were listed as 'male' in the U.S census, and why the original datasets were filtered by female only.
 
@@
 The west and east coast was able to bounce back better than the midwest, but as the cost of living is still expensive [@McMahon1991], this could deter women from having children quite early until they are financially ready.
-In the south, the cost of living is also quite high but as these states tend to be more religious and conservative than the liberal west coast, birth rates did not decline as far as the west coast.The east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher despite the expensive cost of living.
+In the south, the cost of living is also quite high but as these states tend to be more religious and conservative than the liberal west coast, birth rates did not decline as far as the west coast.
+The religious level of the south also corresponds to higher teen birth rates [@Strayhorn2009].
+The east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher despite the expensive cost of living.
 Southern states also tend to have more problems with abortion and aim to restrict them, which could also be a reason behind the comparatively higher birth rates.
-However as generation Z tends to be more liberal than their millennial counterparts they are less likely to have more kids nationwide.
+However as generation Z tends to be more liberal than their millennial counterparts they are less likely to have more children nationwide.
 
@@
 There were datasets in which we had to make generalizations about a specific population based on data of a general population.
-For example, in Figure *US MAP FIGURE*, the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group were not available.
+For example, in (Figure \@ref(fig:figseven)), the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group were not available.
 Due to this, conclusions that are made from this map are likely not accurate as they are not the true statistics of the 15 to 24 age group.
@@
 Due to this, instead of being able to compare 2000 and 2020 explicitly, we had to generalize trends that we see in 2019 to be at least similar to trends that we’d see in 2020.
-We had the same problem with *Table 2*, in which we had to approximate 2004 to 2000 even though situations would be different.
+We had the same problem with (Table \@ref(tab:tabtwo)), in which we had to approximate 2004 to 2000 even though situations would be different.
 In addition to this, the years for millennial and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
@@
 
+\newpage
+
 # References

### 2022-02-28 Output/paper 2.Rmd
Last changes to code and writing

@@
 ---
-title: "Decline in U.S Birthrates is Indicative of Socio-economic Changes"
+title: "The Decline in U.S Birthrates in Recent Years is Indicative of Socio-economic Changes"
 author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
@@
 We find that birth rates of young women of different races have been on a steep decline over the past 20 years, and factors that are known to decrease birth rates have increased.
-This report shows that the decline in birth rates are a symptom of cultural and economic changes between two generations of women.
-While this indicates that the U.S has gotten more culturally liberated, the declining birth rates shows the systemic effects of the worsening economy and increasing costs on the average american woman."
+This report shows that the decline in birth rates is the result of cultural and economic changes between two generations of women.
+While this indicates that the U.S has become more culturally liberated, the declining birth rates showcases systemic effects of the worsening economy and increasing costs on the average American woman."
 output:
@@
 
-A nation's birthrate for a particular year is the total number of live births per 1000 people for that year.
+A nation's birth rate for a particular year is the total number of live births per 1000 people for that year.
 In order to measure the birth rates and fertility levels to forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors.
@@
 It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics.
-The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER).
+The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiology, and End results program (CDC SEER).
 This information also breaks down the population by age, race, and socio-economic factors.
 In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographics of the mother.
-We are also interested in the relationship between the change in economic, social and cultural aspects of the United states and the change in these birth rates.
+We are also interested in the relationship between the change in economic, social and cultural aspects of the United States and the change in these birth rates.
 
 We obtained the dataset from the American Economic Association [@citewebsite].
-We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr],`haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
+We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr` [@citedplyr], `tidyr` [@citetidyr], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages.
 Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegridExtra] and `kableExtra` [@citekable].
@@
 Young women are in no rush to have children, as this report shows that birth rates have declined for the 15-24 age groups over the past 20 years.
-The report also shows that birth rates have declined for all races analyzed over the same period of time.
+The report also shows that birth rates have declined for all races analyzed over the same period.
 This implies that this decline is a general national issue attributed to national factors, as opposed to factors that mainly affect a specific demographic.
@@
 Birth rates for our age demographics have experienced large decreases in some states, while those in other states have had a relatively smaller decrease or no decrease at all.
-This may be due to the fact that some economic or social changes of the country affect some geographic areas more than others.
-These economic and social factors suspected to have an influence on birth rates also changed over the past 20 years in ways that would correlate with declining birth rates.
-According to [@citeadsera], these rates of birth decline are similar to declines in other high income, developed countries.
+This may be because some economic or social changes of the country affect some geographic areas more than others.
+These economic and social factors that are suspected to have an influence on birth rates have also changed over the past 20 years in ways that would correlate with declining birth rates.
+According to the study in [@citeadsera], this decline in birth rates are similar to declines in other high income, developed countries.
 
 While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population.
-The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry].
-This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford children without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society.
+The declining birth rates also brings light to the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry].
+This is problematic as although positive cultural changes have led to more freedom for women than before and decreased teenage births,now  many women are no longer able to afford children without a strong financial background.
+The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their discrimination in society.
 The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being irresponsible with their bodies.
@@
 
-The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite].
-The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer],the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) , the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums].
+The datasets for this report were obtained from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite].
+The data used by the paper was compiled by various sources such as the NCHS Natality Database [@citenchs], CDC SEER [@citecdcseer], the Current Population Survey Annual Social and Economic Supplement (CPS ASEC) [@citeipums], the New York Federal Reserve [@citefed], and the Integrated Public Use Microdata Series (IPUMS) [@citeipums].
 The paper and datasets were last accessed on February 16th, 2022.
 The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019.
-The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, and this was randomly sampled [@citenchs].
-The national population data comes from the census, a long running survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer].
+The birth rate data was obtained from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which were randomly sampled [@citenchs].
+The national population data comes from the census, a long running non-voluntary survey funded by the government released on a predetermined cycle, and counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer].
 The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics.
-These datasets were gotten from surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or estimated.The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity.
+These datasets were gotten from voluntary surveys so not everyone responded, thus the data was often recorded as 0, `NA`, or was estimated.
+The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity.
 Meanwhile, categorical data mainly appeared in the form of state.
@@
 For data cleaning, our target sample and focus was young mothers from two generations: millennials, who were between 15 to 24 years in 2000, and generation Z, who were between 15 to 24 years old in 2020.
-We observed birthrates from 2000 to 2020, in order to see the effect of relatively short-term effects on birthrates between the generations.
+We observed birth rates from 2000 to 2020, in order to see the effect of relatively short-term effects on birth rates between the generations.
 Millennials are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock].
-We also considered birth rates for White, Black, and Hispanic women that were in our target age ranges as we wanted to compare the birth rates between them.
-We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e.
-2019 data is used if 2020 data is not available.
+We also considered birth rates for White, Black, and Hispanic women that were in our age ranges on interest, as we wanted to compare the birth rates between them.
+We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e., 2019 data is used if 2020 data is not available.
 We dropped any `NA` values from the dataset if they appeared.
@@
 
-We included datasets and variables if they were considered important for answering the question behind birthrates or revealed important information.
+We included datasets and variables if they were considered important for answering the research question,  or if they revealed important information.
 We included variables on race, as this was needed to compare birth rates for different races.
@@
 We created several of our own datasets by modifying and joining variables.
-One dataset was gotten by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year.
+One dataset was obtained by adding all the educational statuses for each age-race combination from another dataset to get the total birth rate for each combination, and then compiling the values by state and year.
 We also got a variable for total population by adding the total population for each state in a year.
 We obtained the final dataset used to create the map of birth rate decline by combining various other datasets by year, and subtracting the birth rates of the later years by that of the earlier years.
-The final dataset that showed how our demographics of interest contributed to total birth rate decline was created by joining several datasets all which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions.
-The final dataset used to display the change in explanatory variables over time was gotten by combining the datasets, each of which explained the changes in a single variable.
+The final dataset that showed how our demographics of interest contributed to total birth rate decline, was created by joining several datasets which compared important information, then dividing the columns of our demographic by the column of total birth rates to get the relative contributions.
+The final dataset used to display the change in explanatory variables over time was obtained by combining the datasets, each of which explained the changes in a single variable.
 Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).
@@
 We modified many variables for the report.
-We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rates was recorded also had their educational status recorded.
+We created variables for total birth rates of each race-age combination by adding up all educational statuses for each race and age group combination and dividing by the total population, as each person whose birth rate was recorded also had their educational status recorded.
 We also modified age group variables by changing the columns of 15-19 and 20-24 age to rows within one column.
 We did the same for race variables, modifying the columns to be rows within one group, and then multiplying the values column by 10,000 and dividing by the total female population to get the birth rates for each race.
-These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.
-We obtained a variable which got the decline in birth rates from 2000 to 2019 by state, by creating dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.
+These variables were used to get the trend in birth rates for our age demographics, and trend in birth rates in race for our age demographics.
+We obtained a variable which displayed the decline in birth rates from 2000 to 2019 by state, by dividing the number of births by population to get birth rates, grouping them by year and state, and then subtracting the values of the later year from the values of the earlier year.
 
@@
 We obtained a variable that got the percentage of population of a demographic by dividing the sum of that demographic by the total population and multiplying by 100.
-We produced a variable on contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birthrates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200.
+We produced a variable of contribution to total birth rate decline by subtracting 2019 birth rates of the age group from the 2000 birth rates of the same age group and dividing that by the total birthrate value for 2000, then multiplying by 200.
 We got the variables for all our explanatory factors by grouping them by year and getting the mean values from all the states.
@@
 We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.
-Note that for these variables we needed to use 2004 as 2000 was not available.
+Note that for these variables we needed to use 2004 data as 2000 data was not available.
 
@@
 This also means that the average number of births will be below the median number of births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012].
-We also see that black teenage mothers are less common, as they have the lowest births for most years.
+We also see that Black teenage mothers are less common, as they have the lowest births for most years.
 This is most likely because they are a much smaller segment of the population than the others.
-However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportionately appearing as teenage mothers.
+However, we see that black mothers still have a moderate number of births, indicating that they are disproportionately appearing as teenage mothers.
 Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year.
@@
 We see that there are far more women giving birth at this age group than in the previous age group.
-We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts.
-Hispanic mothers also tend to be at the lower end of the scale, but are more spread out.
-White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.
+We see that Black mothers are overwhelmingly on the lower end of births, which brings into question why teenage Black mothers had more prevalent births compared to their elder counterparts.
+Hispanic mothers also tend to be at the lower end of the scale but are more spread out.
+White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.
+We see that births for Black and Hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while White mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in Black and hispanic women may be more common.
 
-In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value.
+In contrast, in (Figure \@ref(fig:figthree)), the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value.
 This is because the population of a nation either increases or decreases, rarely staying the same.
-Due to this we can say that the mean and median population are approximately the same.
-However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year.
-From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase.
-This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates.
-However, the two dips in the scatterplot correspond to the previous histogram of which had years with similar values.
+Due to this, we can say that the mean and median population are approximately the same.
+However there are some years which had around the same population, this could be due to declining population as a later year has the same population as an earlier year.
+From the scatter plot we see that the total female population has been increasing in recent years, with a seemingly exponential increase.
+This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birth rates.
+However, the two dips in the scatter plot correspond to the previous histogram of which had years with similar values.
 
@@
 This indicates a decrease in teenage pregnancy nationwide.
-The birth rates for the 20-24 age dropped by around 40% over the same time period.
+The birth rates for the 20-24 age dropped by around 40% over the same period.
 This may be because generation Z women are deciding to have children later in life.
 There was a baby boom in 2006 [@citecatholic], which is the reason for the slight increase in birth rates around that time.
-However overall birth rates for young women has dropped.
-According to [@Caldwell2006] there is little reason to believe these birth rates will increase any time soon.
+However overall birth rates for young women have dropped.
+According to [@Caldwell2006], there is little reason to believe these birth rates will increase any time soon.
 There are some possible explanations behind this phenomenon, including better sex education for teenagers leading to a decrease in teen pregnancy, more women focusing on their careers and thus postponing births [@citeexecutive], or the increased acceptance of abortions.
@@
 
-```{r figfour, out.width="90%", fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020", echo = FALSE}
+```{r figfour, out.width="90%", fig.cap = "Trend in birth rates for the 15-19 and 20-24 age groups from 2000 to 2020", echo = FALSE}
 
@@
 We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population.
-This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased.
-The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000.
-The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years.
-We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation Z they have converged more closely due to the decrease in birth rates for all races.
-This could be due to economic factors directly related to race, such as inequality in the health care system [@Davis2018] .
-As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
+This can either be because the population of Hispanic women has increased due to immigration, or that the number of births has decreased.
+The birth rates for Black women have also halved, going from 50 births per 1000 to just 25 births per 1000.
+The White birth rates however have decreased the most, going from about 125 births to about 65 births through the 20 years.
+We see that millennial mothers (mothers in 2000) had a large difference in the birth rates for different races.
+However, for generation Z they have converged more closely due to the decrease in birth rates for all races.
+This could be due to economic factors directly related to race, such as inequality in the health care system [@Davis2018].
+As the income for Hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
 
-```{r figfive, out.width="90%", fig.cap = "Trend in Birth rates by Race from 2000 to 2020", echo = FALSE}
+```{r figfive, out.width="90%", fig.cap = "Trend in birth rates by race from 2000 to 2020", echo = FALSE}
 
@@
 
-(Table \@ref(tab:tabone)) shows the contribution of our main groups to the total decline in birth rates from 2000 to 2019.
-It also shows each group's share of the total female population for those years.
+(Table \@ref(tab:tabone)) shows the contribution of the demographics of interest to the total decline in birth rates from 2000 to 2019.
+It also shows each demographic's share of the total female population for those years.
 We see that white women in both age groups have contributed the most to the decline, with a total of 18.2% of the total decrease.
@@
 White 20 to 24 year old women have contributed the most to the decline, possibly due to economic and cultural factors over the past 20 years [@Maher2007].
-Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy for this rates have not decreased significantly or because they are a much lower percentage of the population.
+Black women for both age groups have not shown a significant contribution to the decline in birth rates over this period however, either because teenage pregnancy rates for this race have not decreased significantly or because they are a much lower percentage of the population.
 Black women are disproportionately poorer [@citepovertyrace], and low economic status leads to higher teenage pregnancy rates [@Gold2001].
 The contribution of Hispanic women of ages 20 to 24 is less than their teenage counterparts, showing that there is not a large decrease in their birth rates as opposed to white women.
-They are also more likely to suffer from disproportional poverty.
-We see that white womens' share of the population from 2000 to 2019 has decreased due to the large decrease in birth rates.
-However black and hispanic womens' share of the population over this time period has either increased or not changed, likely due to increasing immigration from these two races over the past 20 years.
+We see that White womens' share of the population from 2000 to 2019 has decreased due to the large decrease in birth rates.
+However Black and Hispanic womens' share of the population over this period has either increased or not changed, likely due to increasing immigration from these two races over the past 20 years.
 
@@
   pivot_wider(names_from = year, values_from = birthratevalue)
-contribution$contribution = ((contribution$"2000" - contribution$"2019")/65.9)*200
+contribution$contribution = round(((contribution$"2000" - contribution$"2019")/65.9)*200,1)
 
@@
   pivot_wider(names_from = year, values_from = value)
-percentpop$"2000" = percentpop$"2000"*100
-percentpop$"2019"= percentpop$"2019"*100
+percentpop$"2000" = round(percentpop$"2000"*100,1)
+percentpop$"2019"= round(percentpop$"2019"*100,1)
 
@@
 
-(Figure \@ref(fig:figsix)) shows the number of births for those aged 15-24 for every from 2000 to 2019, according to the year's unemployment rate.
+(Figure \@ref(fig:figsix)) shows the number of births for those aged 15-24 from 2000 to 2019, according to the year's unemployment rate.
 We notice that the graph generally has a decreasing pattern in terms of the average number of births per year in the United States which is in line with what we observed earlier.
@@
 According to [@citecatholic], in 2006, the United States had their highest number of births than in the last 45 years prior, causing a baby boom.
-However after this, the birth rate began to quickly decline and unemployment rates were high, reaching 8% by 2008 and continuing until 2012.
+However, after this, the birth rate began to quickly decline and unemployment rates were high, reaching 8% by 2008 and continuing until 2012.
 This is likely due to the great recession that occurred during this period, such that it is possible women could not afford bringing new children.
@@
 The initial conclusion made by this is that low unemployment rates are related to a higher number of births and higher unemployment rates are related to a lower number of births.
-However beyond 2014, despite the low unemployment rate, we see the birth rate continues to decrease.
+However, beyond 2014, despite the low unemployment rate, we see the birth rate continues to decrease.
 We note that the unemployment rate is more likely to affect women in the 20-24 age group compared to their younger counterparts, as they are more involved in the workforce.
 
-```{r figsix, out.width="90%", fig.cap = "Average Birth Rates Vs level of Unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020", echo = FALSE}
+```{r figsix, out.width="90%", fig.cap = "Average birth Rates against level of unemployment for the 15 to 19 and 20 to 24 age groups from 2000 to 2020", echo = FALSE}
 
@@
 
-# creating scatterplot for the 15 to 19 age group
+# creating scatter plot for the 15 to 19 age group
 g1519 <- graph %>% ggplot(mapping = aes(x = year, y = brate_1519, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x = "Year", y = "Birth Rate for Ages 15-19") +
@@
 
-# creating scatterplot for the 20 to 24 age group
+# creating scatter plot for the 20 to 24 age group
 g2024 <- graph %>% ggplot(mapping = aes(x = year, y = brate_2024, color = Unemployment)) + geom_point() + theme_minimal() + scale_x_continuous(breaks = seq(1980,2020,by = 2))+ labs(x= "Year", y = "Birth Rate for Ages 20-24") +
@@
 
-(Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically the the 20-34 age group, by state from 2000 to 2019.
-We see that throughout the U.S there has been a decline in birth rates in most states, with some states having a more severe decline than others.
+(Figure \@ref(fig:figseven)) shows the percentage decline in birth rates for young women, specifically for the 20-34 age group, by state from 2000 to 2019.
+We see that throughout the U.S, there has been a decline in birth rates in most states, with some states having a more severe decline than others.
 We see that states on the west coast tend to have a larger decline than states on the east coast, with California having a very high decline in birth rates, and North Dakota having relatively little change.
-This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering  gender roles, such that women have children later or not at all.
-In comparison the east coast, midwestern and southern states have less severe changes in their birth rates, with some having no change or a slight increase in birth rates.
+This could be attributed to the west coast being more liberal [@citeleft] and thus less adhering to gender roles, such that women have children later or not at all.
+In comparison, the east coast, midwestern and southern states have less severe changes in their birth rates, with some having no change or a slight increase in birth rates.
 Coincidentally, these states tend to be more conservative or religious, so they accept gender norms and refusal of new changes in culture.
@@
 
-```{r figseven, out.width="90%", fig.cap = "Decline in Birth rates for young women from 1999-2000 to 2018-2019 by state", echo = FALSE}
+```{r figseven, out.width="90%", fig.cap = "Decline in birth rates for young women from 1999-2000 to 2018-2019 by state", echo = FALSE}
 # plotting the map with change in birthrates for each state
@@
 
-(Table \@ref(tab:tabtwo)) compares average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019.
+(Table \@ref(tab:tabtwo)) compares the average student debt, rent, child costs and the percentage of people that follow a religion in the years 2004 and 2019.
 The mean student debt is lifetime debt while the mean rent is monthly rent.
@@
 We notice that these costs have all risen past inflation, indicating that it is now more expensive to live in the U.S in 2019 than it was in the early 2000s.
-As generation Z mothers want to be more financially comfortable before having children, this causes them to postpone having children, or not having as many children due to expense.
+As generation Z mothers want to be more financially comfortable before having children, this causes them to postpone having children, or not having as many children due to the expenses.
 Expensive childcare and lack of affordable living has caused lower fertility rates [@citenargund].
@@
   group_by(year) %>%
-  summarize(meanrent = mean(avg_grossrent))
+  summarize(meanrent = mean(avg_grossrent)) %>%
+  mutate(meanrent = round(meanrent,1))
 
@@
   summarize(meanpercentimportant = mean(important)) %>%
-  select(year, meanpercentimportant)
+  select(year, meanpercentimportant) %>%
+  mutate(meanpercentimportant =round(meanpercentimportant,1))
 
@@
   group_by(year) %>%
-  summarize(meanchildcare = mean(coll_c3))
+  summarize(meanchildcare = mean(coll_c3)) %>%
+  mutate(meanchildcare = round(meanchildcare,1))
 
@@
 explanatory %>%
-  knitr::kable(col.names = c("Year","Average Student Debt", "Average Rent", "Percentage of Religious People","Average Child Care Cost"), caption = "Changes in Variables that may Explain Decline in Birth Rates")%>%
+  knitr::kable(col.names = c("Year","Average Student Debt", "Average Rent", "Percentage of Religious People","Average Child Care Cost"), caption = "Changes in Variables that may Explain Decline in Birth Rates") %>%
   kable_styling(latex_options = "HOLD_position")
+
 ```
@@
 As this group of people and their experiences are not included in the original study, it hinders both the accuracy and the main points that the author intended to make.
-Non binary people are systematically removed from society in many official and unofficial processes [@Tabaac2017], in such a way that the official U.S census even in 2020 did not include factor variables for non binary or other genders.
-Even if the data did include an 'other' option, that is grouping a lot of different gender identities into one large group; genders that are  completely different from each other.
+Nonbinary people are systematically removed from society in many official and unofficial processes [@Tabaac2017], in such a way that the official U.S census even in 2020 did not include factor variables for nonbinary or other genders.
+Even if the data did include an 'other' option, that groups a lot of different gender identities together; genders that are completely different from each other.
 This would then make analysis confusing as different gender identities may have different experiences [@Tabaac2017].
@@
 
-Adding to this, non hispanic black and white people got their own group but hispanic black and white people did not, as they were all grouped into one race.
-This is inaccurate as hispanics can be any race and although they are all within the hispanic ethnicity, the experiences and thus data for black hispanics will likely be a lot different than that of white hispanics due to racism within the hispanic community [@Haywood2017].
-This problem may also arise because most of society sees all hispanics as mixed indigenous and european, thus erasing black and indigenous hispanic experiences.
+Adding to this, non-Hispanic black and white people got their own group but black and white Hispanics did not, as they were all grouped into one race.
+This is inaccurate as Hispanics can be any race and although they are all within the hispanic ethnicity, the experiences and thus data for black hispanics will likely be a lot different than that of white hispanics due to racism within the hispanic community [@Haywood2017].
+This problem may also arise because much of society sees all hispanics as a mix of Indigenous and European, thus erasing Black and Indigenous Hispanic experiences.
 
@@
 This is because undocumented mothers are much more likely to give birth at home or less formal places, due to fear of eviction from the country.
-Thus they often do not have birth certificates, or have inaccurate birth certificates.
+Thus, they often do not have birth certificates, or have inaccurate birth certificates.
 As these records are used for planning infrastructure and amenities to benefit the populace, not including them perpetuates their exclusion (many of whom pay taxes [@citetax]) from society and societal improvements.
 This perpetuates discrimination against them as they are also misused for labor, and targeted by ICE.
-ALthough census information does include everybody irregardless of status, many undocumented immigrants also fear coming out due to fear of being extrajudicially removed.
+Although census information does include everybody regardless of status, many undocumented immigrants also fear coming out due to fear of being extrajudicially removed.
 This census data is also used for funding in particular areas and for infrastructure improvement, so these inaccurate results mean that not only are their experiences being left out, but areas where there are many undocumented are not getting the funding required.
@@
 While the datasets included information on race, there are possible problems in data collection due to the subjectivity of race.
-There was no option for mixed race, so it is assumed that every person in the data were classified according to three arbitrary groups.
-For example, the sources did not provide at what point someone was no longer considered to be black, or if white hispanics were classified under white or under hispanic.
-What may be considered white or considered black may also change depending on the state or the community; i.e a mixed race person in a black community would be considered biracial, but in a white community may be considered black.
-If during data collection, the sources decided on self identification of survey takers, this could lead to inaccurate results.
+There was no option for mixed race, so it is assumed that every person in the data was classified according to three arbitrary groups.
+For example, the sources did not provide at what point someone was no longer considered to be black, or if white Hispanics were classified under white or under hispanic.
+What may be considered white or black may also change depending on the state or the community; i.e.
+a mixed-race person in a Black community may be considered to be biracial, but in a white community may be considered to be Black.
+If during data collection, the sources decided on self-identification of survey takers, this could lead to inaccurate results.
 However if the data collectors decided to choose the race for themselves, this could strip people of their ethnic identity.
@@
 The article uses data from the 2002 National Survey of Family Growth (NSFG) to show the results.
-Factors such as unwanted fertility, age at childbearing, or degree of fertility postponement seem not to contribute to religiosity differentials in fertility.The article states how studies consistently find higher fertility among more religiously active women across denominations, although the magnitude of the effect varies with the measures used for fertility and for religiosity.
+Factors such as unwanted fertility, age at childbearing, or degree of fertility postponement seem not to contribute to religiosity differentials in fertility.
+The article states how studies consistently find higher fertility among more religiously active women across denominations, although the magnitude of the effect varies with the measures used for fertility and for religiosity.
 The results from the survey show how women who value religion higher in their daily life have more children compared to those who value their religion somewhat to none.
-This shows how being religious can have a direct correlation to a higher fertility which causes more birth.
-Society in general has become less religious which is shown in the declining birth rates.
+This shows how being religious can have a direct correlation to higher fertility, which leads to more births.
+Society in general has become less religious, which is shown in the declining birth rates.
 [@citeusatrends] shows how marriage for women in the 20 to 24 age group has decreased, showing that generation Z are choosing the marriage and children route at later stages in their lives.
@@
 Since women are marrying late, they are giving birth at a later stage in life compared to before.
-However it should be noted that the amount of women who have children without being married has increased in the past 20 years.
+However it should be noted that the number of women who have children without being married has increased in the past 20 years.
 
@@
 The choice to not have children is not as shameful in society compared to 20 years ago.
-Having children results in more responsibility being taken up which can restrict career advancements [@citeexecutive], as western culture still has expectations that women do  childcare when men go to work, so women often feel they can't do both [@Maher2007].
+Having children results in more responsibility being taken up which can restrict career advancements [@citeexecutive], as western culture still has expectations that women take care of children and men go to work, so women often feel they can't do both [@Maher2007].
 The cost of delaying their careers to have children also increases as they get deeper into their profession.
-Women who do 'have the best of both worlds' need to juggle both work and home, creating difficulties for themselves which can have major effect on their health.
-Maternity leave in the US on an average result in 10 weeks while the paid leave is 8 weeks.
-This can cause an issue in terms of finance as living expenses has increased from the 2000s, especially since the recession in 2008.
+Women who do 'have the best of both worlds' need to juggle both work and home, creating difficulties for themselves which can have a major effect on their health.
+Maternity leave in the US on average results in 10 weeks while the paid leave is 8 weeks.
+This can cause an issue in terms of finance as living expenses have increased from the 2000s, especially since the recession in 2008 [@citeusatrends].
 Two-parent households are on the decline in the United States as divorce, remarriage and cohabitation are on the rise.
@@
 The rise of LGBT families for generation Z has also led to an increase in adoptions and decrease in birth rates.
-Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed.The increase in women achieving higher education [@citeusatrends] has also led to a pause to the child rearing stage for many, which has also caused a decrease in the birth rates for young women as this causes women to have children later in life after they are done with education.The causes of these decrease in births are tied to increases in educational attainment, growing labor force participation.
+Not only are Americans having fewer children, but the circumstances surrounding parenthood have changed.
+The increase in women achieving higher education [@citeusatrends] has also led to a pause to the child rearing stage for many, which has also caused a decrease in the birth rates for young women as this causes women to have children later in life after they are done with education.
+The causes of these decrease in births are tied to increases in educational attainment, growing labor force participation.
 
@@
 
-White birth rates have decreased significantly, and the 15-24 groups contribute at more than 18% of declining birth rates.
+White birth rates have decreased significantly, and the 15-24 age group contributes more than 18% of declining birth rates.
 This decrease in white birth rates is a well known talking point of the far right, which is used to harass people of color, especially mothers.
@@
 These feelings of animosity are also fueled by immigrants tendency to have more children than the citizen populations.
-However these arguments often do not take into account the fact that second and third generation immigrants tend to reduce their birth rates to match the rest of the population [@immigbirth].
+However these arguments often do not consider the fact that second and third generation immigrants tend to reduce their birth rates to match the rest of the population [@immigbirth].
 This decrease in birth rates is thus used both as a justification for immigration and job outsourcing, as well as an 'invasion crisis'.
@@
 Hispanic women also have less than stellar access to fertility health care, as well as have lower IVF success rates [@citeracefertility].
-Black people are concentrated in the south, which has very high levels of teenage pregnancy [@citeteen] and low level of resources on sexual education, so they are disproportionately affected.
+Black people are concentrated in the south, which has very high levels of teenage pregnancy [@citeteen] and low levels of resources on sexual education, so they are disproportionately affected compared to white people.
 This may be why they did not contribute much to declining birth rates in the report.
-Black teenagers are also less likely to have access to quality education,careers and outside opportunities[@citeusatrends], which is a major factor to the decline in birth rates.
-Hispanics are also affected by many of the same problems; lack of funding and proper education in their communities.
-As birth rates for people of color has not decreased as much as those for white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.
+Black teenagers are also less likely to have access to quality education, careers and outside opportunities[@citeusatrends], which is a major factor towards the comparatively small decline in birth rates.
+Hispanics are also affected by many of the same problems: lack of funding and proper education in their communities.
+As birth rates for people of color have not decreased as much compared to white people, it could also perpetuate harmful stereotypes of minority women having too many children, too young, depending on welfare.
 Black women are also at a higher risk of unwanted pregnancies compared to other populations, as shown by [@Bryant2010], which surveyed women who had recently gotten pregnant and asked questions such as if they intended on getting pregnant.
 Among the proportion of women who answered no to this question, black women represented 33% of the women not intending for pregnancy.
-Teenage birth rates have decreased for generation z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted even with the cultural change.
+Teenage birth rates have decreased for generation Z, however black and hispanic women still have higher teen birth rates compared to white women, and this has persisted even with the cultural change.
 This is due to lack of opportunities, as well as higher rates of poverty and low education [@Gold2001].
-While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back.This shows yet another way in which women of color are negatively treated in society.
+While these problems are due to external problems from lack of funding to these communities, this data does not provide the vital background to dispel systemic biases that black and hispanic women are just lazy and have children on the taxpayer's back.
+This shows yet another way in which women of color are negatively treated in society.
 
@@
 
-Table 1 examined economic factors that may play a role in the living standards and motherhood decision of young women.
+(Table \@ref(tab:tabone)) examined economic factors that may play a role in the living standards and motherhood decision of young women.
 Student debt, rent and childcare costs increased from the early 2000s to late 2010s.
@@
 Increasing rent and lack of social safety nets due to the U.S's individualistic tendencies contributes to why generation Z couples have second thoughts about having a child[@citerentbabies].
-In some cities in 2022, rent is increasing by up to 40% [@cite40] and increases like that  push women against having children in less favorable positions [@livingston2020].
+In some cities in 2022, rent is increasing by up to 40% [@cite40], and increases like that push women against having children in less favorable positions [@livingston2020].
 In a similar way, the rise of student debt in the United States is also leaving young adults, more so young women, with less money to care for their families.
 According to [@citewomenloans], women in the United States hold approximately 67% of student loans and many of them state that their debt influences their decision to have a child.
-When considering these two costs, and considering the fact that there are child care expenses to tend to as well, it becomes apparent how expensive it is becoming to have a child.
+When considering these two costs, and the fact that there are child care expenses to tend to as well, it becomes apparent how expensive it is becoming to have a child.
 According to [@citechildcarecost], a dollar increase in weekly childcare costs implies a 2% decrease in birth rate.
-Along with this, a dollar increase in childcare cost also means that the rate of entering employment decreases by 3%.
-From this, one can see how there is not only a correlation between increasing childcare costs and unemployment but also a correlation with increasing childcare costs and a decreasing birth rate.In addition to these, the rise of urbanization has caused a decrease in birth rates [@citenargund], and urbanization is directly correlated to increase in rent.
+Along with this, a dollar increase in childcare costs also means that the rate of entering employment decreases by 3%.
+From this, one can see how there is not only a correlation between increasing childcare costs and unemployment but also a correlation with increasing childcare costs and a decreasing birth rate.
+In addition to these, the rise of urbanization has caused a decrease in birth rates [@citenargund], and urbanization is directly correlated to increases in rent.
 
@@
 Studies show that younger adults struggle to be financially stable, so the opportunity that an expansion brings for them would positively affect their financial status putting them in a better position to have a child [@citeyoungadult].
-This growth is not evident when looking at birthrates for the 15-19 age group as the declining birth rate continues until 2005.
-In contrast, the results of this expansion period are more visible in the 20-24 age range as it seems that the birth rate from 2001 to 2005 decreased at a lower rate each consecutive year until 2005, where it began to increase.
+This growth is not evident when looking at birth rates for the 15-19 age group as the declining birth rate continues until 2005.
+In contrast, the results of this expansion period are more visible in the 20-24 age range as it seems that the birth rate from 2001 to 2005 decreased at a lower rate each consecutive year until 2005, when it began to increase.
 This is likely because of the fact that the 15-19 age group is not affected by an economic expansion as much as the 20-24 age group.
 On the contrary, an economic expansion provides a lot more jobs and income for young adults who are done or almost done with school.
-In addition, in times of expansion families are more likely to be in a position where they can afford to have a child because interest rates and the real value of debts are lower than usual.
+In addition, in times of expansion, families are more likely to be in a position where they can afford to have a child because interest rates and the real value of debts are lower than usual.
 With more people employed, there is generally more income for families to support the needs of children.
-However, that growth began to decline in 2007 where consumers began to pay more for debt that they had to repay.
+However, that growth began to decline in 2007 when consumers began to pay more for their debt.
 It is stated that economic growth had decreased to 2% in the third quarter compared to the 5.6% of first quarter growth the country had.
 This time also marked the end of the housing boom which is generally caused by economic prosperity and lower credits.
-Due to this, the decline of the economy also meant the end of the housing boom and also, the beginning of the declining birth rates [@citeusaecon2006].
-In (Figure \@ref(fig:figsix)), we observe a large decrease in births for both age groups as seen by the steep negative slope,  likely caused by the great recession which began in late 2006 and early 2007.
+Due to this, the decline of the economy also meant the beginning of the declining birth rates [@citeusaecon2006].
+In (Figure \@ref(fig:figsix)), we observe a large decrease in births for both age groups, likely caused by the great recession which began in late 2006 and early 2007.
 The points on the graph also show higher unemployment rates due to the recession.
@@
 
-# State by state changes
+# State Difference
 
-The change in birthrates for some states were very different than others.
-Some states were able to bounce back from the economic depression, while some states have always had cheaper costs of living.
-From *FIGURE MAP* These states, primarily the midwest, are still heavily affected by the economic recession that started due to the fall of the steel industry [@Meyer2003] and outside sourcing of labor.
-However cost of living and childcare are also relatively cheap in these states [@citerent], so it birth rates were not heavily affected.
+The change in birth rates for some states were very different from others.
+Some states were able to recover from the economic depression, while some states did not.
+From (Figure \@ref(fig:figseven)), these states, primarily the midwest, are still heavily affected by the economic recession that started due to the fall of the steel industry [@Meyer2003] and outside sourcing of labor.
+However, the cost of living and child care are also relatively cheap in these states [@citerent], so birth rates were not heavily affected.
 The west and east coast was able to bounce back better than the midwest, but as the cost of living is still expensive [@McMahon1991], this could deter women from having children quite early until they are financially ready.
 In the south, the cost of living is also quite high but as these states tend to be more religious and conservative than the liberal west coast, birth rates did not decline as far as the west coast.
-The religious level of the south also corresponds to higher teen birth rates [@Strayhorn2009].
-The east coast is still socially liberal and wealthier so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are a bit higher despite the expensive cost of living.
+The higher religious levels of the south also corresponds to higher teen birth rates [@Strayhorn2009].
+The east coast is still socially liberal and relatively wealthier, so they provide more safety nets in terms of resources and money for mothers than the south, which is why birth rates are higher despite the expensive cost of living.
 Southern states also tend to have more problems with abortion and aim to restrict them, which could also be a reason behind the comparatively higher birth rates.
-However as generation Z tends to be more liberal than their millennial counterparts they are less likely to have more children nationwide.
+However as generation Z tends to be more liberal than their millennial counterparts, they are less likely to have more children nationwide.
 
@@
 
-There were datasets in which we had to make generalizations about a specific population based on data of a general population.
-For example, in (Figure \@ref(fig:figseven)), the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group were not available.
+There were datasets in which we had to generalize about a specific population based on data of a general population.
+For example, in (Figure \@ref(fig:figseven)), the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group was not available.
 Due to this, conclusions that are made from this map are likely not accurate as they are not the true statistics of the 15 to 24 age group.
 Our research question wanted to focus on the trend of birth rates over a 20 year period, specifically from 2000 to 2020.
-However, for several datasets only data up to 2019 was available.
-Due to this, instead of being able to compare 2000 and 2020 explicitly, we had to generalize trends that we see in 2019 to be at least similar to trends that we’d see in 2020.
-We had the same problem with (Table \@ref(tab:tabtwo)), in which we had to approximate 2004 to 2000 even though situations would be different.
-In addition to this, the years for millennial and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
+However, for several datasets, only data up to 2019 was available.
+Due to this, instead of being able to compare 2000 and 2020 explicitly, we had to generalize trends that we see in 2019 to be at least similar to trends that we would see in 2020.
+We had the same problem with (Table \@ref(tab:tabtwo)), in which we had to approximate 2004 data for 2000 trends, even though situations would be different.
+In addition to this, the year ranges for millennials and generation Z are subjective, as many different years have been suggested as the start and end for these two generations.
 Thus the years chosen for this report are completely arbitrary.
@@
 There has also been much controversy in the past two decades over the falling sperm counts in men [@JohnAitken2013].
-We are interested in if this has had any effect on birth rates, and in future research we would like to explore the relationship between them.
+We are interested in whether this has had any effect on birth rates, and in future research we would like to explore the relationship between them.
 We would also like to download the data directly from the sources to clean and form our own initial datasets.

### 2022-02-28 Main Code repository with data and paper/3 Graph Replication.R

@@ -1,9 +1,13 @@
 # Loading in datasets
+library("tidyverse")
+library("haven")
 
 birth = read_csv("Inputs/Data/fig_1.csv")
 race = read_csv("Inputs/Data/figs_2a_2b.csv")
-write_csv(birth, "Replication-Birth Rates/birth.csv")
-write_csv(race, "Replication-Birth Rates/race.csv")
+write_csv(birth, "Main Code repository with data and paper/birth.csv")
+write_csv(race, "Main Code repository with data and paper/race.csv")
+birth = read_csv(here::here("Main Code repository with data and paper/birth.csv"))
+race = read_csv(here::here("Main Code repository with data and paper/race.csv"))
 
 # Replication of Figure 1
 birth %>%

### 2022-02-28 Output/paper 2.Rmd
Final changes

@@
 ---
-title: "The Decline in U.S Birthrates in Recent Years is Indicative of Socio-economic Changes"
+title: "The Decline in U.S Birthrates in Recent Years is Indicative of Cultural and Economic Changes"
 author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
-thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
+thanks: 'Code, data and reproduction package are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
 
-## Results
+# Results
 
@@
 This concern over the white birth rate decline (often coded as 'True Americans') and the notion of 'being replaced' has led to harmful anti-immigration policies, as well as the rise of far right hate groups throughout the country [@Chermak2013].The increase and proliferation of far right ideologies that white birth rates are decreasing leading to a less white country are fueled in part by the way that these statistics can be shown; without the proper context to show why they are decreasing.
-Many pro-life advocates have also noted declining birth rates as a major reason for their antipathy towards abortion, which has caused an increase in attacks on planned parenthood clinics as they are believed to be a symptom of societal decay.
-[@Primrose].
+Many pro-life advocates have also noted declining birth rates as a major reason for their antipathy towards abortion, which has caused an increase in attacks on planned parenthood clinics as they are believed to be a symptom of societal decay [@Primrose].
 
@@
 
-# State Difference
+## State Differences
 
@@
 There were datasets in which we had to generalize about a specific population based on data of a general population.
-For example, in (Figure \@ref(fig:figseven)), the map data was composed of responses taken from women aged from 20-34, as information for the 15 to 24 age group was not available.
+For example, in (Figure \@ref(fig:figseven)), the map data was composed of responses taken from women aged from 20 to 34, as information for the 15 to 24 age group was not available.
 Due to this, conclusions that are made from this map are likely not accurate as they are not the true statistics of the 15 to 24 age group.

### 2022-02-28 Output/paper 2.Rmd
Final changes.

@@
 For the average debt, we multiplied by 10,000 as that is how it was shown in the original source.
-We got the variable on the importance of religion by dividing the number of people who felt religion was important over the total number surveys.
+We got the variable on the importance of religion by dividing the percentage of people who felt religion was important over the total percentage of people surveyed.
 Note that for these variables we needed to use 2004 data as 2000 data was not available.
